---
title: "Parameter Recovery and Identification"
subtitle: "The aggregate model: simulation, estimation, and what identification buys"
author:
  - name: Dan Yavorsky
  - name: Geoffery Zheng
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

## What this notebook does

This notebook is the evidence behind **Parameter Recovery and Identification**
in the paper's simulation study, aggregate half. It runs the whole program:

1. a detailed walkthrough of one cell, from behavioral draws to recovered
   probability-scale cut points;
2. the full recovery grid, three parameter configurations for each of the two
   links;
3. the negative control, where an intercept is appended and identification
   fails in exactly the predicted way;
4. the factorization check, confirming that the joint distribution of the two
   responses is the product the derivation claims;
5. repeated sampling across independent datasets, to check bias and
   standard-error calibration;
6. a scaling check confirming the residual bias is the ordinary $O(1/n)$
   finite-sample effect rather than something structural.

Parts 3 and 4 are the ones the article does not spend space on, because
identification is argued analytically in the model section. They belong here.

This notebook is the **source** of three results files the article reads:
`identification_test.rds`, `replication_study.rds`, and
`bias_scaling_check.rds`. Its seeds are the article's seeds, so the numbers
below are the numbers in the paper, not a fresh draw.

The functions are deliberate copies of `R/lib/dgp.R` and `R/lib/likelihood.R`,
unrolled and commented so the mechanics are visible without opening the
archive.

A word on why the data are simulated behaviorally rather than from the closed
form. If we drew responses from `alpha_w^S - alpha_{w-1}^S` and then fit
`alpha_w^S - alpha_{w-1}^S`, recovery would confirm only that `optim` works. By
drawing utilities, taking an argmax, and applying the reporting rule, recovery
tests the derivation itself: the claim that the maximum of Gumbel utilities is
Gumbel with location equal to the log-sum inclusive value, and that it is
independent of which alternative attained it.

```{r}
#| label: setup
set.seed(1)
options(digits = 5)
```

## 1. The model in one screen

A consumer faces $J$ profiles. Profile $j$ carries attributes $x_j$ and utility

$$
u_j = x_j'\beta + \varepsilon_j, \qquad \varepsilon_j \sim \text{Gumbel}(0, 1) \text{ iid}.
$$

**First response.** She is asked which she prefers and picks the best one,
$j^* = \arg\max_j u_j$. This is the usual multinomial logit.

**Second response.** She is asked how likely she is to buy it. She knows
everything about the profiles on screen, but not her valuation of the outside
good $\eta_0 \sim \text{Gumbel}(0,1)$. So what she holds is not a decision but a
probability,

$$
p = \Pr(\eta_0 < u^* \mid u^*) = F(u^*), \qquad u^* = \max_j u_j,
$$

where $F(z) = \exp(-e^{-z})$ is the standard Gumbel CDF.

**Model A** says she reports the interval containing that probability. Given cut
points $0 = \alpha_0 < \alpha_1 < \cdots < \alpha_{W-1} < \alpha_W = 1$ on the
probability scale,

$$
y = w \iff p \in [\alpha_{w-1}, \alpha_w).
$$

From the researcher's seat $u^*$ is random. The key lemma is that
$u^* \sim \text{Gumbel}(\overline{\mu}, 1)$ with $\overline{\mu} = \ln S$ and
$S = \sum_j e^{V_j}$, and that $u^*$ is independent of $j^*$. Evaluating that
CDF at $F^{-1}(\alpha)$ collapses to a power, which gives the whole second-stage
likelihood:

$$
\Pr(y = w) = \alpha_w^{\,S} - \alpha_{w-1}^{\,S}.
$$

Equivalently, from the researcher's perspective the consumer's purchase
probability is $\text{Beta}(S, 1)$ distributed, with the inclusive value as its
only shape parameter.

## 2. A design

Two three-level attributes, dummy coded against a reference level, plus a
continuous price-like attribute. Five columns, no intercept. The absence of an
intercept is not cosmetic, as Section 9 below shows.

```{r}
#| label: design
make_design <- function(n_tasks, J, seed, intercept = FALSE) {
  set.seed(seed)
  n_rows <- n_tasks * J
  a     <- sample(1:3, n_rows, replace = TRUE)
  b     <- sample(1:3, n_rows, replace = TRUE)
  price <- runif(n_rows, 0.5, 2.5)
  X <- cbind(
    a2    = as.numeric(a == 2),
    a3    = as.numeric(a == 3),
    b2    = as.numeric(b == 2),
    b3    = as.numeric(b == 3),
    price = price
  )
  if (intercept) X <- cbind(X, const = 1)
  list(X = X, n_tasks = n_tasks, J = J, P = ncol(X))
}

N_TASKS <- 20000   # matches the paper's aggregate experiments
J       <- 4
W       <- 5       # a five-point purchase-likelihood scale

design <- make_design(N_TASKS, J, seed = 101)
head(design$X, 8)
```

Rows are stacked task-major: rows 1 to 4 are the four profiles of task 1, rows
5 to 8 are task 2, and so on.

## 3. True parameters

The scale labels are set on the probability scale, which is what makes Model A
interpretable: $\alpha_1 = 0.10$ says that answering "1" means a purchase
probability below ten percent.

```{r}
#| label: truth
beta_true  <- c(a2 = 0.8, a3 = -0.5, b2 = 0.4, b3 = 1.0, price = -0.9)
alpha_true <- c(0.10, 0.30, 0.60, 0.85)

# Cut points live on the utility scale internally. F^{-1}(a) = -log(-log a).
alpha_to_cut <- function(a) -log(-log(a))
cut_true <- alpha_to_cut(alpha_true)

rbind(alpha = alpha_true, cut = cut_true)
```

## 4. Simulate the behavioral process

Nothing here uses the likelihood. We draw utilities, take the argmax, convert the
winning utility into the consumer's purchase probability, and bin it.

```{r}
#| label: simulate
rgumbel <- function(n) -log(-log(runif(n)))

simulate_modelA <- function(design, beta, cut, seed) {
  set.seed(seed)
  n <- design$n_tasks
  J <- design$J

  # Deterministic utilities, reshaped to one row per task.
  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)

  # Add iid Gumbel noise: this is the consumer's single, internally
  # consistent utility draw for the task.
  u <- V + matrix(rgumbel(n * J), n, J)

  # First response: which profile is best.
  jstar <- max.col(u, ties.method = "first")

  # The utility she actually attains.
  ustar <- u[cbind(seq_len(n), jstar)]

  # Model A: she reports the interval holding p = F(ustar). Binning p against
  # alpha is the same as binning ustar against cut, since F is increasing.
  y <- findInterval(ustar, cut) + 1L

  list(design = design, jstar = jstar, y = y, W = length(cut) + 1L,
       ustar = ustar, V = V)
}

dat <- simulate_modelA(design, beta_true, cut_true, seed = 501)
```

### What the two responses look like

```{r}
#| label: responses
rbind(
  `first response (chosen profile)` = table(dat$jstar) / N_TASKS,
  `second response (scale point)`   = table(dat$y)     / N_TASKS
)
```

The first response is roughly uniform because the design randomizes attributes
across positions. The second response is not, and its shape is what the cut
points control.

### The lemma is visible in the simulated data

Two claims underlie the whole factorization. Both can be read straight off the
draws, before any estimation.

```{r}
#| label: lemma-check
S     <- rowSums(exp(dat$V))
mubar <- log(S)

# (i) ustar is Gumbel(mubar, 1). Standardize and compare to a standard Gumbel.
z <- dat$ustar - mubar
cat("mean of standardized ustar:", round(mean(z), 4),
    " (Euler-Mascheroni = 0.5772)\n")
cat("sd   of standardized ustar:", round(sd(z), 4),
    " (pi/sqrt(6) = 1.2825)\n")
cat("KS test vs standard Gumbel, p =",
    round(ks.test(sample(z, 5000), function(q) exp(-exp(-q)))$p.value, 3), "\n")

# (ii) ustar is independent of jstar: the distribution of the standardized
# maximum should not depend on which profile won.
round(tapply(z, dat$jstar, mean), 4)

# Test it rather than eyeball it. Under independence the four groups share a
# distribution, so a Kruskal-Wallis test should not reject.
kruskal.test(z, factor(dat$jstar))
```

The group means sit within sampling error of one another and the test does not
reject. That independence is what lets the joint likelihood factor into a choice
term and an ordinal term, and it is the part of the lemma that is easiest to
doubt: it says a consumer who picked profile 3 is no more or less enthusiastic,
on average, than one who picked profile 1.

## 5. The likelihood

The joint probability of one task is the product of the two pieces:

$$
\underbrace{\frac{e^{V_{j^*}}}{S}}_{\text{first response}}
\times
\underbrace{\left( \alpha_{y}^{\,S} - \alpha_{y-1}^{\,S} \right)}_{\text{second response}}.
$$

We optimize over an unconstrained parameterization that keeps the cut points
ordered by construction: the first cut point free, then the logs of successive
gaps.

```{r}
#| label: likelihood
par_to_cut <- function(par, P, W) {
  if (W == 2) return(par[P + 1])
  par[P + 1] + c(0, cumsum(exp(par[(P + 2):(P + W - 1)])))
}
cut_to_par <- function(cut) c(cut[1], log(diff(cut)))

negloglik <- function(par, dat) {
  design <- dat$design
  P <- design$P; W <- dat$W
  n <- design$n_tasks; J <- design$J

  beta <- par[1:P]
  cut  <- par_to_cut(par, P, W)

  V     <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  m     <- do.call(pmax, as.data.frame(V))     # row maxima, for stability
  logS  <- m + log(rowSums(exp(V - m)))        # log inclusive value

  # First response: multinomial logit log-probability of the chosen profile.
  ll_choice <- V[cbind(seq_len(n), dat$jstar)] - logS

  # Second response: difference of Gumbel CDFs at the standardized cut points.
  caug <- c(-Inf, cut, Inf)
  lo <- caug[dat$y]      - logS
  hi <- caug[dat$y + 1L] - logS
  p_ord <- exp(-exp(-hi)) - exp(-exp(-lo))

  -(sum(ll_choice) + sum(log(pmax(p_ord, 1e-312))))
}
```

A note on that second-response line. Writing it as a plain difference of two
`exp(-exp(-z))` terms is clear but loses precision when both cut points sit far
into a tail and the two CDFs nearly cancel. The production code uses the
algebraically equivalent but numerically stable rearrangement
$e^{-e^{-h}}\left(1 - e^{-(e^{-l} - e^{-h})}\right)$. At the parameter values
here the difference is invisible; in the stress configurations of the paper it
is not.

## 6. Estimate

Starting values: $\beta = 0$, and cut points backed out of the observed category
frequencies through the link. No knowledge of the truth goes in.

```{r}
#| label: fit
fit_modelA <- function(dat) {
  P <- dat$design$P; W <- dat$W

  freq <- tabulate(dat$y, nbins = W)
  cumq <- cumsum(freq)[1:(W - 1)] / sum(freq)
  cut0 <- log(dat$design$J) + alpha_to_cut(cumq)
  start <- c(rep(0, P), cut_to_par(cut0))

  fn  <- function(p) negloglik(p, dat)
  opt <- optim(start, fn, method = "BFGS",
               control = list(maxit = 1000, reltol = 1e-12))
  H   <- optimHess(opt$par, fn)      # observed information
  Vc  <- solve(H)

  cut_hat <- par_to_cut(opt$par, P, W)

  # Delta method for the cut points, which are a nonlinear function of the
  # working parameters.
  Jc <- matrix(0, W - 1, length(opt$par))
  Jc[, P + 1] <- 1
  if (W > 2) {
    d <- exp(opt$par[(P + 2):(P + W - 1)])
    for (w in 2:(W - 1)) Jc[w, (P + 2):(P + w)] <- d[1:(w - 1)]
  }

  list(beta = opt$par[1:P], se_beta = sqrt(diag(Vc)[1:P]),
       cut = cut_hat, se_cut = sqrt(diag(Jc %*% Vc %*% t(Jc))),
       nll = opt$value, convergence = opt$convergence,
       eigen = eigen(H, symmetric = TRUE, only.values = TRUE)$values,
       par = opt$par)
}

fit <- fit_modelA(dat)
cat("converged:", fit$convergence == 0, "\n")
```

## 7. Did the parameters come back?

```{r}
#| label: recovery
recovery <- data.frame(
  parameter = c(names(beta_true), paste0("c", 1:(W - 1))),
  truth     = c(beta_true, cut_true),
  estimate  = c(fit$beta, fit$cut),
  std_error = c(fit$se_beta, fit$se_cut)
)
recovery$z <- (recovery$estimate - recovery$truth) / recovery$std_error
knitr::kable(recovery, digits = 4)
```

Every $z$ statistic is small. With nine parameters we would expect roughly one
value above two in absolute terms about forty percent of the time, so a table
with none is unremarkable rather than suspicious.

### The cut points on the probability scale

This is where Model A pays off. Pushing the estimated cut points back through
the link turns scale labels into purchase probabilities, with standard errors.

```{r}
#| label: alpha
alpha_hat <- exp(-exp(-fit$cut))
se_alpha  <- fit$se_cut * alpha_hat * (-log(alpha_hat))   # delta method

knitr::kable(
  data.frame(
    label      = paste("answer", 1:(W - 1), "or below"),
    alpha_true = alpha_true,
    alpha_hat  = alpha_hat,
    std_error  = se_alpha
  ), digits = 4)
```

A respondent choosing the top box is saying her purchase probability exceeds
`r sprintf("%.3f", alpha_hat[4])`, and we can attach a standard error to that
statement. This is the quantity that dichotomizing the scale throws away and that
assigning fixed probabilities to scale points gets wrong, because the meaning of
a label depends on what was on the screen.

## 8. Diagnostics

**Is the truth inside the confidence region?** Compare twice the
log-likelihood gap between the truth and the MLE against a $\chi^2_9$ critical
value.

```{r}
#| label: lrtest
nll_truth <- negloglik(c(beta_true, cut_to_par(cut_true)), dat)
lr   <- 2 * (nll_truth - fit$nll)
npar <- length(fit$par)
cat(sprintf("LR statistic %.2f vs chi-square(%d) 95%% critical value %.2f -> %s\n",
            lr, npar, qchisq(0.95, npar),
            ifelse(lr < qchisq(0.95, npar), "inside the region", "OUTSIDE")))
```

**Is the information matrix well conditioned?** A strongly positive definite
Hessian is the numerical face of identification.

```{r}
#| label: eigen
cat("smallest eigenvalue:", format(min(fit$eigen), digits = 4), "\n")
cat("condition number:   ", format(max(fit$eigen) / min(fit$eigen), digits = 4), "\n")
```


## 9. The full recovery grid

The walkthrough above is one cell. The article's claim is broader: recovery
holds across parameter configurations that put the response distribution in
very different places, and for both links. Three configurations, two links, six
fits.

```{r}
#| label: grid-setup
W <- 5

param_sets <- list(
  set1_moderate = list(
    beta   = c(a2 = 0.8, a3 = -0.5, b2 = 0.4, b3 = 1.0, price = -0.9),
    cutB   = c(-1.0, 0.2, 1.2, 2.2),
    alphaA = c(0.10, 0.30, 0.60, 0.85)
  ),
  set2_negative = list(
    beta   = c(a2 = -1.2, a3 = 0.3, b2 = -0.6, b3 = 0.7, price = -1.5),
    cutB   = c(-2.5, -1.2, 0.5, 1.0),
    alphaA = c(0.05, 0.25, 0.55, 0.90)
  ),
  set3_stress = list(
    beta   = c(a2 = 2.0, a3 = -1.8, b2 = 1.2, b3 = -0.7, price = -2.2),
    cutB   = c(-3.5, -2.2, 0.8, 3.0),
    alphaA = c(0.02, 0.10, 0.70, 0.97)
  )
)
```

The `set3_stress` configuration is the demanding one. Its $\alpha_1 = 0.02$ puts
the bottom category at two percent of the probability scale, so only about one
task in fifty lands there and the corresponding cut point has to be pinned down
from little data.

Model B needs the general machinery: a logistic link rather than a Gumbel one.
The only thing that changes is the interval probability.

```{r}
#| label: general-likelihood
# Interval probability G(hi) - G(lo), tail-stable, for either link.
# Model A: G = standard Gumbel CDF   (she reports the probability she holds)
# Model B: G = standard logistic CDF (she resolves, then grades)
ord_prob_z <- function(lo, hi, model) {
  if (model == "B") {
    plogis(hi) * plogis(-lo) * (-expm1(lo - hi))
  } else {
    ea <- exp(-lo); eb <- exp(-hi)
    p <- exp(-eb) * (-expm1(-(ea - eb)))
    p[!is.finite(eb)] <- 0
    p
  }
}

ord_prob <- function(mubar, cut, y, model) {
  caug <- c(-Inf, cut, Inf)
  ord_prob_z(caug[y] - mubar, caug[y + 1L] - mubar, model)
}

row_max <- function(M) do.call(pmax, as.data.frame(M))

negloglik <- function(par, dat) {
  design <- dat$design
  P <- design$P; W <- dat$W
  n <- design$n_tasks; J <- design$J
  beta <- par[1:P]
  cut  <- par_to_cut(par, P, W)

  V    <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  m    <- row_max(V)
  logS <- m + log(rowSums(exp(V - m)))

  ll_choice <- V[cbind(seq_len(n), dat$jstar)] - logS
  ll_ord    <- log(pmax(ord_prob(logS, cut, dat$y, dat$model), 1e-312))
  -(sum(ll_choice) + sum(ll_ord))
}

num_grad <- function(f, x, eps = 1e-6) {
  vapply(seq_along(x), function(k) {
    h <- eps * max(1, abs(x[k]))
    xp <- x; xp[k] <- xp[k] + h
    xm <- x; xm[k] <- xm[k] - h
    (f(xp) - f(xm)) / (2 * h)
  }, numeric(1))
}
```

The simulator likewise generalizes. Model B differs in one line: she resolves
her uncertainty by drawing the outside good, and grades the comparison
$u^* - \eta_0$ rather than reporting $F(u^*)$.

```{r}
#| label: general-sim
simulate_dual <- function(design, beta, cut, model = c("A", "B"), seed) {
  model <- match.arg(model)
  set.seed(seed)
  n <- design$n_tasks; J <- design$J
  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  u <- V + matrix(rgumbel(n * J), n, J)
  jstar <- max.col(u, ties.method = "first")
  ustar <- u[cbind(seq_len(n), jstar)]
  latent <- if (model == "A") ustar else ustar - rgumbel(n)
  y <- findInterval(latent, cut) + 1L
  list(design = design, jstar = jstar, y = y, W = length(cut) + 1L,
       model = model, beta_true = beta, cut_true = cut)
}

fit_dual_mle <- function(dat, start = NULL) {
  design <- dat$design; P <- design$P; W <- dat$W
  if (is.null(start)) {
    freq <- tabulate(dat$y, nbins = W)
    cumq <- pmin(pmax(cumsum(freq)[1:(W - 1)] / sum(freq), 1e-4), 1 - 1e-4)
    ginv <- if (dat$model == "B") qlogis else function(q) -log(-log(q))
    cut0 <- log(design$J) + ginv(cumq)
    if (W > 2) for (k in 2:(W - 1)) cut0[k] <- max(cut0[k], cut0[k - 1] + 1e-3)
    start <- c(rep(0, P), cut_to_par(cut0))
  }
  fn  <- function(p) negloglik(p, dat)
  opt <- optim(start, fn, method = "BFGS",
               control = list(maxit = 1000, reltol = 1e-12))
  H  <- optimHess(opt$par, fn)
  g  <- num_grad(fn, opt$par)
  ev <- eigen(H, symmetric = TRUE, only.values = TRUE)$values

  cut_hat <- par_to_cut(opt$par, P, W)
  Jc <- matrix(0, W - 1, length(opt$par)); Jc[, P + 1] <- 1
  if (W > 2) {
    d <- exp(opt$par[(P + 2):(P + W - 1)])
    for (w in 2:(W - 1)) Jc[w, (P + 2):(P + w)] <- d[1:(w - 1)]
  }
  Vpar <- tryCatch(solve(H), error = function(e)
    matrix(NA, length(opt$par), length(opt$par)))

  out <- list(beta = opt$par[1:P], se_beta = sqrt(diag(Vpar)[1:P]),
              cut = cut_hat, se_cut = sqrt(diag(Jc %*% Vpar %*% t(Jc))),
              nll = opt$value, convergence = opt$convergence,
              grad_max = max(abs(g)), eigen = ev, min_eig = min(ev),
              cond = max(ev) / max(min(ev), .Machine$double.eps),
              par = opt$par, counts = opt$counts)
  if (dat$model == "A") {
    a <- exp(-exp(-cut_hat))
    out$alpha <- a
    out$se_alpha <- out$se_cut * a * (-log(a))
  }
  out
}

id_rank_check <- function(X) {
  aug <- cbind(X, 1)
  list(rank = qr(aug)$rank, required = ncol(aug), pass = qr(aug)$rank == ncol(aug))
}
```

Now the grid. Seeds increment across the six cells exactly as the production
script does, so cell `A / set1_moderate` reproduces the walkthrough above.

```{r}
#| label: grid-run
N_TASKS <- 20000
J <- 4

recover_one <- function(model, set_name, pars, design_seed, sim_seed) {
  design <- make_design(N_TASKS, J, seed = design_seed)
  rc <- id_rank_check(design$X)
  cut_true <- if (model == "A") alpha_to_cut(pars$alphaA) else pars$cutB
  dat <- simulate_dual(design, pars$beta, cut_true, model = model, seed = sim_seed)
  fit <- fit_dual_mle(dat)

  truth <- c(pars$beta, cut_true)
  est   <- c(fit$beta, fit$cut)
  se    <- c(fit$se_beta, fit$se_cut)
  tab <- data.frame(param = c(names(pars$beta), paste0("c", 1:(W - 1))),
                    truth = truth, est = est, se = se, z = (est - truth) / se)
  if (model == "A") {
    tab <- rbind(tab, data.frame(
      param = paste0("alpha", 1:(W - 1)),
      truth = pars$alphaA, est = fit$alpha, se = fit$se_alpha,
      z = (fit$alpha - pars$alphaA) / fit$se_alpha))
  }
  rownames(tab) <- NULL
  list(model = model, set = set_name, rank_check = rc, table = tab,
       y_freq = tabulate(dat$y, nbins = W) / N_TASKS,
       nll = fit$nll, convergence = fit$convergence, grad_max = fit$grad_max,
       min_eig = fit$min_eig, cond = fit$cond,
       nll_truth = negloglik(c(pars$beta, cut_to_par(cut_true)), dat))
}

results <- list()
seed_counter <- 0
for (model in c("A", "B")) {
  for (s in seq_along(param_sets)) {
    seed_counter <- seed_counter + 1
    nm <- names(param_sets)[s]
    results[[paste(model, nm, sep = "_")]] <-
      recover_one(model, nm, param_sets[[s]],
                  design_seed = 100 + seed_counter, sim_seed = 500 + seed_counter)
  }
}
```

### Grid summary

```{r}
#| label: grid-summary
grid_tab <- do.call(rbind, lapply(results, function(r) {
  z_free <- r$table$z[!grepl("alpha", r$table$param)]
  lr <- 2 * (r$nll_truth - r$nll)
  data.frame(
    cell          = paste(r$model, r$set),
    `max |z|`     = round(max(abs(z_free)), 2),
    `LR vs truth` = round(lr, 2),
    `chi2 .95`    = round(qchisq(0.95, 9), 2),
    `inside`      = lr < qchisq(0.95, 9),
    `min eig(H)`  = signif(r$min_eig, 3),
    check.names = FALSE)
}))
knitr::kable(grid_tab, row.names = FALSE)
```

Every cell puts the truth inside the 95% likelihood-ratio region, and every
observed information matrix is strongly positive definite. Across all six fits:

```{r}
#| label: grid-z
zs <- unlist(lapply(results, function(r) r$table$z[!grepl("alpha", r$table$param)]))
cat(sprintf("%d free parameters over 6 fits: max|z| = %.2f, share |z| > 1.96 = %.3f (expect ~0.05)\n",
            length(zs), max(abs(zs)), mean(abs(zs) > 1.96)))
```

The stress configuration is worth looking at directly, since it is the one the
article quotes. A true $\alpha_1$ of 0.02 recovers to:

```{r}
#| label: stress-alpha
st <- results$A_set3_stress$table
knitr::kable(st[grepl("alpha", st$param), ], row.names = FALSE, digits = 4)
```

## 10. The negative control: identification made to fail

@prp-ident requires $[X, \iota]$ to have full column rank $P + 1$. The canonical
violation is a constant common to every inside good. Adding one leaves the
model's *probabilities* untouched, because shifting every $V_j$ by $\delta$
shifts $\overline{\mu}$ by exactly $\delta$, which the cut points absorb one for
one. What breaks is the ability to separate the two.

```{r}
#| label: nc-rank
pars     <- param_sets$set1_moderate
beta_aug <- c(pars$beta, const = 0.5)

design_nc <- make_design(N_TASKS, J, seed = 42, intercept = TRUE)
rc_nc     <- id_rank_check(design_nc$X)
cat(sprintf("rank([X,1]) = %d, required %d -> %s\n",
            rc_nc$rank, rc_nc$required, ifelse(rc_nc$pass, "PASS", "FAIL")))

dat_nc <- simulate_dual(design_nc, beta_aug, pars$cutB, model = "B", seed = 4242)
```

Fit it twice: once from the default start, once displaced two units along the
invariance ridge.

```{r}
#| label: nc-fits
P_nc <- design_nc$P

fit1 <- fit_dual_mle(dat_nc)

start2 <- fit1$par
start2[P_nc]     <- start2[P_nc] + 2       # the intercept coefficient
start2[P_nc + 1] <- start2[P_nc + 1] + 2   # and c1, moving with it
fit2 <- fit_dual_mle(dat_nc, start = start2)

nc_summary <- data.frame(
  quantity = c("nll", "min eigenvalue(H)", "cond(H)", "beta_const",
               paste0("c", 1:4), paste0("c", 1:4, " - beta_const")),
  start1 = c(fit1$nll, fit1$min_eig, fit1$cond, fit1$beta[P_nc], fit1$cut,
             fit1$cut - fit1$beta[P_nc]),
  start2 = c(fit2$nll, fit2$min_eig, fit2$cond, fit2$beta[P_nc], fit2$cut,
             fit2$cut - fit2$beta[P_nc]),
  truth  = c(NA, NA, NA, 0.5, pars$cutB, pars$cutB - 0.5)
)
knitr::kable(nc_summary, row.names = FALSE, digits = 5)
```

Read that table in three parts.

The **log-likelihood** is the same to many decimals from both starts: the two
answers fit equally well. The **smallest eigenvalue** of the observed
information has collapsed, which is the numerical signature of a flat direction.
The **raw** intercept and cut points differ between the starts, by roughly the
two units of displacement, while the **differences** $c_w - \beta_{\text{const}}$
agree with each other and with the truth.

The non-constant coefficients are unaffected:

```{r}
#| label: nc-betas
knitr::kable(
  rbind(start1 = fit1$beta[1:5], start2 = fit2$beta[1:5], truth = pars$beta),
  digits = 4)
```

This is the proposition made concrete. The model is identified up to a single
location, and the design condition is exactly what rules that location out.

## 11. The factorization is exact

@eq-joint claims the joint probability of the two responses factors into a
multinomial logit term and an ordinal term. That claim rests on @lem-gumbel(ii),
the independence of $u^*$ from $j^*$. Here it is tested directly: one fixed
task, one million behavioral draws, empirical joint distribution against the
product form.

```{r}
#| label: factorization
X0 <- rbind(
  c(1, 0, 0, 0, 1.0),
  c(0, 1, 1, 0, 2.0),
  c(0, 0, 0, 1, 0.8),
  c(0, 0, 1, 0, 1.6)
)
M <- 1e6

fact_check <- function(model) {
  beta <- param_sets$set1_moderate$beta
  cut  <- if (model == "A") alpha_to_cut(param_sets$set1_moderate$alphaA)
          else param_sets$set1_moderate$cutB
  V <- as.numeric(X0 %*% beta)

  set.seed(777)
  u      <- matrix(V, M, 4, byrow = TRUE) + matrix(rgumbel(M * 4), M, 4)
  jstar  <- max.col(u, ties.method = "first")
  ustar  <- u[cbind(seq_len(M), jstar)]
  latent <- if (model == "A") ustar else ustar - rgumbel(M)
  y      <- findInterval(latent, cut) + 1L

  emp <- table(factor(jstar, 1:4), factor(y, 1:W)) / M

  # The closed form: outer product of the two marginals.
  mnl   <- exp(V) / sum(exp(V))
  mubar <- log(sum(exp(V)))
  pord  <- ord_prob(rep(mubar, W), cut, 1:W, model)
  theo  <- outer(mnl, pord)

  chisq <- M * sum((emp - theo)^2 / theo)
  pval  <- pchisq(chisq, df = 4 * W - 1, lower.tail = FALSE)
  cat(sprintf("Model %s: max |empirical - product| = %.5f (MC noise ~ %.5f); chi-sq(%d) = %.1f, p = %.3f\n",
              model, max(abs(emp - theo)),
              sqrt(max(theo * (1 - theo)) / M), 4 * W - 1, chisq, pval))
  list(model = model, emp = emp, theo = theo, chisq = chisq, pval = pval)
}

fact <- lapply(c("A", "B"), fact_check)
```

Both links pass. The largest cell-wise deviation sits at Monte Carlo noise, and
neither chi-square rejects. Note what this test would catch: if the maximum
utility were correlated with which alternative attained it, the empirical joint
would not be an outer product, and the deviation would be systematic rather than
noise.

## 12. Repeated sampling: bias and standard-error calibration

One dataset per cell shows recovery. Forty independent datasets per link show
whether the estimator's *sampling* behavior matches the asymptotics: is it
unbiased, and are the model-based standard errors the right size?

```{r}
#| label: replication
R_REPS <- 40
N_REP  <- 5000

rep_pars <- list(
  beta   = c(a2 = 0.8, a3 = -0.5, b2 = 0.4, b3 = 1.0, price = -0.9),
  cutB   = c(-1.0, 0.2, 1.2, 2.2),
  alphaA = c(0.10, 0.30, 0.60, 0.85)
)

run_model <- function(model) {
  cut_true <- if (model == "A") alpha_to_cut(rep_pars$alphaA) else rep_pars$cutB
  truth <- c(rep_pars$beta, cut_true)
  est <- se <- matrix(NA_real_, R_REPS, length(truth))
  conv <- integer(R_REPS)
  for (r in seq_len(R_REPS)) {
    design <- make_design(N_REP, J, seed = 10000 + r)
    dat <- simulate_dual(design, rep_pars$beta, cut_true, model = model,
                         seed = 20000 + r)
    fit <- fit_dual_mle(dat)
    est[r, ] <- c(fit$beta, fit$cut)
    se[r, ]  <- c(fit$se_beta, fit$se_cut)
    conv[r]  <- fit$convergence
  }
  bias   <- colMeans(est) - truth
  emp_sd <- apply(est, 2, sd)
  data.frame(
    model = model,
    param = c(names(rep_pars$beta), paste0("c", 1:(W - 1))),
    truth = truth, mean_est = colMeans(est), bias = bias,
    t_bias = bias / (emp_sd / sqrt(R_REPS)), emp_sd = emp_sd,
    mean_se = colMeans(se), se_ratio = colMeans(se) / emp_sd,
    n_nonconv = sum(conv != 0))
}

rep_res <- do.call(rbind, lapply(c("A", "B"), run_model))
rownames(rep_res) <- NULL
knitr::kable(rep_res, row.names = FALSE, digits = 4)
```

```{r}
#| label: replication-summary
cat(sprintf("max |t_bias| = %.2f over %d tests (5%% critical ~ 2.0; Bonferroni ~ 3.0)\n",
            max(abs(rep_res$t_bias)), nrow(rep_res)))
cat(sprintf("standard-error calibration: mean ratio = %.3f (1 = perfect)\n",
            mean(rep_res$se_ratio)))
```

The calibration ratio compares the average model-based standard error to the
empirical standard deviation of the estimates across the forty datasets. At one,
the standard errors are telling the truth.

## 13. Is the residual bias just $O(1/n)$?

One parameter in the table above carries a $t$ statistic near the conventional
threshold. Two explanations compete: ordinary finite-sample MLE bias, which
shrinks like $1/n$, or something structural, which would not shrink. The test is
to quadruple $n$ on fresh seeds and see whether the bias falls by about four.

```{r}
#| label: bias-scaling
N_BIG <- 20000
cut_A <- alpha_to_cut(rep_pars$alphaA)
truth_A <- c(rep_pars$beta, cut_A)

est_big <- matrix(NA_real_, R_REPS, length(truth_A))
for (r in seq_len(R_REPS)) {
  design <- make_design(N_BIG, J, seed = 30000 + r)
  dat <- simulate_dual(design, rep_pars$beta, cut_A, model = "A", seed = 40000 + r)
  fit <- fit_dual_mle(dat)
  est_big[r, ] <- c(fit$beta, fit$cut)
}

bias_big   <- colMeans(est_big) - truth_A
emp_sd_big <- apply(est_big, 2, sd)
bias_res <- data.frame(
  param = c(names(rep_pars$beta), paste0("c", 1:(W - 1))),
  truth = truth_A, mean_est = colMeans(est_big), bias = bias_big,
  t_bias = bias_big / (emp_sd_big / sqrt(R_REPS)), emp_sd = emp_sd_big)
rownames(bias_res) <- NULL
knitr::kable(bias_res, row.names = FALSE, digits = 4)
```

```{r}
#| label: bias-scaling-verdict
b2_small <- rep_res$bias[rep_res$model == "A" & rep_res$param == "b2"]
b2_big   <- bias_res$bias[bias_res$param == "b2"]
cat(sprintf("b2 bias at n = 5,000:  %+.4f\n", b2_small))
cat(sprintf("b2 bias at n = 20,000: %+.4f\n", b2_big))
cat(sprintf("ratio %.1fx (O(1/n) predicts ~4x); max |t_bias| now %.2f\n",
            b2_small / b2_big, max(abs(bias_res$t_bias))))
```

The bias shrinks in proportion to $n$, which is what the finite-sample
explanation predicts and what a structural problem would not do.

## 14. Results written

These three files are read by the article's simulation section.

```{r}
#| label: save
# Quarto executes some passes from the project root and others from this
# file's directory, so anchor the output path on the project marker rather
# than trusting the working directory.
PROJ <- if (file.exists("_quarto.yml")) "." else ".."
OUT  <- file.path(PROJ, "R", "output")
dir.create(OUT, showWarnings = FALSE, recursive = TRUE)

saveRDS(list(
  recovery = results,
  negative_control = list(fit1 = fit1, fit2 = fit2, summary = nc_summary),
  factorization = fact,
  settings = list(N_TASKS = N_TASKS, J = J, W = W, param_sets = param_sets)
), file.path(OUT, "identification_test.rds"))

saveRDS(rep_res,  file.path(OUT, "replication_study.rds"))
saveRDS(bias_res, file.path(OUT, "bias_scaling_check.rds"))

cat("wrote identification_test.rds, replication_study.rds, bias_scaling_check.rds\n")
```

## 15. Related material

| where | what |
|---|---|
| `R/lib/dgp.R` | designs and behavioral simulators |
| `R/lib/likelihood.R` | the joint likelihood and the MLE fitter |
| notebook 02 | the hierarchical sampler, its calibration, and a cross-check against an independent implementation |
| Appendix E | the same aggregate model estimated in Apollo, as a third independent check |
