---
title: "Aggregate Recovery"
subtitle: "Both links, three configurations, and forty independent datasets"
author:
  - name: Dan Yavorsky
  - name: Geoffery Zheng
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

{{< include ../latex_commands.qmd >}}

## What this notebook does

Notebook 01 worked one configuration of Model A end to end. This one asks
whether that was luck.

Three questions, in increasing order of what they would catch:

1. **Does recovery hold across configurations and both links?** One cell showing
   recovery is an anecdote; six is a pattern.
2. **Is the estimator unbiased, and are its standard errors honest?** Recovery on
   a single dataset says nothing about repeated sampling. Forty independent
   datasets per link do.
3. **Is the residual bias the ordinary finite-sample kind?** One parameter comes
   in near the conventional threshold. Ordinary maximum-likelihood bias shrinks
   like $O(1/n)$; a structural problem does not.

The machinery below is the general form of what notebook 01 built: it carries
both links rather than Model A alone, and it is defined once here rather than
twice.

This notebook is the source of `aggregate_recovery.rds`,
`replication_study.rds` and `bias_scaling_check.rds`.

```{r}
#| label: setup
set.seed(1)
options(digits = 5)
```

## 1. The machinery

The general form of what notebook 01 built. Two differences: these carry both
links rather than Model A alone, and they are defined once rather than twice.

```{r}
#| label: machinery-base
#| code-fold: true
#| code-summary: "Designs, the scale maps, and the cut-point parameterization"

# Standard Gumbel quantile function and CDF. Exact inverses; used in both
# directions. rgumbel draws by inverse transform.
alpha_to_cut <- function(a) -log(-log(a))   # probability -> utility,  F^{-1}
cut_to_alpha <- function(c) exp(-exp(-c))   # utility -> probability,  F
rgumbel      <- function(n) alpha_to_cut(runif(n))

make_design <- function(n_tasks, J, seed, intercept = FALSE) {
  set.seed(seed)
  n_rows <- n_tasks * J
  a <- sample(1:3, n_rows, replace = TRUE)
  b <- sample(1:3, n_rows, replace = TRUE)
  price <- runif(n_rows, 0.5, 2.5)
  X <- cbind(a2 = as.numeric(a == 2), a3 = as.numeric(a == 3),
             b2 = as.numeric(b == 2), b3 = as.numeric(b == 3), price = price)
  if (intercept) X <- cbind(X, const = 1)
  list(X = X, n_tasks = n_tasks, J = J, P = ncol(X))
}

# Unconstrained working parameters: (beta, c_1, log gaps). See notebook 01 for
# the derivation and the Jacobian used for delta-method standard errors.
par_to_cut <- function(par, P, W) {
  if (W == 2) return(par[P + 1])
  par[P + 1] + c(0, cumsum(exp(par[(P + 2):(P + W - 1)])))
}
cut_to_par <- function(cut) c(cut[1], log(diff(cut)))
```

## 2. Recovery across configurations and links


The walkthrough above is one cell. The article's claim is broader: recovery
holds across parameter configurations that put the response distribution in
very different places, and for both links. Three configurations, two links, six
fits.

```{r}
#| label: grid-setup
W <- 5

param_sets <- list(
  set1_moderate = list(
    beta   = c(a2 = 0.8, a3 = -0.5, b2 = 0.4, b3 = 1.0, price = -0.9),
    cutB   = c(-1.0, 0.2, 1.2, 2.2),
    alphaA = c(0.10, 0.30, 0.60, 0.85)
  ),
  set2_negative = list(
    beta   = c(a2 = -1.2, a3 = 0.3, b2 = -0.6, b3 = 0.7, price = -1.5),
    cutB   = c(-2.5, -1.2, 0.5, 1.0),
    alphaA = c(0.05, 0.25, 0.55, 0.90)
  ),
  set3_stress = list(
    beta   = c(a2 = 2.0, a3 = -1.8, b2 = 1.2, b3 = -0.7, price = -2.2),
    cutB   = c(-3.5, -2.2, 0.8, 3.0),
    alphaA = c(0.02, 0.10, 0.70, 0.97)
  )
)
```

The `set3_stress` configuration is the demanding one. Its $\alpha_1 = 0.02$ puts
the bottom category at two percent of the probability scale, so only about one
task in fifty lands there and the corresponding cut point has to be pinned down
from little data.

Model B needs the general machinery: a logistic link rather than a Gumbel one.
The only thing that changes is the interval probability.

```{r}
#| label: general-likelihood
# Interval probability G(hi) - G(lo), tail-stable, for either link.
# Model A: G = standard Gumbel CDF   (she reports the probability she holds)
# Model B: G = standard logistic CDF (she resolves, then grades)
ord_prob_z <- function(lo, hi, model) {
  if (model == "B") {
    plogis(hi) * plogis(-lo) * (-expm1(lo - hi))
  } else {
    ea <- exp(-lo); eb <- exp(-hi)
    p <- exp(-eb) * (-expm1(-(ea - eb)))
    p[!is.finite(eb)] <- 0
    p
  }
}

ord_prob <- function(mubar, cut, y, model) {
  caug <- c(-Inf, cut, Inf)
  ord_prob_z(caug[y] - mubar, caug[y + 1L] - mubar, model)
}

row_max <- function(M) do.call(pmax, as.data.frame(M))

negloglik <- function(par, dat) {
  design <- dat$design
  P <- design$P; W <- dat$W
  n <- design$n_tasks; J <- design$J
  beta <- par[1:P]
  cut  <- par_to_cut(par, P, W)

  V    <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  m    <- row_max(V)
  logS <- m + log(rowSums(exp(V - m)))

  ll_choice <- V[cbind(seq_len(n), dat$jstar)] - logS
  ll_ord    <- log(pmax(ord_prob(logS, cut, dat$y, dat$model), 1e-312))
  -(sum(ll_choice) + sum(ll_ord))
}

num_grad <- function(f, x, eps = 1e-6) {
  vapply(seq_along(x), function(k) {
    h <- eps * max(1, abs(x[k]))
    xp <- x; xp[k] <- xp[k] + h
    xm <- x; xm[k] <- xm[k] - h
    (f(xp) - f(xm)) / (2 * h)
  }, numeric(1))
}
```

The simulator likewise generalizes. Model B differs in one line: she resolves
her uncertainty by drawing the outside good, and grades the comparison
$u^* - \eta_0$ rather than reporting $F(u^*)$.

```{r}
#| label: general-sim
simulate_dual <- function(design, beta, cut, model = c("A", "B"), seed) {
  model <- match.arg(model)
  set.seed(seed)
  n <- design$n_tasks; J <- design$J
  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  u <- V + matrix(rgumbel(n * J), n, J)
  jstar <- max.col(u, ties.method = "first")
  ustar <- u[cbind(seq_len(n), jstar)]
  latent <- if (model == "A") ustar else ustar - rgumbel(n)
  y <- findInterval(latent, cut) + 1L
  list(design = design, jstar = jstar, y = y, W = length(cut) + 1L,
       model = model, beta_true = beta, cut_true = cut)
}

fit_dual_mle <- function(dat, start = NULL) {
  design <- dat$design; P <- design$P; W <- dat$W
  if (is.null(start)) {
    freq <- tabulate(dat$y, nbins = W)
    cumq <- pmin(pmax(cumsum(freq)[1:(W - 1)] / sum(freq), 1e-4), 1 - 1e-4)
    ginv <- if (dat$model == "B") qlogis else function(q) -log(-log(q))
    cut0 <- log(design$J) + ginv(cumq)
    if (W > 2) for (k in 2:(W - 1)) cut0[k] <- max(cut0[k], cut0[k - 1] + 1e-3)
    start <- c(rep(0, P), cut_to_par(cut0))
  }
  fn  <- function(p) negloglik(p, dat)
  opt <- optim(start, fn, method = "BFGS",
               control = list(maxit = 1000, reltol = 1e-12))
  H  <- optimHess(opt$par, fn)
  g  <- num_grad(fn, opt$par)
  ev <- eigen(H, symmetric = TRUE, only.values = TRUE)$values

  cut_hat <- par_to_cut(opt$par, P, W)
  Jc <- matrix(0, W - 1, length(opt$par)); Jc[, P + 1] <- 1
  if (W > 2) {
    d <- exp(opt$par[(P + 2):(P + W - 1)])
    for (w in 2:(W - 1)) Jc[w, (P + 2):(P + w)] <- d[1:(w - 1)]
  }
  Vpar <- tryCatch(solve(H), error = function(e)
    matrix(NA, length(opt$par), length(opt$par)))

  out <- list(beta = opt$par[1:P], se_beta = sqrt(diag(Vpar)[1:P]),
              cut = cut_hat, se_cut = sqrt(diag(Jc %*% Vpar %*% t(Jc))),
              nll = opt$value, convergence = opt$convergence,
              grad_max = max(abs(g)), eigen = ev, min_eig = min(ev),
              cond = max(ev) / max(min(ev), .Machine$double.eps),
              par = opt$par, counts = opt$counts)
  if (dat$model == "A") {
    a <- cut_to_alpha(cut_hat)
    out$alpha <- a
    out$se_alpha <- out$se_cut * a * (-log(a))
  }
  out
}

id_rank_check <- function(X) {
  aug <- cbind(X, 1)
  list(rank = qr(aug)$rank, required = ncol(aug), pass = qr(aug)$rank == ncol(aug))
}
```

Now the grid. Seeds increment across the six cells exactly as the production
script does, so cell `A / set1_moderate` reproduces the walkthrough above.

```{r}
#| label: grid-run
N_TASKS <- 20000
J <- 4

recover_one <- function(model, set_name, pars, design_seed, sim_seed) {
  design <- make_design(N_TASKS, J, seed = design_seed)
  rc <- id_rank_check(design$X)
  cut_true <- if (model == "A") alpha_to_cut(pars$alphaA) else pars$cutB
  dat <- simulate_dual(design, pars$beta, cut_true, model = model, seed = sim_seed)
  fit <- fit_dual_mle(dat)

  truth <- c(pars$beta, cut_true)
  est   <- c(fit$beta, fit$cut)
  se    <- c(fit$se_beta, fit$se_cut)
  tab <- data.frame(param = c(names(pars$beta), paste0("c", 1:(W - 1))),
                    truth = truth, est = est, se = se, z = (est - truth) / se)
  if (model == "A") {
    tab <- rbind(tab, data.frame(
      param = paste0("alpha", 1:(W - 1)),
      truth = pars$alphaA, est = fit$alpha, se = fit$se_alpha,
      z = (fit$alpha - pars$alphaA) / fit$se_alpha))
  }
  rownames(tab) <- NULL
  list(model = model, set = set_name, rank_check = rc, table = tab,
       y_freq = tabulate(dat$y, nbins = W) / N_TASKS,
       nll = fit$nll, convergence = fit$convergence, grad_max = fit$grad_max,
       min_eig = fit$min_eig, cond = fit$cond,
       nll_truth = negloglik(c(pars$beta, cut_to_par(cut_true)), dat))
}

results <- list()
seed_counter <- 0
for (model in c("A", "B")) {
  for (s in seq_along(param_sets)) {
    seed_counter <- seed_counter + 1
    nm <- names(param_sets)[s]
    results[[paste(model, nm, sep = "_")]] <-
      recover_one(model, nm, param_sets[[s]],
                  design_seed = 100 + seed_counter, sim_seed = 500 + seed_counter)
  }
}
```

### Grid summary

```{r}
#| label: grid-summary
grid_tab <- do.call(rbind, lapply(results, function(r) {
  z_free <- r$table$z[!grepl("alpha", r$table$param)]
  lr <- 2 * (r$nll_truth - r$nll)
  data.frame(
    cell          = paste(r$model, r$set),
    `max |z|`     = round(max(abs(z_free)), 2),
    `LR vs truth` = round(lr, 2),
    `chi2 .95`    = round(qchisq(0.95, 9), 2),
    `inside`      = lr < qchisq(0.95, 9),
    `min eig(H)`  = signif(r$min_eig, 3),
    check.names = FALSE)
}))
knitr::kable(grid_tab, row.names = FALSE)
```

Every cell puts the truth inside the 95% likelihood-ratio region, and every
observed information matrix is strongly positive definite. Across all six fits:

```{r}
#| label: grid-z
zs <- unlist(lapply(results, function(r) r$table$z[!grepl("alpha", r$table$param)]))
cat(sprintf("%d free parameters over 6 fits: max|z| = %.2f, share |z| > 1.96 = %.3f (expect ~0.05)\n",
            length(zs), max(abs(zs)), mean(abs(zs) > 1.96)))
```

The stress configuration is worth looking at directly, since it is the one the
article quotes. A true $\alpha_1$ of 0.02 recovers to:

```{r}
#| label: stress-alpha
st <- results$A_set3_stress$table
knitr::kable(st[grepl("alpha", st$param), ], row.names = FALSE, digits = 4)
```

## 3. Repeated sampling: bias and standard-error calibration


One dataset per cell shows recovery. Forty independent datasets per link show
whether the estimator's *sampling* behavior matches the asymptotics: is it
unbiased, and are the model-based standard errors the right size?

```{r}
#| label: replication
R_REPS <- 40
N_REP  <- 5000

rep_pars <- list(
  beta   = c(a2 = 0.8, a3 = -0.5, b2 = 0.4, b3 = 1.0, price = -0.9),
  cutB   = c(-1.0, 0.2, 1.2, 2.2),
  alphaA = c(0.10, 0.30, 0.60, 0.85)
)

run_model <- function(model) {
  cut_true <- if (model == "A") alpha_to_cut(rep_pars$alphaA) else rep_pars$cutB
  truth <- c(rep_pars$beta, cut_true)
  est <- se <- matrix(NA_real_, R_REPS, length(truth))
  conv <- integer(R_REPS)
  for (r in seq_len(R_REPS)) {
    design <- make_design(N_REP, J, seed = 10000 + r)
    dat <- simulate_dual(design, rep_pars$beta, cut_true, model = model,
                         seed = 20000 + r)
    fit <- fit_dual_mle(dat)
    est[r, ] <- c(fit$beta, fit$cut)
    se[r, ]  <- c(fit$se_beta, fit$se_cut)
    conv[r]  <- fit$convergence
  }
  bias   <- colMeans(est) - truth
  emp_sd <- apply(est, 2, sd)
  data.frame(
    model = model,
    param = c(names(rep_pars$beta), paste0("c", 1:(W - 1))),
    truth = truth, mean_est = colMeans(est), bias = bias,
    t_bias = bias / (emp_sd / sqrt(R_REPS)), emp_sd = emp_sd,
    mean_se = colMeans(se), se_ratio = colMeans(se) / emp_sd,
    n_nonconv = sum(conv != 0))
}

rep_res <- do.call(rbind, lapply(c("A", "B"), run_model))
rownames(rep_res) <- NULL
knitr::kable(rep_res, row.names = FALSE, digits = 4)
```

```{r}
#| label: replication-summary
cat(sprintf("max |t_bias| = %.2f over %d tests (5%% critical ~ 2.0; Bonferroni ~ 3.0)\n",
            max(abs(rep_res$t_bias)), nrow(rep_res)))
cat(sprintf("standard-error calibration: mean ratio = %.3f (1 = perfect)\n",
            mean(rep_res$se_ratio)))
```

The calibration ratio compares the average model-based standard error to the
empirical standard deviation of the estimates across the forty datasets. At one,
the standard errors are telling the truth.

## 4. Is the residual bias just $O(1/n)$?


One parameter in the table above carries a $t$ statistic near the conventional
threshold. Two explanations compete: ordinary finite-sample MLE bias, which
shrinks like $1/n$, or something structural, which would not shrink. The test is
to quadruple $n$ on fresh seeds and see whether the bias falls by about four.

```{r}
#| label: bias-scaling
N_BIG <- 20000
cut_A <- alpha_to_cut(rep_pars$alphaA)
truth_A <- c(rep_pars$beta, cut_A)

est_big <- matrix(NA_real_, R_REPS, length(truth_A))
for (r in seq_len(R_REPS)) {
  design <- make_design(N_BIG, J, seed = 30000 + r)
  dat <- simulate_dual(design, rep_pars$beta, cut_A, model = "A", seed = 40000 + r)
  fit <- fit_dual_mle(dat)
  est_big[r, ] <- c(fit$beta, fit$cut)
}

bias_big   <- colMeans(est_big) - truth_A
emp_sd_big <- apply(est_big, 2, sd)
bias_res <- data.frame(
  param = c(names(rep_pars$beta), paste0("c", 1:(W - 1))),
  truth = truth_A, mean_est = colMeans(est_big), bias = bias_big,
  t_bias = bias_big / (emp_sd_big / sqrt(R_REPS)), emp_sd = emp_sd_big)
rownames(bias_res) <- NULL
knitr::kable(bias_res, row.names = FALSE, digits = 4)
```

```{r}
#| label: bias-scaling-verdict
b2_small <- rep_res$bias[rep_res$model == "A" & rep_res$param == "b2"]
b2_big   <- bias_res$bias[bias_res$param == "b2"]
cat(sprintf("b2 bias at n = 5,000:  %+.4f\n", b2_small))
cat(sprintf("b2 bias at n = 20,000: %+.4f\n", b2_big))
cat(sprintf("ratio %.1fx (O(1/n) predicts ~4x); max |t_bias| now %.2f\n",
            b2_small / b2_big, max(abs(bias_res$t_bias))))
```

The bias shrinks in proportion to $n$, which is what the finite-sample
explanation predicts and what a structural problem would not do.

## 5. Results written

```{r}
#| label: save
PROJ <- if (file.exists("_quarto.yml")) "." else ".."
OUT  <- file.path(PROJ, "R", "output")
dir.create(OUT, showWarnings = FALSE, recursive = TRUE)

saveRDS(list(recovery = results,
             settings = list(N_TASKS = N_TASKS, J = J, W = W,
                             param_sets = param_sets)),
        file.path(OUT, "aggregate_recovery.rds"))
saveRDS(rep_res,  file.path(OUT, "replication_study.rds"))
saveRDS(bias_res, file.path(OUT, "bias_scaling_check.rds"))

cat("wrote aggregate_recovery.rds, replication_study.rds, bias_scaling_check.rds\n")
```

## 6. Related material

| where | what |
|---|---|
| notebook 01 | the same model worked end to end on one configuration, with the simplest code that works |
| notebook 03 | the hierarchical estimator: recovery, posterior calibration, and two cross-checks |
| `R/lib/likelihood.R` | the production likelihood and MLE fitter this copies |
