---
title: "The Cost of Dichotomization"
subtitle: "What collapsing the scale to a yes/no answer actually costs"
author:
  - name: Dan Yavorsky
  - name: Geoffery Zheng
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

## What this notebook does

The binary dual-response model is the incumbent. If a practitioner fields a
five-point purchase-likelihood scale and then collapses it ("top-2-box counts as
a buy"), the resulting binary model is **correctly specified**: collapsing a
cumulative-link model at one of its own cut points gives back a cumulative-link
model with one cut point. So dichotomization costs no consistency. What it costs
is precision, and this notebook measures how much, three ways:

1. **Exactly**, from the Fisher information. No simulation noise at all: the
   asymptotic variance of every estimator is computed in closed form across four
   parameter configurations.
2. **By paired Monte Carlo**, fitting the ordinal model and every binary variant
   to the *same* simulated datasets, so nothing differs except the coarsening of
   the second response.
3. **At the individual level**, in the hierarchical model a practitioner would
   actually estimate, where each respondent contributes only twelve tasks and
   pooled cut points can no longer rescue a coarsened second stage.

The third is where the argument turns. The aggregate penalties are real but
modest, and an honest reading of them alone would say dichotomization is a minor
sin. It is not, and the reason is that aggregate prediction variance is
dominated by the forced-choice stage that both estimators share.

This notebook is the source of `dichotomization_study.rds` and
`hb_simstudy_dichot.rds`.

```{r}
#| label: setup
set.seed(1)
options(digits = 5)
```

## 1. Why coarsening is lossy but not biased

Under the model, $\Pr(y \ge k) = 1 - G(c_{k-1} - \overline{\mu})$. Define a
binary variable $b = \mathbb{1}\{y \ge k\}$ and its likelihood is exactly a
binary dual-response model with the single cut point $c_{k-1}$. The parameters
$\bfbeta$ and $c_{k-1}$ are still identified and still consistently estimated.

What is lost is the information in the other $W - 2$ thresholds. The paper's
information decomposition makes this precise: the second response contributes
$\omega(\overline{\mu}; \mathbf{c}) \, \bar{x} \bar{x}'$ to the information
matrix, where

$$
\omega = \sum_{w} \frac{\left( g_w - g_{w-1} \right)^2}{\pi_w}
$$

sums over the categories. Dropping categories drops terms. The quantity
$\omega$ is computable in closed form, so we can say exactly what fraction of
the second response's information each coarsening retains, independent of any
particular design.

## 2. Machinery

Copies of `R/lib/dgp.R`, `R/lib/likelihood.R` and `R/homogeneous/fisher.R`. The
Fisher block is the part worth reading: it implements the paper's information
decomposition directly.

```{r}
#| label: machinery
#| code-fold: true
#| code-summary: "Design, simulator, likelihood, and exact Fisher information"
rgumbel <- function(n) -log(-log(runif(n)))

make_design <- function(n_tasks, J, seed, intercept = FALSE) {
  set.seed(seed)
  n_rows <- n_tasks * J
  a <- sample(1:3, n_rows, replace = TRUE); b <- sample(1:3, n_rows, replace = TRUE)
  price <- runif(n_rows, 0.5, 2.5)
  X <- cbind(a2 = as.numeric(a == 2), a3 = as.numeric(a == 3),
             b2 = as.numeric(b == 2), b3 = as.numeric(b == 3), price = price)
  if (intercept) X <- cbind(X, const = 1)
  list(X = X, n_tasks = n_tasks, J = J, P = ncol(X))
}
id_rank_check <- function(X) {
  aug <- cbind(X, 1); list(rank = qr(aug)$rank, required = ncol(aug),
                           pass = qr(aug)$rank == ncol(aug))
}
par_to_cut <- function(par, P, W) {
  if (W == 2) return(par[P + 1])
  par[P + 1] + c(0, cumsum(exp(par[(P + 2):(P + W - 1)])))
}
cut_to_par <- function(cut) if (length(cut) == 1) cut else c(cut[1], log(diff(cut)))
row_max <- function(M) do.call(pmax, as.data.frame(M))

ord_prob_z <- function(lo, hi, model) {
  if (model == "B") plogis(hi) * plogis(-lo) * (-expm1(lo - hi))
  else { ea <- exp(-lo); eb <- exp(-hi)
         p <- exp(-eb) * (-expm1(-(ea - eb))); p[!is.finite(eb)] <- 0; p }
}
ord_prob <- function(mubar, cut, y, model) {
  caug <- c(-Inf, cut, Inf); ord_prob_z(caug[y] - mubar, caug[y + 1L] - mubar, model)
}
negloglik <- function(par, dat) {
  design <- dat$design; P <- design$P; W <- dat$W
  n <- design$n_tasks; J <- design$J
  beta <- par[1:P]; cut <- par_to_cut(par, P, W)
  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  m <- row_max(V); logS <- m + log(rowSums(exp(V - m)))
  -(sum(V[cbind(seq_len(n), dat$jstar)] - logS) +
    sum(log(pmax(ord_prob(logS, cut, dat$y, dat$model), 1e-312))))
}
num_grad <- function(f, x, eps = 1e-6) {
  vapply(seq_along(x), function(k) {
    h <- eps * max(1, abs(x[k])); xp <- x; xp[k] <- xp[k] + h
    xm <- x; xm[k] <- xm[k] - h; (f(xp) - f(xm)) / (2 * h) }, numeric(1))
}
simulate_dual <- function(design, beta, cut, model = c("A","B"), seed) {
  model <- match.arg(model); set.seed(seed)
  n <- design$n_tasks; J <- design$J
  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  u <- V + matrix(rgumbel(n * J), n, J)
  jstar <- max.col(u, ties.method = "first")
  ustar <- u[cbind(seq_len(n), jstar)]
  latent <- if (model == "A") ustar else ustar - rgumbel(n)
  list(design = design, jstar = jstar, y = findInterval(latent, cut) + 1L,
       W = length(cut) + 1L, model = model, beta_true = beta, cut_true = cut)
}
fit_dual_mle <- function(dat, start = NULL) {
  design <- dat$design; P <- design$P; W <- dat$W
  if (is.null(start)) {
    freq <- tabulate(dat$y, nbins = W)
    cumq <- pmin(pmax(cumsum(freq)[1:(W - 1)] / sum(freq), 1e-4), 1 - 1e-4)
    ginv <- if (dat$model == "B") qlogis else function(q) -log(-log(q))
    cut0 <- log(design$J) + ginv(cumq)
    if (W > 2) for (k in 2:(W - 1)) cut0[k] <- max(cut0[k], cut0[k - 1] + 1e-3)
    start <- c(rep(0, P), cut_to_par(cut0))
  }
  fn <- function(p) negloglik(p, dat)
  opt <- optim(start, fn, method = "BFGS", control = list(maxit = 1000, reltol = 1e-12))
  H <- optimHess(opt$par, fn); ev <- eigen(H, symmetric = TRUE, only.values = TRUE)$values
  cut_hat <- par_to_cut(opt$par, P, W)
  Jc <- matrix(0, W - 1, length(opt$par)); Jc[, P + 1] <- 1
  if (W > 2) { d <- exp(opt$par[(P + 2):(P + W - 1)])
    for (w in 2:(W - 1)) Jc[w, (P + 2):(P + w)] <- d[1:(w - 1)] }
  Vpar <- tryCatch(solve(H), error = function(e) matrix(NA, length(opt$par), length(opt$par)))
  list(beta = opt$par[1:P], se_beta = sqrt(diag(Vpar)[1:P]), cut = cut_hat,
       se_cut = sqrt(diag(Jc %*% Vpar %*% t(Jc))), nll = opt$value,
       convergence = opt$convergence, grad_max = max(abs(num_grad(fn, opt$par))),
       eigen = ev, min_eig = min(ev),
       cond = max(ev) / max(min(ev), .Machine$double.eps), par = opt$par)
}

# ---- exact Fisher information (R/homogeneous/fisher.R) ---------------------
link_cdf <- function(z, model) if (model == "B") plogis(z) else exp(-exp(-z))
link_pdf <- function(z, model) if (model == "B") dlogis(z) else exp(-(z + exp(-z)))

# Information from one task: the MNL block plus the ordinal-stage block.
fisher_task <- function(Xt, beta, cut, model) {
  P <- ncol(Xt); W <- length(cut) + 1L; K <- P + W - 1L
  V <- as.numeric(Xt %*% beta)
  eV <- exp(V - max(V)); p <- eV / sum(eV)
  mubar <- max(V) + log(sum(eV))
  I <- matrix(0, K, K)
  I[1:P, 1:P] <- t(Xt) %*% (diag(p) - tcrossprod(p)) %*% Xt   # choice stage
  z <- cut - mubar
  Gz <- c(0, link_cdf(z, model), 1); gz <- c(0, link_pdf(z, model), 0)
  pi_w <- diff(Gz)
  xbar <- as.numeric(t(p) %*% Xt)
  for (w in seq_len(W)) {                                      # ordinal stage
    s <- numeric(K)
    s[1:P] <- -(gz[w + 1] - gz[w]) / pi_w[w] * xbar
    if (w <= W - 1) s[P + w] <- s[P + w] + gz[w + 1] / pi_w[w]
    if (w >= 2)     s[P + w - 1] <- s[P + w - 1] - gz[w] / pi_w[w]
    I <- I + pi_w[w] * tcrossprod(s)
  }
  I
}
fisher_total <- function(design, beta, cut, model) {
  n <- design$n_tasks; J <- design$J; K <- design$P + length(cut)
  I <- matrix(0, K, K)
  for (t in seq_len(n)) {
    rows <- ((t - 1) * J + 1):(t * J)
    I <- I + fisher_task(design$X[rows, , drop = FALSE], beta, cut, model)
  }
  I
}
# The scalar weight omega: how much the second response contributes per task.
omega_weight <- function(mubar, cut, model) {
  z <- outer(mubar, cut, function(m, c) c - m)
  Gz <- cbind(0, link_cdf(z, model), 1); gz <- cbind(0, link_pdf(z, model), 0)
  W <- length(cut) + 1L
  om <- 0
  for (w in seq_len(W)) om <- om + (gz[, w + 1] - gz[, w])^2 / (Gz[, w + 1] - Gz[, w])
  om
}
```

```{r}
#| label: config
MODEL <- "B"; N_TASKS <- 3600; J <- 4; W <- 5; P <- 5
R_REPS <- 100; N_EVAL <- 200
MC_SETS <- c("set1_moderate", "set4_skewed")

param_sets <- list(
  set1_moderate = list(beta = c(a2=0.8,a3=-0.5,b2=0.4,b3=1.0,price=-0.9),
                       cut = c(-1.0, 0.2, 1.2, 2.2)),
  set2_negative = list(beta = c(a2=-1.2,a3=0.3,b2=-0.6,b3=0.7,price=-1.5),
                       cut = c(-2.5, -1.2, 0.5, 1.0)),
  set3_stress   = list(beta = c(a2=2.0,a3=-1.8,b2=1.2,b3=-0.7,price=-2.2),
                       cut = c(-3.5, -2.2, 0.8, 3.0)),
  set4_skewed   = list(beta = c(a2=0.8,a3=-0.5,b2=0.4,b3=1.0,price=-0.9),
                       cut = c(0.2, 1.2, 2.4, 3.6))
)

design      <- make_design(N_TASKS, J, seed = 3001)
eval_design <- make_design(N_EVAL, J, seed = 777)
stopifnot(id_rank_check(design$X)$pass)
```

`set4_skewed` is the configuration that matters most for practice. Its cut points
push the response distribution downward, producing the bottom-heavy purchase
intent that fielded scales actually show, and leaving the top-2-box convention
badly placed.

```{r}
#| label: helpers
mubar_of <- function(des, beta) {
  V <- matrix(des$X %*% beta, nrow = des$n_tasks, ncol = des$J, byrow = TRUE)
  m <- row_max(V); m + log(rowSums(exp(V - m)))
}
xbar_of <- function(des, beta) {
  V <- matrix(des$X %*% beta, nrow = des$n_tasks, ncol = des$J, byrow = TRUE)
  m <- row_max(V); pr <- exp(V - m) / rowSums(exp(V - m))
  out <- matrix(0, des$n_tasks, ncol(des$X))
  for (j in seq_len(des$J))
    out <- out + pr[, j] * des$X[seq(j, by = des$J, length.out = des$n_tasks), , drop = FALSE]
  out
}
exceed_pred <- function(des, beta, cut_scalar)
  1 - link_cdf(cut_scalar - mubar_of(des, beta), MODEL)

# Which position in each estimator's cut vector can predict P(y >= k); NA = it cannot.
cutpos <- function(est, k) switch(est,
  ord5 = k - 1, ord3 = if (k %in% 3:4) k - 2 else NA,
  bin2 = if (k == 2) 1 else NA, bin3 = if (k == 3) 1 else NA,
  bin4 = if (k == 4) 1 else NA, bin5 = if (k == 5) 1 else NA)
est_cut <- function(est, cut_true) switch(est,
  ord5 = cut_true, ord3 = cut_true[2:3], bin2 = cut_true[1],
  bin3 = cut_true[2], bin4 = cut_true[3], bin5 = cut_true[4])
EST <- c("ord5", "ord3", "bin2", "bin3", "bin4", "bin5")

grad_exceed <- function(Xt, beta, cut, pos, npar) {
  V <- as.numeric(Xt %*% beta); m <- max(V)
  mubar <- m + log(sum(exp(V - m))); p <- exp(V - mubar)
  g <- link_pdf(cut[pos] - mubar, MODEL)
  gr <- numeric(npar); gr[1:P] <- g * as.numeric(t(p) %*% Xt); gr[P + pos] <- -g
  gr
}
```

Note what `cutpos` encodes: a binary estimator cut at $k$ can answer questions
about threshold $k$ and no other. A top-box model cannot tell you about top-2-box
demand. This is not an efficiency loss, it is an inability, and it returns in the
pricing counterfactual of notebook 07.

## 3. Exact efficiency, from the Fisher information

```{r}
#| label: analytic
analytic <- list()
for (sn in names(param_sets)) {
  ps <- param_sets[[sn]]
  mu_est_design <- mubar_of(design, ps$beta)
  catprob <- diff(c(0, colMeans(link_cdf(outer(mu_est_design, ps$cut,
                                               function(m, c) c - m), MODEL)), 1))
  u <- colMeans(xbar_of(design, ps$beta)); u <- u / sqrt(sum(u^2))

  Vcov <- list(); sdtab <- list()
  for (e in EST) {
    ct <- est_cut(e, ps$cut)
    Vc <- solve(fisher_total(design, ps$beta, ct, MODEL))
    Vcov[[e]] <- Vc
    Vb <- Vc[1:P, 1:P]
    row <- data.frame(estimator = e, mean_sd_beta = mean(sqrt(diag(Vb))),
                      sd_u_beta = sqrt(t(u) %*% Vb %*% u),
                      mean_omega = mean(omega_weight(mu_est_design, ct, MODEL)))
    for (k in 2:W) {
      pos <- cutpos(e, k)
      if (!is.na(pos)) {
        grads <- vapply(seq_len(N_EVAL), function(t) {
          rows <- ((t - 1) * J + 1):(t * J)
          grad_exceed(eval_design$X[rows, , drop = FALSE], ps$beta, ct, pos, nrow(Vc))
        }, numeric(nrow(Vc)))
        pv <- colSums(grads * (Vc %*% grads))
        gbar <- rowMeans(grads)
        row[[paste0("sd_pred_k", k)]]   <- sqrt(mean(pv))
        row[[paste0("sd_demand_k", k)]] <- sqrt(t(gbar) %*% Vc %*% gbar)
        row[[paste0("sd_c", k - 1)]]    <- sqrt(Vc[P + pos, P + pos])
      } else {
        row[[paste0("sd_pred_k", k)]] <- NA
        row[[paste0("sd_demand_k", k)]] <- NA
        row[[paste0("sd_c", k - 1)]] <- NA
      }
    }
    sdtab[[e]] <- row
  }
  sdtab <- do.call(rbind, sdtab); rownames(sdtab) <- NULL
  analytic[[sn]] <- list(sd = sdtab, u = u, catprob = catprob)
}
```

### Implied response distributions

```{r}
#| label: catprobs
knitr::kable(
  do.call(rbind, lapply(names(param_sets), function(sn) {
    cp <- analytic[[sn]]$catprob
    data.frame(configuration = sn, `1` = cp[1], `2` = cp[2], `3` = cp[3],
               `4` = cp[4], `5` = cp[5], `top-2-box` = sum(cp[4:5]),
               check.names = FALSE)})),
  row.names = FALSE, digits = 3)
```

### What each coarsening costs

```{r}
#| label: analytic-table
eff_tab <- do.call(rbind, lapply(names(param_sets), function(sn) {
  a <- analytic[[sn]]$sd
  data.frame(configuration = sn,
             `top-2-box, beta` = round((a$mean_sd_beta[5] / a$mean_sd_beta[1])^2, 3),
             `worst cut, beta` = round(max((a$mean_sd_beta / a$mean_sd_beta[1])^2, na.rm = TRUE), 3),
             `omega share, top-2-box` = round(a$mean_omega[5] / a$mean_omega[1], 2),
             `omega share, worst` = round(min(a$mean_omega[-1] / a$mean_omega[1], na.rm = TRUE), 2),
             check.names = FALSE)}))
knitr::kable(eff_tab, row.names = FALSE)
```

Two columns, two different things. The **beta** columns are variance ratios: a
value of 1.05 means the binary analyst needs five percent more sample for the
same precision on preference parameters. The **omega share** columns say what
fraction of the second response's *own* information survives the coarsening,
and those numbers are much worse. The gap between them exists because $\bfbeta$
is mostly identified by the forced choice, which both estimators keep intact.

The analyst must commit to the cut before seeing the data, so the "worst cut"
column is the relevant risk, not the best case.

### Full detail, per configuration

```{r}
#| label: analytic-detail
for (sn in names(param_sets)) {
  cat("\n###", sn, "\n")
  a <- analytic[[sn]]$sd
  ratio <- function(col) round((a[[col]] / a[[col]][1])^2, 3)
  print(knitr::kable(
    data.frame(estimator = a$estimator,
               beta_avg = round((a$mean_sd_beta / a$mean_sd_beta[1])^2, 3),
               beta_u_dir = ratio("sd_u_beta"),
               omega_share = round(a$mean_omega / a$mean_omega[1], 3),
               pred_k4 = ratio("sd_pred_k4"),
               demand_k4 = ratio("sd_demand_k4")),
    row.names = FALSE))
}
```

`beta_u_dir` is the variance in the "average attractiveness" direction, the
rank-one direction the ordinal stage informs and the forced choice identifies
worst. It is consistently the most damaged, which is exactly what the theory
predicts: that is where the second response carries information the first does
not.

## 4. Monte Carlo validation

The analytic numbers are asymptotic. Do they describe finite samples? Fit every
estimator to the same simulated datasets and compare empirical spread to the
predicted spread.

```{r}
#| label: monte-carlo
collapse3 <- function(y) 1L + (y >= 3L) + (y >= 4L)
mc <- list()
for (sn in MC_SETS) {
  ps <- param_sets[[sn]]
  mubar_eval_true <- mubar_of(eval_design, ps$beta)
  true_ex <- sapply(2:W, function(k) 1 - link_cdf(ps$cut[k - 1] - mubar_eval_true, MODEL))

  beta_est <- array(NA_real_, c(R_REPS, P, length(EST)),
                    dimnames = list(NULL, names(ps$beta), EST))
  c_own <- array(NA_real_, c(R_REPS, 4, length(EST)),
                 dimnames = list(NULL, paste0("c", 1:4), EST))
  demand <- array(NA_real_, c(R_REPS, W - 1, length(EST)),
                  dimnames = list(NULL, paste0("k", 2:W), EST))
  pred_sse <- matrix(0, length(EST), W - 1, dimnames = list(EST, paste0("k", 2:W)))
  pred_n <- pred_sse

  for (r in seq_len(R_REPS)) {
    dat5 <- simulate_dual(design, ps$beta, ps$cut, model = MODEL, seed = 60000 + r)
    fits <- list()
    fits$ord5 <- fit_dual_mle(dat5)
    dat3 <- dat5; dat3$y <- collapse3(dat5$y); dat3$W <- 3L
    fits$ord3 <- fit_dual_mle(dat3)
    for (k in 2:W) {
      datb <- dat5; datb$y <- 1L + (dat5$y >= k); datb$W <- 2L
      fits[[paste0("bin", k)]] <- fit_dual_mle(datb)
    }
    for (e in EST) {
      beta_est[r, , e] <- fits[[e]]$beta
      for (k in 2:W) {
        pos <- cutpos(e, k); if (is.na(pos)) next
        c_own[r, k - 1, e] <- fits[[e]]$cut[pos]
        pr <- exceed_pred(eval_design, fits[[e]]$beta, fits[[e]]$cut[pos])
        demand[r, k - 1, e] <- mean(pr)
        pred_sse[e, k - 1] <- pred_sse[e, k - 1] + mean((pr - true_ex[, k - 1])^2)
        pred_n[e, k - 1] <- pred_n[e, k - 1] + 1
      }
    }
  }
  pred_rmse <- sqrt(pred_sse / pmax(pred_n, 1)); pred_rmse[pred_n == 0] <- NA
  vr <- sapply(2:W, function(k)
    (pred_rmse[paste0("bin", k), k - 1] / pred_rmse["ord5", k - 1])^2)
  names(vr) <- paste0("k", 2:W)
  mc[[sn]] <- list(beta_est = beta_est, c_own = c_own, demand = demand,
                   pred_rmse = pred_rmse, pred_var_ratio = vr, true_ex = true_ex)
}
```

### Empirical versus analytic

```{r}
#| label: mc-validate
for (sn in MC_SETS) {
  cat("\n###", sn, "-- cut-point sd, empirical vs analytic\n")
  rows <- list()
  for (k in 2:W) for (e in EST) {
    if (!is.na(cutpos(e, k))) {
      rows[[length(rows) + 1]] <- data.frame(
        cut = paste0("c", k - 1), estimator = e,
        empirical = round(sd(mc[[sn]]$c_own[, k - 1, e]), 4),
        analytic = round(analytic[[sn]]$sd[[paste0("sd_c", k - 1)]][match(e, EST)], 4))
    }
  }
  print(knitr::kable(do.call(rbind, rows), row.names = FALSE))
}
```

The two columns track each other closely, which validates both the Fisher
calculation and the standard-error formulas.

### Is anything biased?

```{r}
#| label: mc-bias
bias_tab <- do.call(rbind, lapply(MC_SETS, function(sn) {
  ps <- param_sets[[sn]]
  data.frame(configuration = sn,
    t(sapply(EST, function(e) {
      b <- colMeans(mc[[sn]]$beta_est[, , e]) - ps$beta
      max(abs(b / (apply(mc[[sn]]$beta_est[, , e], 2, sd) / sqrt(R_REPS))))
    })))}))
knitr::kable(bias_tab, row.names = FALSE, digits = 2)
```

Maximum $|t|$ statistics for bias, by estimator. Nothing systematic anywhere,
confirming that dichotomization costs efficiency and not consistency. This is
the claim that lets the rest of the comparison be about precision alone.

## 5. The individual level, where it matters

Everything above pools 3,600 tasks and estimates one parameter vector. That is
not what practitioners do. They estimate a hierarchical model in which each
respondent contributes twelve tasks, and their deliverable is individual-level
part-worths and individual purchase probabilities.

At twelve tasks per person, there is no pooling to rescue the coarsened second
stage.

```{r}
#| label: hb-machinery
#| code-fold: true
#| code-summary: "Panel simulation and the hierarchical sampler"
make_panel_design <- function(N, T_tasks, J, seed, intercept = FALSE) {
  des <- make_design(N * T_tasks, J, seed = seed, intercept = intercept)
  des$N <- N; des$T_tasks <- T_tasks
  des$resp <- rep(seq_len(N), each = T_tasks)
  des$resp_row <- rep(des$resp, each = J)
  des
}
draw_betas <- function(N, beta_bar, Sigma, seed) {
  set.seed(seed); P <- length(beta_bar)
  sweep(matrix(rnorm(N * P), N, P) %*% chol(Sigma), 2, beta_bar, "+")
}
simulate_dual_hb <- function(design, Bmat, cut, model = c("A","B"), seed) {
  model <- match.arg(model); set.seed(seed)
  n <- design$n_tasks; J <- design$J; N <- design$N
  cutmat <- if (is.matrix(cut)) cut else matrix(cut, N, length(cut), byrow = TRUE)
  W <- ncol(cutmat) + 1L
  Bx <- Bmat[design$resp_row, , drop = FALSE]
  V <- matrix(rowSums(design$X * Bx), nrow = n, ncol = J, byrow = TRUE)
  u <- V + matrix(rgumbel(n * J), n, J)
  jstar <- max.col(u, ties.method = "first")
  ustar <- u[cbind(seq_len(n), jstar)]
  latent <- if (model == "A") ustar else ustar - rgumbel(n)
  y <- 1L + rowSums(latent >= cutmat[design$resp, , drop = FALSE])
  list(design = design, resp = design$resp, resp_row = design$resp_row, N = N,
       jstar = jstar, y = as.integer(y), W = W, model = model,
       Bmat_true = Bmat, cut_true = cut)
}
split_holdout <- function(dat, n_holdout) {
  des <- dat$design; T_tasks <- des$T_tasks
  keep_t <- rep(seq_len(T_tasks) <= T_tasks - n_holdout, times = des$N)
  sub <- function(keep) {
    rows <- rep(keep, each = des$J); d <- des
    d$X <- des$X[rows, , drop = FALSE]; d$n_tasks <- sum(keep)
    d$T_tasks <- d$n_tasks / des$N; d$resp <- des$resp[keep]
    d$resp_row <- rep(d$resp, each = des$J)
    out <- dat; out$design <- d; out$resp <- d$resp; out$resp_row <- d$resp_row
    out$jstar <- dat$jstar[keep]; out$y <- dat$y[keep]; out
  }
  list(est = sub(keep_t), holdout = sub(!keep_t))
}
phi_to_cutmat <- function(Phi, P, W) {
  c1 <- Phi[, P + 1]
  if (W == 2) return(matrix(c1, ncol = 1))
  D <- exp(Phi[, (P + 2):(P + W - 1), drop = FALSE])
  cm <- matrix(0, nrow(Phi), W - 1); cm[, 1] <- c1
  for (w in 2:(W - 1)) cm[, w] <- cm[, w - 1] + D[, w - 1]
  cm
}
resp_loglik_all <- function(dat, Bmat, cutmat, choice_only = FALSE) {
  des <- dat$design; n <- des$n_tasks; J <- des$J
  Bx <- Bmat[dat$resp_row, , drop = FALSE]
  V <- matrix(rowSums(des$X * Bx), nrow = n, ncol = J, byrow = TRUE)
  m <- row_max(V); logS <- m + log(rowSums(exp(V - m)))
  ll <- V[cbind(seq_len(n), dat$jstar)] - logS
  if (!choice_only) {
    ca <- cbind(-Inf, cutmat, Inf)
    ll <- ll + log(pmax(ord_prob_z(ca[cbind(dat$resp, dat$y)] - logS,
                                   ca[cbind(dat$resp, dat$y + 1L)] - logS,
                                   dat$model), 1e-312))
  }
  as.numeric(rowsum(ll, dat$resp, reorder = TRUE))
}
default_hb_priors <- function(d) list(phibar0 = rep(0, d), A = 1/100,
  nu = d + 3, V0 = (d + 3) * diag(d), zeta_prec = 1/100)
draw_phibar <- function(Phi, Sigma, pr) {
  N <- nrow(Phi); Si <- chol2inv(chol(Sigma))
  Vb <- chol2inv(chol(N * Si + pr$A * diag(ncol(Phi))))
  m <- Vb %*% (Si %*% (N * colMeans(Phi)) + pr$A * pr$phibar0)
  as.numeric(m + t(chol(Vb)) %*% rnorm(ncol(Phi)))
}
draw_sigma <- function(Phi, phibar, pr) {
  Vn <- pr$V0 + crossprod(sweep(Phi, 2, phibar))
  chol2inv(chol(rWishart(1, pr$nu + nrow(Phi), chol2inv(chol(Vn)))[,,1]))
}
fit_dual_hb <- function(dat, mcmc = list(R=30000, burn=10000, thin=10),
                        het_cut = FALSE, choice_only = FALSE, priors = NULL,
                        keep_phi = TRUE, seed = NULL, verbose = TRUE) {
  if (!is.null(seed)) set.seed(seed)
  des <- dat$design; N <- dat$N; P <- des$P; W <- dat$W
  dim_phi <- if (het_cut) P + W - 1 else P
  if (is.null(priors)) priors <- default_hb_priors(dim_phi)
  R_iter <- mcmc$R; burn <- mcmc$burn; thin <- mcmc$thin
  nkeep <- floor((R_iter - burn) / thin)
  Phi <- matrix(0, N, dim_phi)
  freq <- tabulate(dat$y, nbins = W)
  cumq <- pmin(pmax(cumsum(freq)[1:(W-1)]/sum(freq), 1e-4), 1-1e-4)
  ginv <- if (dat$model == "B") qlogis else function(q) -log(-log(q))
  cut0 <- log(des$J) + ginv(cumq)
  if (W > 2) for (k in 2:(W-1)) cut0[k] <- max(cut0[k], cut0[k-1] + 1e-3)
  zeta <- cut_to_par(cut0)
  if (het_cut) Phi[, (P+1):(P+W-1)] <- matrix(zeta, N, W-1, byrow = TRUE)
  phibar <- colMeans(Phi); Sigma <- diag(dim_phi)
  cur_cutmat <- if (het_cut) phi_to_cutmat(Phi, P, W) else
    matrix(par_to_cut(zeta, 0, W), N, W-1, byrow = TRUE)
  cur_ll <- resp_loglik_all(dat, Phi[, 1:P, drop=FALSE], cur_cutmat, choice_only)
  s_phi <- rep(2.93/sqrt(dim_phi), N); acc_phi <- rep(0, N)
  s_zeta <- 0.05; acc_zeta <- 0; window <- 100
  keep_betabar <- matrix(NA_real_, nkeep, dim_phi)
  keep_Sigma <- array(NA_real_, c(nkeep, dim_phi, dim_phi))
  keep_cut <- if (!het_cut && !choice_only) matrix(NA_real_, nkeep, W-1) else NULL
  keep_ll <- numeric(nkeep); keep_ll_resp <- matrix(NA_real_, nkeep, N)
  keep_Phi <- if (keep_phi) array(NA_real_, c(nkeep, N, dim_phi)) else NULL
  t0 <- Sys.time(); ki <- 0
  for (it in seq_len(R_iter)) {
    U <- chol(Sigma)
    Prop <- Phi + (matrix(rnorm(N*dim_phi), N, dim_phi) %*% U) * s_phi
    prop_cutmat <- if (het_cut) phi_to_cutmat(Prop, P, W) else cur_cutmat
    prop_ll <- resp_loglik_all(dat, Prop[, 1:P, drop=FALSE], prop_cutmat, choice_only)
    Si <- chol2inv(chol(Sigma))
    dcur <- sweep(Phi, 2, phibar); dprop <- sweep(Prop, 2, phibar)
    la <- (prop_ll - 0.5*rowSums((dprop %*% Si)*dprop)) -
          (cur_ll  - 0.5*rowSums((dcur  %*% Si)*dcur))
    acc <- log(runif(N)) < la
    Phi[acc, ] <- Prop[acc, ]; cur_ll[acc] <- prop_ll[acc]
    if (het_cut && any(acc)) cur_cutmat[acc, ] <- prop_cutmat[acc, , drop=FALSE]
    acc_phi <- acc_phi + acc
    if (!het_cut && !choice_only) {
      zp <- zeta + rnorm(W-1, sd = s_zeta)
      cp <- matrix(par_to_cut(zp, 0, W), N, W-1, byrow = TRUE)
      lp_ll <- resp_loglik_all(dat, Phi[, 1:P, drop=FALSE], cp, choice_only)
      if (log(runif(1)) < (sum(lp_ll) - 0.5*priors$zeta_prec*sum(zp^2)) -
                          (sum(cur_ll) - 0.5*priors$zeta_prec*sum(zeta^2))) {
        zeta <- zp; cur_cutmat <- cp; cur_ll <- lp_ll; acc_zeta <- acc_zeta + 1
      }
    }
    phibar <- draw_phibar(Phi, Sigma, priors)
    Sigma <- draw_sigma(Phi, phibar, priors)
    if (it <= burn && it %% window == 0) {
      s_phi <- pmin(pmax(s_phi*exp(1.5*(acc_phi/window - 0.23)), 0.01), 10)
      acc_phi <- rep(0, N)
      if (!het_cut && !choice_only) {
        s_zeta <- min(max(s_zeta*exp(1.5*(acc_zeta/window - 0.30)), 1e-3), 2)
        acc_zeta <- 0
      }
    }
    if (it == burn) { acc_phi <- rep(0, N); acc_zeta <- 0 }
    if (it > burn && (it - burn) %% thin == 0) {
      ki <- ki + 1
      keep_betabar[ki, ] <- phibar; keep_Sigma[ki, , ] <- Sigma
      if (!is.null(keep_cut)) keep_cut[ki, ] <- par_to_cut(zeta, 0, W)
      keep_ll[ki] <- sum(cur_ll); keep_ll_resp[ki, ] <- cur_ll
      if (keep_phi) keep_Phi[ki, , ] <- Phi
    }
  }
  post <- R_iter - burn
  structure(list(draws = list(phibar = keep_betabar, Sigma = keep_Sigma,
                              cut = keep_cut, Phi = keep_Phi),
                 ll = keep_ll, ll_resp = keep_ll_resp,
                 accept = list(phi = mean(acc_phi/post),
                               zeta = if (het_cut || choice_only) NA else acc_zeta/post),
                 settings = list(mcmc = mcmc, het_cut = het_cut,
                                 choice_only = choice_only, P = P, W = W, N = N,
                                 model = dat$model, priors = priors),
                 runtime_min = as.numeric(difftime(Sys.time(), t0, units="mins"))),
            class = "dr_hb_fit")
}
hb_beta_i <- function(fit)
  apply(fit$draws$Phi[, , 1:fit$settings$P, drop = FALSE], c(2,3), mean)
```

```{r}
#| label: hb-config
N <- 300; T_EST <- 12; T_HOLD <- 2
beta_bar <- c(a2=0.8, a3=-0.5, b2=0.4, b3=1.0, price=-0.9)
Sigma_true <- diag(c(0.6, 0.6, 0.6, 0.6, 0.3))
cut_true <- c(-1.0, 0.2, 1.2, 2.2)
KBUY <- 4              # "purchase" means top-2-box throughout
REPS <- 20
ITER <- 20000
MCMC <- list(R = ITER, burn = floor(ITER * 0.4), thin = 10)

gen_panel <- function(seed, model = "B") {
  des <- make_panel_design(N, T_EST + T_HOLD, J, seed = seed)
  B <- draw_betas(N, beta_bar, Sigma_true, seed = seed + 1)
  split_holdout(simulate_dual_hb(des, B, cut_true, model = model, seed = seed + 2), T_HOLD)
}
task_mubar <- function(design, Bmat) {
  Bx <- Bmat[design$resp_row, , drop = FALSE]
  V <- matrix(rowSums(design$X * Bx), ncol = design$J, byrow = TRUE)
  m <- row_max(V); m + log(rowSums(exp(V - m)))
}
true_exceed_panel <- function(sp) {
  d <- sp$holdout$design
  mu <- task_mubar(d, sp$holdout$Bmat_true)
  cutm <- matrix(sp$holdout$cut_true, N, W - 1, byrow = TRUE)
  1 - plogis(cutm[d$resp, KBUY - 1] - mu)
}
pred_exceed_panel <- function(fit, sp, k = KBUY) {
  d <- sp$holdout$design
  mu <- task_mubar(d, hb_beta_i(fit))
  cutvec <- colMeans(fit$draws$cut)
  cc <- rep(cutvec[min(k - 1, length(cutvec))], length(mu))
  G <- if (fit$settings$model == "B") plogis else function(z) exp(-exp(-z))
  1 - G(cc - mu)
}
hit_rate <- function(fit, sp) {
  d <- sp$holdout$design
  Bx <- hb_beta_i(fit)[d$resp_row, , drop = FALSE]
  V <- matrix(rowSums(d$X * Bx), ncol = d$J, byrow = TRUE)
  mean(max.col(V, ties.method = "first") == sp$holdout$jstar)
}
dichotomize <- function(dat, k = KBUY) { dat$y <- 1L + (dat$y >= k); dat$W <- 2L; dat }
```

Twenty paired replications. Each generates one panel, then fits the ordinal
model and the top-2-box binary model to **the same** forced-choice data, so
every difference is attributable to the coarsened second response.

```{r}
#| label: hb-dichot
m <- matrix(NA_real_, REPS, 8, dimnames = list(NULL, c(
  "rmse_bi_ord","rmse_bi_bin","hit_ord","hit_bin",
  "rmse_ex_ord","rmse_ex_bin","sd_c3_ord","sd_c3_bin")))
for (r in seq_len(REPS)) {
  sp <- gen_panel(seed = 60000 + 10 * r)
  f_ord <- fit_dual_hb(sp$est, mcmc = MCMC, seed = 61000 + r, verbose = FALSE)
  f_bin <- fit_dual_hb(dichotomize(sp$est), mcmc = MCMC, seed = 62000 + r, verbose = FALSE)
  tex <- true_exceed_panel(sp)
  m[r, ] <- c(
    sqrt(mean((hb_beta_i(f_ord) - sp$est$Bmat_true)^2)),
    sqrt(mean((hb_beta_i(f_bin) - sp$est$Bmat_true)^2)),
    hit_rate(f_ord, sp), hit_rate(f_bin, sp),
    sqrt(mean((pred_exceed_panel(f_ord, sp) - tex)^2)),
    sqrt(mean((pred_exceed_panel(f_bin, sp, k = 2) - tex)^2)),
    sd(f_ord$draws$cut[, KBUY - 1]), sd(f_bin$draws$cut[, 1]))
}
res_dichot <- list(metrics = m, mean = colMeans(m), se = apply(m, 2, sd)/sqrt(REPS))
knitr::kable(rbind(mean = res_dichot$mean, mc_se = res_dichot$se), digits = 4)
```

```{r}
#| label: hb-ratios
rm_ <- res_dichot$mean
knitr::kable(data.frame(
  quantity = c("individual part-worth RMSE", "individual purchase-probability RMSE",
               "posterior sd of the shared cut point", "holdout hit rate"),
  ordinal = c(rm_["rmse_bi_ord"], rm_["rmse_ex_ord"], rm_["sd_c3_ord"], rm_["hit_ord"]),
  binary  = c(rm_["rmse_bi_bin"], rm_["rmse_ex_bin"], rm_["sd_c3_bin"], rm_["hit_bin"]),
  `variance ratio` = c((rm_["rmse_bi_bin"]/rm_["rmse_bi_ord"])^2,
                       (rm_["rmse_ex_bin"]/rm_["rmse_ex_ord"])^2,
                       (rm_["sd_c3_bin"]/rm_["sd_c3_ord"])^2, NA),
  check.names = FALSE), row.names = FALSE, digits = 4)
```

Read the last column. The cut point and the individual purchase probabilities,
which are what the dual response exists to measure, degrade substantially. The
holdout hit rate does not move at all, and should not: the first response
carries the ranking information and dichotomization leaves it untouched.

That is the finding. Dichotomization does no harm to what the forced choice
already tells you, and real harm to the one thing the second question was added
to measure.

## 6. Results written

```{r}
#| label: save
# Quarto executes some passes from the project root and others from this
# file's directory, so anchor the output path on the project marker rather
# than trusting the working directory.
PROJ <- if (file.exists("_quarto.yml")) "." else ".."
OUT  <- file.path(PROJ, "R", "output")
dir.create(OUT, showWarnings = FALSE, recursive = TRUE)
saveRDS(list(analytic = analytic, mc = mc,
             settings = list(MODEL = MODEL, N_TASKS = N_TASKS, J = J, W = W,
                             R_REPS = R_REPS, param_sets = param_sets)),
        file.path(OUT, "dichotomization_study.rds"))
saveRDS(list(cell = "dichot", res = res_dichot,
             settings = list(N = N, T_EST = T_EST, T_HOLD = T_HOLD, J = J, W = W,
                             REPS = REPS, ITER = ITER, KBUY = KBUY)),
        file.path(OUT, "hb_simstudy_dichot.rds"))
cat("wrote dichotomization_study.rds and hb_simstudy_dichot.rds\n")
```

## 7. Related material

| where | what |
|---|---|
| notebook 06 | how many scale points to field, and where to place the labels |
| notebook 07 | what these precision losses cost in a pricing decision |
| `R/homogeneous/fisher.R` | the production Fisher information code |
