---
title: "The Hierarchical Sampler"
subtitle: "Recovery, posterior calibration, and two independent cross-checks"
author:
  - name: Dan Yavorsky
  - name: Geoffery Zheng
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

{{< include ../latex_commands.qmd >}}

## What this notebook does

Notebook 02 established that the aggregate estimator works. This one does the
same for the hierarchical Bayesian estimator, which is the one a practitioner
actually runs, and which has more ways to go quietly wrong.

Five checks, in increasing order of how much they would embarrass us:

1. **Recovery.** Do population means, covariances, and cut points come back on
   synthetic panels, for both links?
2. **Calibration.** Do the posterior intervals mean what they say? A sampler can
   recover point estimates and still lie about uncertainty.
3. **Cross-check against an independent implementation.** The forced-choice-only
   special case of our sampler is an ordinary hierarchical MNL, which `bayesm`
   has estimated for two decades. On identical data, do we agree?
4. **The degenerate case.** With no heterogeneity in the data, does the
   hierarchical posterior collapse toward the aggregate MLE?
5. **Marginal likelihood.** How optimistic is the Newton-Raftery estimator that
   this literature reports, measured against bridge sampling?

Checks 1 through 4 back **Parameter Recovery and Identification** in the
simulation study. Check 5 backs the marginal-likelihood appendix.

This notebook is the source of `hb_coverage.rds` and `hb_validation.rds`.

::: {.callout-note}
## On the bayesm comparison

An earlier version of this cross-check ran against a local fork of `bayesm`
with a modified `lgtdata` format, and set no seed. It was therefore not
reproducible by anyone else, including us. The version below uses stock CRAN
`bayesm` with its standard data layout and an explicit seed, so the number it
reports can be checked by a reader.
:::

```{r}
#| label: setup
set.seed(1)
options(digits = 5)
stopifnot(requireNamespace("bayesm", quietly = TRUE))
```

## 1. The model

Each respondent $i$ has her own taste vector, drawn from a population
distribution:

$$
\bfbeta_i \sim \mathcal{N}\left( \betabar, \Sigma \right).
$$

Conditional on $\bfbeta_i$, each of her $T$ tasks generates the same two
responses as before: a forced choice among $J$ profiles, and an ordinal purchase
report read off the task's inclusive value. The cut points are common across
respondents here; the heterogeneous-cut extension is exercised in notebook 06.

The sampler is random-walk Metropolis within Gibbs, three blocks per iteration:

1. every respondent's $\bfbeta_i$ proposed at once, with covariance
   $s_i^2 \Sigma$, accepted or rejected individually;
2. the common cut-point block, proposed on an unconstrained scale that keeps the
   cut points ordered by construction;
3. the population moments $\left( \betabar, \Sigma \right)$, drawn from their
   conjugate normal and inverse-Wishart conditionals.

Step sizes adapt toward target acceptance rates during burn-in only, so the
kept draws come from a fixed transition kernel.

## 2. Machinery

Copies of `R/lib/dgp.R`, `R/lib/likelihood.R`, `R/hierarchical/dgp_panel.R` and
`R/hierarchical/hb.R`. Folded because it is long, not because it is
uninteresting: the sampler is the object under test.

```{r}
#| label: machinery
#| code-fold: true
#| code-summary: "Designs, the behavioral panel simulator, and the sampler"

# ---- designs and shocks (R/lib/dgp.R) --------------------------------------
rgumbel <- function(n) -log(-log(runif(n)))

make_design <- function(n_tasks, J, seed, intercept = FALSE) {
  set.seed(seed)
  n_rows <- n_tasks * J
  a <- sample(1:3, n_rows, replace = TRUE)
  b <- sample(1:3, n_rows, replace = TRUE)
  price <- runif(n_rows, 0.5, 2.5)
  X <- cbind(a2 = as.numeric(a == 2), a3 = as.numeric(a == 3),
             b2 = as.numeric(b == 2), b3 = as.numeric(b == 3), price = price)
  if (intercept) X <- cbind(X, const = 1)
  list(X = X, n_tasks = n_tasks, J = J, P = ncol(X))
}

make_panel_design <- function(N, T_tasks, J, seed, intercept = FALSE) {
  des <- make_design(N * T_tasks, J, seed = seed, intercept = intercept)
  des$N <- N
  des$T_tasks <- T_tasks
  des$resp <- rep(seq_len(N), each = T_tasks)
  des$resp_row <- rep(des$resp, each = J)
  des
}

# ---- cut-point parameterization and link (R/lib/likelihood.R) --------------
par_to_cut <- function(par, P, W) {
  if (W == 2) return(par[P + 1])
  par[P + 1] + c(0, cumsum(exp(par[(P + 2):(P + W - 1)])))
}
cut_to_par <- function(cut) if (length(cut) == 1) cut else c(cut[1], log(diff(cut)))
row_max <- function(M) do.call(pmax, as.data.frame(M))

ord_prob_z <- function(lo, hi, model) {
  if (model == "B") {
    plogis(hi) * plogis(-lo) * (-expm1(lo - hi))
  } else {
    ea <- exp(-lo); eb <- exp(-hi)
    p <- exp(-eb) * (-expm1(-(ea - eb)))
    p[!is.finite(eb)] <- 0
    p
  }
}
ord_prob <- function(mubar, cut, y, model) {
  caug <- c(-Inf, cut, Inf)
  ord_prob_z(caug[y] - mubar, caug[y + 1L] - mubar, model)
}

negloglik <- function(par, dat) {
  design <- dat$design; P <- design$P; W <- dat$W
  n <- design$n_tasks; J <- design$J
  beta <- par[1:P]; cut <- par_to_cut(par, P, W)
  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  m <- row_max(V); logS <- m + log(rowSums(exp(V - m)))
  ll_choice <- V[cbind(seq_len(n), dat$jstar)] - logS
  ll_ord <- log(pmax(ord_prob(logS, cut, dat$y, dat$model), 1e-312))
  -(sum(ll_choice) + sum(ll_ord))
}

fit_dual_mle <- function(dat, start = NULL) {
  design <- dat$design; P <- design$P; W <- dat$W
  if (is.null(start)) {
    freq <- tabulate(dat$y, nbins = W)
    cumq <- pmin(pmax(cumsum(freq)[1:(W - 1)] / sum(freq), 1e-4), 1 - 1e-4)
    ginv <- if (dat$model == "B") qlogis else function(q) -log(-log(q))
    cut0 <- log(design$J) + ginv(cumq)
    if (W > 2) for (k in 2:(W - 1)) cut0[k] <- max(cut0[k], cut0[k - 1] + 1e-3)
    start <- c(rep(0, P), cut_to_par(cut0))
  }
  fn <- function(p) negloglik(p, dat)
  opt <- optim(start, fn, method = "BFGS", control = list(maxit = 1000, reltol = 1e-12))
  list(beta = opt$par[1:P], cut = par_to_cut(opt$par, P, W),
       nll = opt$value, par = opt$par)
}

# ---- panel simulation (R/hierarchical/dgp_panel.R) -------------------------
draw_betas <- function(N, beta_bar, Sigma, seed) {
  set.seed(seed)
  P <- length(beta_bar)
  Z <- matrix(rnorm(N * P), N, P)
  sweep(Z %*% chol(Sigma), 2, beta_bar, "+")
}

simulate_dual_hb <- function(design, Bmat, cut, model = c("A", "B"), seed,
                             shock = c("gumbel", "normal")) {
  model <- match.arg(model); shock <- match.arg(shock)
  set.seed(seed)
  n <- design$n_tasks; J <- design$J; N <- design$N
  cutmat <- if (is.matrix(cut)) cut else matrix(cut, N, length(cut), byrow = TRUE)
  W <- ncol(cutmat) + 1L
  Bx <- Bmat[design$resp_row, , drop = FALSE]
  V <- matrix(rowSums(design$X * Bx), nrow = n, ncol = J, byrow = TRUE)
  eta <- if (shock == "gumbel") rgumbel(n * J) else rnorm(n * J, sd = pi / sqrt(6))
  u <- V + matrix(eta, n, J)
  jstar <- max.col(u, ties.method = "first")
  ustar <- u[cbind(seq_len(n), jstar)]
  latent <- if (model == "A") ustar else ustar - rgumbel(n)
  eff_cut <- cutmat[design$resp, , drop = FALSE]
  y <- 1L + rowSums(latent >= eff_cut)
  list(design = design, resp = design$resp, resp_row = design$resp_row,
       N = N, jstar = jstar, y = as.integer(y), W = W, model = model,
       Bmat_true = Bmat, cut_true = cut)
}

# ---- the sampler (R/hierarchical/hb.R) -------------------------------------
phi_to_cutmat <- function(Phi, P, W) {
  c1 <- Phi[, P + 1]
  if (W == 2) return(matrix(c1, ncol = 1))
  D <- exp(Phi[, (P + 2):(P + W - 1), drop = FALSE])
  cutmat <- matrix(0, nrow(Phi), W - 1)
  cutmat[, 1] <- c1
  for (w in 2:(W - 1)) cutmat[, w] <- cutmat[, w - 1] + D[, w - 1]
  cutmat
}

# N-vector of per-respondent joint log-likelihoods. Vectorized over the whole
# panel: this is what makes the all-respondents-at-once MH step affordable.
resp_loglik_all <- function(dat, Bmat, cutmat, choice_only = FALSE) {
  des <- dat$design; n <- des$n_tasks; J <- des$J
  Bx <- Bmat[dat$resp_row, , drop = FALSE]
  V <- matrix(rowSums(des$X * Bx), nrow = n, ncol = J, byrow = TRUE)
  m <- row_max(V); logS <- m + log(rowSums(exp(V - m)))
  ll <- V[cbind(seq_len(n), dat$jstar)] - logS
  if (!choice_only) {
    cut_aug <- cbind(-Inf, cutmat, Inf)
    lo <- cut_aug[cbind(dat$resp, dat$y)] - logS
    hi <- cut_aug[cbind(dat$resp, dat$y + 1L)] - logS
    ll <- ll + log(pmax(ord_prob_z(lo, hi, dat$model), 1e-312))
  }
  as.numeric(rowsum(ll, dat$resp, reorder = TRUE))
}

default_hb_priors <- function(dim_phi) {
  list(phibar0 = rep(0, dim_phi), A = 1 / 100,
       nu = dim_phi + 3, V0 = (dim_phi + 3) * diag(dim_phi),
       zeta_prec = 1 / 100)
}

draw_phibar <- function(Phi, Sigma, priors) {
  N <- nrow(Phi)
  Sinv <- chol2inv(chol(Sigma))
  prec <- N * Sinv + priors$A * diag(ncol(Phi))
  Vb <- chol2inv(chol(prec))
  m <- Vb %*% (Sinv %*% (N * colMeans(Phi)) + priors$A * priors$phibar0)
  as.numeric(m + t(chol(Vb)) %*% rnorm(ncol(Phi)))
}

draw_sigma <- function(Phi, phibar, priors) {
  R <- sweep(Phi, 2, phibar)
  Vn <- priors$V0 + crossprod(R)
  nun <- priors$nu + nrow(Phi)
  Winv <- chol2inv(chol(Vn))
  Wdraw <- rWishart(1, df = nun, Sigma = Winv)[, , 1]
  chol2inv(chol(Wdraw))
}

fit_dual_hb <- function(dat, mcmc = list(R = 30000, burn = 10000, thin = 10),
                        het_cut = FALSE, choice_only = FALSE, priors = NULL,
                        keep_phi = TRUE, seed = NULL, verbose = TRUE) {
  if (!is.null(seed)) set.seed(seed)
  des <- dat$design; N <- dat$N; P <- des$P; W <- dat$W
  dim_phi <- if (het_cut) P + W - 1 else P
  if (is.null(priors)) priors <- default_hb_priors(dim_phi)
  R_iter <- mcmc$R; burn <- mcmc$burn; thin <- mcmc$thin
  nkeep <- floor((R_iter - burn) / thin)

  Phi <- matrix(0, N, dim_phi)
  freq <- tabulate(dat$y, nbins = W)
  cumq <- pmin(pmax(cumsum(freq)[1:(W - 1)] / sum(freq), 1e-4), 1 - 1e-4)
  ginv <- if (dat$model == "B") qlogis else function(q) -log(-log(q))
  cut0 <- log(des$J) + ginv(cumq)
  if (W > 2) for (k in 2:(W - 1)) cut0[k] <- max(cut0[k], cut0[k - 1] + 1e-3)
  zeta <- cut_to_par(cut0)
  if (het_cut) Phi[, (P + 1):(P + W - 1)] <- matrix(zeta, N, W - 1, byrow = TRUE)

  phibar <- colMeans(Phi); Sigma <- diag(dim_phi)
  cur_cutmat <- if (het_cut) phi_to_cutmat(Phi, P, W)
                else matrix(par_to_cut(zeta, 0, W), N, W - 1, byrow = TRUE)
  cur_ll <- resp_loglik_all(dat, Phi[, 1:P, drop = FALSE], cur_cutmat, choice_only)

  s_phi <- rep(2.93 / sqrt(dim_phi), N); acc_phi <- rep(0, N)
  s_zeta <- 0.05; acc_zeta <- 0; window <- 100

  keep_betabar <- matrix(NA_real_, nkeep, dim_phi)
  keep_Sigma <- array(NA_real_, c(nkeep, dim_phi, dim_phi))
  keep_cut <- if (!het_cut && !choice_only) matrix(NA_real_, nkeep, W - 1) else NULL
  keep_ll <- numeric(nkeep)
  keep_ll_resp <- matrix(NA_real_, nkeep, N)
  keep_Phi <- if (keep_phi) array(NA_real_, c(nkeep, N, dim_phi)) else NULL

  t0 <- Sys.time(); ki <- 0
  for (it in seq_len(R_iter)) {
    # (1) all respondents at once
    U <- chol(Sigma)
    Z <- matrix(rnorm(N * dim_phi), N, dim_phi) %*% U
    Prop <- Phi + Z * s_phi
    prop_cutmat <- if (het_cut) phi_to_cutmat(Prop, P, W) else cur_cutmat
    prop_ll <- resp_loglik_all(dat, Prop[, 1:P, drop = FALSE], prop_cutmat, choice_only)
    Sinv <- chol2inv(chol(Sigma))
    dcur <- sweep(Phi, 2, phibar); dprop <- sweep(Prop, 2, phibar)
    qcur <- rowSums((dcur %*% Sinv) * dcur)
    qprop <- rowSums((dprop %*% Sinv) * dprop)
    log_alpha <- (prop_ll - 0.5 * qprop) - (cur_ll - 0.5 * qcur)
    accept <- log(runif(N)) < log_alpha
    Phi[accept, ] <- Prop[accept, ]
    cur_ll[accept] <- prop_ll[accept]
    if (het_cut && any(accept)) cur_cutmat[accept, ] <- prop_cutmat[accept, , drop = FALSE]
    acc_phi <- acc_phi + accept

    # (2) common cut block
    if (!het_cut && !choice_only) {
      zeta_prop <- zeta + rnorm(W - 1, sd = s_zeta)
      cut_prop <- matrix(par_to_cut(zeta_prop, 0, W), N, W - 1, byrow = TRUE)
      ll_prop <- resp_loglik_all(dat, Phi[, 1:P, drop = FALSE], cut_prop, choice_only)
      lp <- (sum(ll_prop) - 0.5 * priors$zeta_prec * sum(zeta_prop^2)) -
            (sum(cur_ll) - 0.5 * priors$zeta_prec * sum(zeta^2))
      if (log(runif(1)) < lp) {
        zeta <- zeta_prop; cur_cutmat <- cut_prop; cur_ll <- ll_prop
        acc_zeta <- acc_zeta + 1
      }
    }

    # (3) hyperparameters
    phibar <- draw_phibar(Phi, Sigma, priors)
    Sigma <- draw_sigma(Phi, phibar, priors)

    if (it <= burn && it %% window == 0) {
      s_phi <- pmin(pmax(s_phi * exp(1.5 * (acc_phi / window - 0.23)), 0.01), 10)
      acc_phi <- rep(0, N)
      if (!het_cut && !choice_only) {
        s_zeta <- min(max(s_zeta * exp(1.5 * (acc_zeta / window - 0.30)), 1e-3), 2)
        acc_zeta <- 0
      }
    }
    if (it == burn) { acc_phi <- rep(0, N); acc_zeta <- 0 }

    if (it > burn && (it - burn) %% thin == 0) {
      ki <- ki + 1
      keep_betabar[ki, ] <- phibar
      keep_Sigma[ki, , ] <- Sigma
      if (!is.null(keep_cut)) keep_cut[ki, ] <- par_to_cut(zeta, 0, W)
      keep_ll[ki] <- sum(cur_ll)
      keep_ll_resp[ki, ] <- cur_ll
      if (keep_phi) keep_Phi[ki, , ] <- Phi
    }
  }

  post_iters <- R_iter - burn
  structure(list(
    draws = list(phibar = keep_betabar, Sigma = keep_Sigma, cut = keep_cut, Phi = keep_Phi),
    ll = keep_ll, ll_resp = keep_ll_resp,
    accept = list(phi = mean(acc_phi / post_iters),
                  zeta = if (het_cut || choice_only) NA else acc_zeta / post_iters),
    settings = list(mcmc = mcmc, het_cut = het_cut, choice_only = choice_only,
                    P = P, W = W, N = N, model = dat$model, priors = priors),
    runtime_min = as.numeric(difftime(Sys.time(), t0, units = "mins"))
  ), class = "dr_hb_fit")
}

hb_beta_i <- function(fit)
  apply(fit$draws$Phi[, , 1:fit$settings$P, drop = FALSE], c(2, 3), mean)

summary_dr_hb <- function(fit, truth = NULL) {
  pm <- colMeans(fit$draws$phibar); psd <- apply(fit$draws$phibar, 2, sd)
  out <- data.frame(param = paste0("phibar", seq_along(pm)), post_mean = pm, post_sd = psd)
  if (!is.null(fit$draws$cut)) {
    out <- rbind(out, data.frame(param = paste0("c", seq_len(ncol(fit$draws$cut))),
                                 post_mean = colMeans(fit$draws$cut),
                                 post_sd = apply(fit$draws$cut, 2, sd)))
  }
  if (!is.null(truth)) { out$truth <- truth; out$z <- (out$post_mean - truth) / out$post_sd }
  out
}
```

```{r}
#| label: config
N <- 300; T_tasks <- 12; J <- 4; W <- 5
beta_bar   <- c(a2 = 0.8, a3 = -0.5, b2 = 0.4, b3 = 1.0, price = -0.9)
Sigma_true <- diag(c(0.6, 0.6, 0.6, 0.6, 0.3))
cut_true   <- c(-1.0, 0.2, 1.2, 2.2)
P <- length(beta_bar)
MCMC <- list(R = 30000, burn = 10000, thin = 10)
```

Three hundred respondents, twelve tasks each, which is a commercially realistic
study rather than an asymptotic fantasy. The heterogeneity is substantial: a
standard deviation of about 0.77 on coefficients whose means run from 0.4 to 1.0,
so respondents genuinely differ in sign on some attributes.

## 3. Recovery

```{r}
#| label: recovery
out <- list()
for (model in c("B", "A")) {
  des <- make_panel_design(N, T_tasks, J, seed = 200 + (model == "A"))
  B   <- draw_betas(N, beta_bar, Sigma_true, seed = 300 + (model == "A"))
  dat <- simulate_dual_hb(des, B, cut_true, model = model, seed = 400 + (model == "A"))

  cat(sprintf("\n=== Model %s ===\ncategory shares: %s\n", model,
              paste(round(tabulate(dat$y, W) / length(dat$y), 3), collapse = " ")))

  fit <- fit_dual_hb(dat, mcmc = MCMC, seed = 500, verbose = FALSE)
  s <- summary_dr_hb(fit, truth = c(beta_bar, cut_true))
  print(knitr::kable(s, row.names = FALSE, digits = 3))

  Sig_pm <- apply(fit$draws$Sigma, c(2, 3), mean)
  cat("posterior mean diag(Sigma):", round(diag(Sig_pm), 3),
      "| truth:", diag(Sigma_true), "\n")
  cat(sprintf("acceptance: phi %.2f, cut block %.2f | runtime %.1f min\n",
              fit$accept$phi, fit$accept$zeta, fit$runtime_min))

  bi <- hb_beta_i(fit)
  cat(sprintf("individual beta_i posterior-mean RMSE: %.3f (sd of true beta_i: %.3f)\n",
              sqrt(mean((bi - dat$Bmat_true)^2)), sqrt(mean(diag(Sigma_true)))))

  out[[paste0("recovery_", model)]] <- list(
    summary = s, Sigma_pm = Sig_pm, accept = fit$accept,
    runtime = fit$runtime_min, beta_i_rmse = sqrt(mean((bi - dat$Bmat_true)^2)))
}
```

The $z$ column is the posterior mean's distance from the truth in posterior
standard deviations. Values inside about two say the truth sits comfortably
within the posterior.

Two things worth noting beyond the table. The individual-level RMSE is
substantially smaller than the population standard deviation, which is the
hierarchical prior doing its job: twelve tasks alone would not pin down five
coefficients per person, and shrinkage toward the population mean supplies the
rest. And the acceptance rates land near their targets of 0.23 and 0.30, which
is the adaptation working as intended.

## 4. Posterior calibration

Recovery is necessary but not sufficient. A sampler can land on the right answer
and still report intervals that are systematically too narrow, which would make
every inference in the paper overconfident.

The check is coverage. Over repeated datasets drawn from fixed true
hyperparameters, the 50% and 90% central posterior intervals should contain the
truth about 50% and 90% of the time. This is the lightweight cousin of
simulation-based calibration; full SBC would draw the truth from the prior, but
fixed-truth coverage at interior values catches sign errors, scale errors, and
hierarchy bugs, which is what we are worried about.

```{r}
#| label: coverage
REPS_B <- 20
REPS_A <- 8
truth <- c(beta_bar, cut_true)
K <- length(truth)
MCMC_COV <- list(R = 20000, burn = 8000, thin = 12)

run_reps <- function(model, reps, seed0) {
  cover50 <- cover90 <- zstats <- matrix(NA, reps, K)
  for (r in seq_len(reps)) {
    des <- make_panel_design(N, T_tasks, J, seed = seed0 + 3 * r)
    B   <- draw_betas(N, beta_bar, Sigma_true, seed = seed0 + 3 * r + 1)
    dat <- simulate_dual_hb(des, B, cut_true, model = model, seed = seed0 + 3 * r + 2)
    fit <- fit_dual_hb(dat, mcmc = MCMC_COV, seed = seed0 + r,
                       verbose = FALSE, keep_phi = FALSE)
    draws <- cbind(fit$draws$phibar, fit$draws$cut)
    for (k in seq_len(K)) {
      q50 <- quantile(draws[, k], c(0.25, 0.75))
      q90 <- quantile(draws[, k], c(0.05, 0.95))
      cover50[r, k] <- truth[k] >= q50[1] && truth[k] <= q50[2]
      cover90[r, k] <- truth[k] >= q90[1] && truth[k] <= q90[2]
      zstats[r, k]  <- (mean(draws[, k]) - truth[k]) / sd(draws[, k])
    }
  }
  list(cover50 = cover50, cover90 = cover90, z = zstats)
}

res_B <- run_reps("B", REPS_B, seed0 = 7000)
res_A <- run_reps("A", REPS_A, seed0 = 9000)
```

```{r}
#| label: coverage-report
cov_tab <- data.frame(
  model = c("B", "A"),
  reps  = c(REPS_B, REPS_A),
  `50% coverage` = c(mean(res_B$cover50), mean(res_A$cover50)),
  `90% coverage` = c(mean(res_B$cover90), mean(res_A$cover90)),
  `mean |z|` = c(mean(abs(res_B$z)), mean(abs(res_A$z))),
  check.names = FALSE)
knitr::kable(cov_tab, row.names = FALSE, digits = 3)
```

Nominal is 0.50 and 0.90. The mean $|z|$ should sit near 0.8, the expected
absolute value of a standard normal, if the posteriors are honest.

Per-parameter, so that a single badly behaved coordinate cannot hide inside an
average:

```{r}
#| label: coverage-byparam
knitr::kable(
  data.frame(parameter = c(names(beta_bar), paste0("c", 1:4)),
             `B: 90%` = round(colMeans(res_B$cover90), 2),
             `A: 90%` = round(colMeans(res_A$cover90), 2),
             check.names = FALSE),
  row.names = FALSE)
```

## 5. Cross-check against bayesm

Everything so far tests our sampler against our own simulator. A coding error
shared between them would pass every check above. So here the sampler is run in
a mode where an independent implementation exists.

Setting `choice_only = TRUE` drops the ordinal stage, leaving an ordinary
hierarchical multinomial logit. That is exactly what `bayesm`'s
`rhierMnlRwMixture` estimates. Same data, two implementations written by
different people years apart.

```{r}
#| label: bayesm-data
des <- make_panel_design(N, T_tasks, J, seed = 210)
B   <- draw_betas(N, beta_bar, Sigma_true, seed = 310)
dat <- simulate_dual_hb(des, B, cut_true, model = "B", seed = 410)

fit_co  <- fit_dual_hb(dat, mcmc = MCMC, choice_only = TRUE, seed = 510, verbose = FALSE)
ours_mu  <- colMeans(fit_co$draws$phibar)
ours_Sig <- diag(apply(fit_co$draws$Sigma, c(2, 3), mean))
```

`bayesm` wants one row block per task, stacked, with the number of alternatives
given separately.

```{r}
#| label: bayesm-fit
lgtdata <- lapply(seq_len(N), function(i) {
  tasks <- which(des$resp == i)
  Xs <- do.call(rbind, lapply(tasks, function(t)
    des$X[((t - 1) * J + 1):(t * J), , drop = FALSE]))
  list(y = dat$jstar[tasks], X = Xs)
})

set.seed(20260913)   # explicit, so this comparison is reproducible
invisible(capture.output(
  bm <- bayesm::rhierMnlRwMixture(
    Data  = list(p = J, lgtdata = lgtdata),
    Prior = list(ncomp = 1),
    Mcmc  = list(R = 30000, keep = 10, nprint = 0))
))

keep_idx  <- seq(1001, 3000)            # drop the first third as burn-in
mu_draws  <- t(vapply(keep_idx, function(r) bm$nmix$compdraw[[r]][[1]]$mu, numeric(P)))
sig_draws <- t(vapply(keep_idx, function(r) {
  rooti <- bm$nmix$compdraw[[r]][[1]]$rooti
  diag(solve(tcrossprod(rooti)))
}, numeric(P)))
bm_mu  <- colMeans(mu_draws)
bm_Sig <- colMeans(sig_draws)
```

```{r}
#| label: bayesm-compare
comp <- data.frame(param = names(beta_bar), truth = as.numeric(beta_bar),
                   ours = round(ours_mu, 3), bayesm = round(bm_mu, 3),
                   diff = round(ours_mu - bm_mu, 3))
knitr::kable(comp, row.names = FALSE)
cat("diag(Sigma) -- ours:", round(ours_Sig, 3),
    "| bayesm:", round(bm_Sig, 3), "| truth:", diag(Sigma_true), "\n")
cat(sprintf("\nlargest disagreement on any population mean: %.4f\n",
            max(abs(comp$diff))))
```

Both implementations sit the same distance from the truth in the same direction,
which is what two correct samplers looking at one finite dataset should do. The
gap between them is far smaller than the gap between either and the truth.

```{r}
#| label: bayesm-store
out$bayesm_check <- list(comp = comp, ours_Sig = ours_Sig, bm_Sig = bm_Sig)
```

## 6. The degenerate case: no heterogeneity

If the data contain no heterogeneity, the hierarchical posterior should collapse
toward the aggregate MLE. This catches a hierarchy that is inventing variation
that is not there.

```{r}
#| label: sigma0
des  <- make_panel_design(N, T_tasks, J, seed = 220)
B0   <- matrix(beta_bar, N, P, byrow = TRUE)     # every respondent identical
dat0 <- simulate_dual_hb(des, B0, cut_true, model = "B", seed = 420)

fit0 <- fit_dual_hb(dat0, mcmc = MCMC, seed = 520, verbose = FALSE)
dat0_pooled <- list(design = dat0$design, jstar = dat0$jstar, y = dat0$y,
                    W = W, model = "B")
mle0 <- fit_dual_mle(dat0_pooled)

comp0 <- data.frame(
  param = c(names(beta_bar), paste0("c", 1:(W - 1))),
  truth = c(beta_bar, cut_true),
  hb_postmean = round(c(colMeans(fit0$draws$phibar), colMeans(fit0$draws$cut)), 3),
  agg_mle = round(c(mle0$beta, mle0$cut), 3))
comp0$diff <- round(comp0$hb_postmean - comp0$agg_mle, 3)
knitr::kable(comp0, row.names = FALSE)

cat("posterior mean diag(Sigma), should be small:",
    round(diag(apply(fit0$draws$Sigma, c(2, 3), mean)), 3), "\n")
out$sigma0_check <- comp0
```

## 7. Marginal likelihood: how optimistic is Newton-Raftery?

This section backs the marginal-likelihood appendix rather than the simulation
study. Model comparisons in this literature conventionally report the
Newton-Raftery harmonic-mean estimator. It is known to be optimistic. In the
aggregate model the parameter space is small enough to measure by how much,
against bridge sampling and a Laplace approximation.

```{r}
#| label: lmd-machinery
#| code-fold: true
#| code-summary: "Aggregate posterior sampler and the three LMD estimators"
logmeanexp <- function(x) { m <- max(x); m + log(mean(exp(x - m))) }

lmd_nr <- function(ll_draws, trim = 0) {
  if (trim > 0) ll_draws <- ll_draws[ll_draws >= quantile(ll_draws, trim)]
  -logmeanexp(-ll_draws)
}

fit_dual_bayes_agg <- function(dat, iter = 20000, burn = 5000, thin = 5,
                               tau2 = 100, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  mle <- fit_dual_mle(dat)
  d <- length(mle$par)
  H <- optimHess(mle$par, function(p) negloglik(p, dat))
  Lp <- t(chol(chol2inv(chol(H)))) * (2.4 / sqrt(d))
  lpost <- function(p) -negloglik(p, dat) - 0.5 * sum(p^2) / tau2
  cur <- mle$par; cur_lp <- lpost(cur)
  nkeep <- floor((iter - burn) / thin)
  draws <- matrix(NA_real_, nkeep, d); lp_draws <- numeric(nkeep)
  acc <- 0; ki <- 0; sc <- 1
  for (it in seq_len(iter)) {
    prop <- cur + as.numeric(Lp %*% rnorm(d)) * sc
    prop_lp <- lpost(prop)
    if (log(runif(1)) < prop_lp - cur_lp) { cur <- prop; cur_lp <- prop_lp; acc <- acc + 1 }
    if (it <= burn && it %% 200 == 0) {
      sc <- min(max(sc * exp(1.5 * (acc / 200 - 0.234)), 0.1), 10); acc <- 0
    }
    if (it == burn) acc <- 0
    if (it > burn && (it - burn) %% thin == 0) {
      ki <- ki + 1; draws[ki, ] <- cur; lp_draws[ki] <- cur_lp
    }
  }
  list(draws = draws, lp = lp_draws, accept = acc / (iter - burn),
       mle = mle, tau2 = tau2, dat = dat)
}

bridge_lmd_agg <- function(fit_agg) {
  dat <- fit_agg$dat; tau2 <- fit_agg$tau2
  lpost_fn <- function(pars, data) {
    -negloglik(pars, data) - 0.5 * sum(pars^2) / tau2 -
      0.5 * length(pars) * log(2 * pi * tau2)
  }
  samples <- fit_agg$draws
  colnames(samples) <- paste0("p", seq_len(ncol(samples)))
  lb <- rep(-Inf, ncol(samples)); ub <- rep(Inf, ncol(samples))
  names(lb) <- names(ub) <- colnames(samples)
  bridgesampling::bridge_sampler(samples = samples, log_posterior = lpost_fn,
                                 data = dat, lb = lb, ub = ub, silent = TRUE)$logml
}

lmd_nr_agg <- function(fit_agg, trim = 0) {
  ll <- apply(fit_agg$draws, 1, function(p) -negloglik(p, fit_agg$dat))
  lmd_nr(ll, trim = trim)
}

lmd_laplace_agg <- function(fit_agg) {
  mle <- fit_agg$mle; d <- length(mle$par)
  H <- optimHess(mle$par, function(p) negloglik(p, fit_agg$dat))
  lprior <- -0.5 * sum(mle$par^2) / fit_agg$tau2 - 0.5 * d * log(2 * pi * fit_agg$tau2)
  -mle$nll + lprior + 0.5 * d * log(2 * pi) -
    0.5 * determinant(H, logarithm = TRUE)$modulus
}

simulate_dual <- function(design, beta, cut, model = c("A", "B"), seed) {
  model <- match.arg(model); set.seed(seed)
  n <- design$n_tasks; J <- design$J
  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  u <- V + matrix(rgumbel(n * J), n, J)
  jstar <- max.col(u, ties.method = "first")
  ustar <- u[cbind(seq_len(n), jstar)]
  latent <- if (model == "A") ustar else ustar - rgumbel(n)
  list(design = design, jstar = jstar, y = findInterval(latent, cut) + 1L,
       W = length(cut) + 1L, model = model)
}
```

```{r}
#| label: lmd-run
des_a <- make_design(3600, J, seed = 230)
dat_a <- simulate_dual(des_a, as.numeric(beta_bar), cut_true, model = "B", seed = 430)
fa <- fit_dual_bayes_agg(dat_a, iter = 30000, burn = 10000, thin = 5, seed = 530)

lmd_tab <- data.frame(
  method = c("bridge", "laplace", "NR", "NR trim 5%", "NR trim 20%"),
  lmd = c(bridge_lmd_agg(fa), lmd_laplace_agg(fa), lmd_nr_agg(fa),
          lmd_nr_agg(fa, 0.05), lmd_nr_agg(fa, 0.20)))
lmd_tab$vs_bridge <- round(lmd_tab$lmd - lmd_tab$lmd[1], 2)
knitr::kable(transform(lmd_tab, lmd = round(lmd, 2)), row.names = FALSE)
out$lmd_check <- lmd_tab
```

Bridge sampling and the Laplace approximation, two methods with nothing in
common, agree closely. Newton-Raftery sits well above both, and trimming the
smallest-likelihood draws makes it worse rather than better, which is the
signature of an estimator whose problem is the tail it is averaging over rather
than a few outliers.

## 8. Results written

```{r}
#| label: save
# Quarto executes some passes from the project root and others from this
# file's directory, so anchor the output path on the project marker rather
# than trusting the working directory.
PROJ <- if (file.exists("_quarto.yml")) "." else ".."
OUT  <- file.path(PROJ, "R", "output")
dir.create(OUT, showWarnings = FALSE, recursive = TRUE)

saveRDS(list(B = res_B, A = res_A, truth = truth,
             settings = list(N = N, T_tasks = T_tasks, MCMC = MCMC_COV)),
        file.path(OUT, "hb_coverage.rds"))

saveRDS(out, file.path(OUT, "hb_validation.rds"))

cat("wrote hb_coverage.rds and hb_validation.rds\n")
```

## 9. Related material

| where | what |
|---|---|
| notebook 02 | the aggregate estimator, identification, and the factorization check |
| notebook 04 | what dichotomizing the scale costs, at the individual level |
| notebook 06 | five misspecifications, including heterogeneous cut points |
| `R/hierarchical/hb.R` | the production sampler this notebook copies |
