---
title: "Misspecification Robustness"
subtitle: "Five ways the model could be wrong, and what each one costs"
author:
  - name: Dan Yavorsky
  - name: Geoffery Zheng
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

## What this notebook does

Every result so far assumes the model is true. This notebook breaks that
assumption five times, in the five ways a sceptical reader would break it, and
measures the damage. In each case the data are generated from something the
model does not describe, and then fit with the model anyway.

| # | the data really are... | the worry |
|---|---|---|
| 1 | generated by the other link | we cannot tell Model A from Model B, so the behavioral story is untestable |
| 2 | generated with respondent-specific cut points | people use scales differently and we pool them |
| 3 | generated with normal rather than Gumbel shocks | the whole derivation rests on Gumbel |
| 4 | generated with correlated alternatives | independence of irrelevant alternatives is false in real choice sets |
| 5 | generated with the outside good drifting across tasks | respondents change as they work through a survey |

A recurring pattern emerges, and it is worth stating before the evidence rather
than after: **the damage lands on the cut points, not on the preference
parameters or the demand predictions built from them.** That is reassuring for
the quantities practitioners use and a genuine caveat for the structural reading
of the scale labels.

This notebook is the source of five results files,
`hb_simstudy_{wronglink,hetcut,nongumbel,nested,drift}.rds`.

```{r}
#| label: setup
set.seed(1)
options(digits = 5)
```

## 1. Machinery

The simulator here is the fullest version: it carries normal shocks, nested
(GEV) shocks, heterogeneous cut points, and two kinds of task-order drift. The
sampler gains the heterogeneous-cut-point mode.

```{r}
#| label: machinery
#| code-fold: true
#| code-summary: "Designs, GEV shocks, the full panel simulator, and the sampler"
rgumbel <- function(n) -log(-log(runif(n)))

# Positive-stable draws (Chambers-Mallows-Stuck). Conditional on a positive
# stable draw, within-nest errors are Gumbel with a common shifted location,
# which reproduces the nested-logit joint distribution while leaving each
# marginal standard Gumbel -- so nested and independent shocks are on the same
# scale and need no variance matching.
rposstable <- function(n, alpha) {
  U <- runif(n, 0, pi); W <- rexp(n)
  sin(alpha * U) / (sin(U))^(1 / alpha) *
    (sin((1 - alpha) * U) / W)^((1 - alpha) / alpha)
}
rnested <- function(n, J, nest) {
  eta <- matrix(NA_real_, n, J)
  for (g in sort(unique(nest$groups))) {
    idx <- which(nest$groups == g); lam <- nest$lambda[g]
    loc <- lam * log(rposstable(n, lam))
    for (j in idx) eta[, j] <- loc + lam * rgumbel(n)
  }
  eta
}

make_design <- function(n_tasks, J, seed, intercept = FALSE) {
  set.seed(seed)
  n_rows <- n_tasks * J
  a <- sample(1:3, n_rows, replace = TRUE); b <- sample(1:3, n_rows, replace = TRUE)
  price <- runif(n_rows, 0.5, 2.5)
  X <- cbind(a2 = as.numeric(a == 2), a3 = as.numeric(a == 3),
             b2 = as.numeric(b == 2), b3 = as.numeric(b == 3), price = price)
  if (intercept) X <- cbind(X, const = 1)
  list(X = X, n_tasks = n_tasks, J = J, P = ncol(X))
}
make_panel_design <- function(N, T_tasks, J, seed, intercept = FALSE) {
  des <- make_design(N * T_tasks, J, seed = seed, intercept = intercept)
  des$N <- N; des$T_tasks <- T_tasks
  des$resp <- rep(seq_len(N), each = T_tasks)
  des$resp_row <- rep(des$resp, each = J)
  des
}
par_to_cut <- function(par, P, W) {
  if (W == 2) return(par[P + 1])
  par[P + 1] + c(0, cumsum(exp(par[(P + 2):(P + W - 1)])))
}
cut_to_par <- function(cut) if (length(cut) == 1) cut else c(cut[1], log(diff(cut)))
row_max <- function(M) do.call(pmax, as.data.frame(M))
ord_prob_z <- function(lo, hi, model) {
  if (model == "B") plogis(hi) * plogis(-lo) * (-expm1(lo - hi))
  else { ea <- exp(-lo); eb <- exp(-hi)
         p <- exp(-eb) * (-expm1(-(ea - eb))); p[!is.finite(eb)] <- 0; p }
}
draw_betas <- function(N, beta_bar, Sigma, seed) {
  set.seed(seed); P <- length(beta_bar)
  sweep(matrix(rnorm(N * P), N, P) %*% chol(Sigma), 2, beta_bar, "+")
}

simulate_dual_hb <- function(design, Bmat, cut, model = c("A","B"), seed,
                             shock = c("gumbel","normal"), drift = NULL, nest = NULL) {
  model <- match.arg(model); shock <- match.arg(shock)
  set.seed(seed)
  n <- design$n_tasks; J <- design$J; N <- design$N
  cutmat <- if (is.matrix(cut)) cut else matrix(cut, N, length(cut), byrow = TRUE)
  W <- ncol(cutmat) + 1L
  Bx <- Bmat[design$resp_row, , drop = FALSE]
  V <- matrix(rowSums(design$X * Bx), nrow = n, ncol = J, byrow = TRUE)
  u <- if (!is.null(nest)) V + rnested(n, J, nest) else {
    eta <- if (shock == "gumbel") rgumbel(n * J) else rnorm(n * J, sd = pi / sqrt(6))
    V + matrix(eta, n, J)
  }
  jstar <- max.col(u, ties.method = "first")
  ustar <- u[cbind(seq_len(n), jstar)]
  m <- row_max(V); logS <- m + log(rowSums(exp(V - m)))
  latent <- if (model == "A") ustar else ustar - rgumbel(n)

  shift <- numeric(n)
  if (!is.null(drift)) {
    T_tasks <- design$T_tasks
    tt <- rep(seq_len(T_tasks), times = N)
    if (drift$type == "linear") shift <- drift$delta * (tt - 1)
    else if (drift$type == "update") {
      psi <- numeric(n)
      for (i in seq_len(N)) {
        rows <- which(design$resp == i)
        for (k in seq_along(rows)[-1])
          psi[rows[k]] <- if (runif(1) < plogis(drift$rho)) logS[rows[k-1]] else psi[rows[k-1]]
      }
      shift <- psi
    }
  }
  eff_cut <- cutmat[design$resp, , drop = FALSE] + shift
  list(design = design, resp = design$resp, resp_row = design$resp_row, N = N,
       jstar = jstar, y = as.integer(1L + rowSums(latent >= eff_cut)), W = W,
       model = model, Bmat_true = Bmat, cut_true = cut, shock = shock,
       drift = drift, nest = nest)
}

split_holdout <- function(dat, n_holdout) {
  des <- dat$design; T_tasks <- des$T_tasks
  keep_t <- rep(seq_len(T_tasks) <= T_tasks - n_holdout, times = des$N)
  sub <- function(keep) {
    rows <- rep(keep, each = des$J); d <- des
    d$X <- des$X[rows, , drop = FALSE]; d$n_tasks <- sum(keep)
    d$T_tasks <- d$n_tasks / des$N; d$resp <- des$resp[keep]
    d$resp_row <- rep(d$resp, each = des$J)
    out <- dat; out$design <- d; out$resp <- d$resp; out$resp_row <- d$resp_row
    out$jstar <- dat$jstar[keep]; out$y <- dat$y[keep]; out
  }
  list(est = sub(keep_t), holdout = sub(!keep_t))
}

phi_to_cutmat <- function(Phi, P, W) {
  c1 <- Phi[, P + 1]
  if (W == 2) return(matrix(c1, ncol = 1))
  D <- exp(Phi[, (P + 2):(P + W - 1), drop = FALSE])
  cm <- matrix(0, nrow(Phi), W - 1); cm[, 1] <- c1
  for (w in 2:(W - 1)) cm[, w] <- cm[, w - 1] + D[, w - 1]
  cm
}
resp_loglik_all <- function(dat, Bmat, cutmat, choice_only = FALSE) {
  des <- dat$design; n <- des$n_tasks; J <- des$J
  Bx <- Bmat[dat$resp_row, , drop = FALSE]
  V <- matrix(rowSums(des$X * Bx), nrow = n, ncol = J, byrow = TRUE)
  m <- row_max(V); logS <- m + log(rowSums(exp(V - m)))
  ll <- V[cbind(seq_len(n), dat$jstar)] - logS
  if (!choice_only) {
    ca <- cbind(-Inf, cutmat, Inf)
    ll <- ll + log(pmax(ord_prob_z(ca[cbind(dat$resp, dat$y)] - logS,
                                   ca[cbind(dat$resp, dat$y + 1L)] - logS,
                                   dat$model), 1e-312))
  }
  as.numeric(rowsum(ll, dat$resp, reorder = TRUE))
}
default_hb_priors <- function(d) list(phibar0 = rep(0, d), A = 1/100,
  nu = d + 3, V0 = (d + 3) * diag(d), zeta_prec = 1/100)
draw_phibar <- function(Phi, Sigma, pr) {
  N <- nrow(Phi); Si <- chol2inv(chol(Sigma))
  Vb <- chol2inv(chol(N * Si + pr$A * diag(ncol(Phi))))
  as.numeric(Vb %*% (Si %*% (N * colMeans(Phi)) + pr$A * pr$phibar0) +
             t(chol(Vb)) %*% rnorm(ncol(Phi)))
}
draw_sigma <- function(Phi, phibar, pr) {
  Vn <- pr$V0 + crossprod(sweep(Phi, 2, phibar))
  chol2inv(chol(rWishart(1, pr$nu + nrow(Phi), chol2inv(chol(Vn)))[,,1]))
}
fit_dual_hb <- function(dat, mcmc = list(R=30000, burn=10000, thin=10),
                        het_cut = FALSE, choice_only = FALSE, priors = NULL,
                        keep_phi = TRUE, seed = NULL, verbose = TRUE) {
  if (!is.null(seed)) set.seed(seed)
  des <- dat$design; N <- dat$N; P <- des$P; W <- dat$W
  dim_phi <- if (het_cut) P + W - 1 else P
  if (is.null(priors)) priors <- default_hb_priors(dim_phi)
  R_iter <- mcmc$R; burn <- mcmc$burn; thin <- mcmc$thin
  nkeep <- floor((R_iter - burn) / thin)
  Phi <- matrix(0, N, dim_phi)
  freq <- tabulate(dat$y, nbins = W)
  cumq <- pmin(pmax(cumsum(freq)[1:(W-1)]/sum(freq), 1e-4), 1-1e-4)
  ginv <- if (dat$model == "B") qlogis else function(q) -log(-log(q))
  cut0 <- log(des$J) + ginv(cumq)
  if (W > 2) for (k in 2:(W-1)) cut0[k] <- max(cut0[k], cut0[k-1] + 1e-3)
  zeta <- cut_to_par(cut0)
  if (het_cut) Phi[, (P+1):(P+W-1)] <- matrix(zeta, N, W-1, byrow = TRUE)
  phibar <- colMeans(Phi); Sigma <- diag(dim_phi)
  cur_cutmat <- if (het_cut) phi_to_cutmat(Phi, P, W) else
    matrix(par_to_cut(zeta, 0, W), N, W-1, byrow = TRUE)
  cur_ll <- resp_loglik_all(dat, Phi[, 1:P, drop=FALSE], cur_cutmat, choice_only)
  s_phi <- rep(2.93/sqrt(dim_phi), N); acc_phi <- rep(0, N)
  s_zeta <- 0.05; acc_zeta <- 0; window <- 100
  keep_betabar <- matrix(NA_real_, nkeep, dim_phi)
  keep_Sigma <- array(NA_real_, c(nkeep, dim_phi, dim_phi))
  keep_cut <- if (!het_cut && !choice_only) matrix(NA_real_, nkeep, W-1) else NULL
  keep_ll <- numeric(nkeep); keep_ll_resp <- matrix(NA_real_, nkeep, N)
  keep_Phi <- if (keep_phi) array(NA_real_, c(nkeep, N, dim_phi)) else NULL
  t0 <- Sys.time(); ki <- 0
  for (it in seq_len(R_iter)) {
    U <- chol(Sigma)
    Prop <- Phi + (matrix(rnorm(N*dim_phi), N, dim_phi) %*% U) * s_phi
    prop_cutmat <- if (het_cut) phi_to_cutmat(Prop, P, W) else cur_cutmat
    prop_ll <- resp_loglik_all(dat, Prop[, 1:P, drop=FALSE], prop_cutmat, choice_only)
    Si <- chol2inv(chol(Sigma))
    dcur <- sweep(Phi, 2, phibar); dprop <- sweep(Prop, 2, phibar)
    la <- (prop_ll - 0.5*rowSums((dprop %*% Si)*dprop)) -
          (cur_ll  - 0.5*rowSums((dcur  %*% Si)*dcur))
    acc <- log(runif(N)) < la
    Phi[acc, ] <- Prop[acc, ]; cur_ll[acc] <- prop_ll[acc]
    if (het_cut && any(acc)) cur_cutmat[acc, ] <- prop_cutmat[acc, , drop=FALSE]
    acc_phi <- acc_phi + acc
    if (!het_cut && !choice_only) {
      zp <- zeta + rnorm(W-1, sd = s_zeta)
      cp <- matrix(par_to_cut(zp, 0, W), N, W-1, byrow = TRUE)
      lp_ll <- resp_loglik_all(dat, Phi[, 1:P, drop=FALSE], cp, choice_only)
      if (log(runif(1)) < (sum(lp_ll) - 0.5*priors$zeta_prec*sum(zp^2)) -
                          (sum(cur_ll) - 0.5*priors$zeta_prec*sum(zeta^2))) {
        zeta <- zp; cur_cutmat <- cp; cur_ll <- lp_ll; acc_zeta <- acc_zeta + 1
      }
    }
    phibar <- draw_phibar(Phi, Sigma, priors); Sigma <- draw_sigma(Phi, phibar, priors)
    if (it <= burn && it %% window == 0) {
      s_phi <- pmin(pmax(s_phi*exp(1.5*(acc_phi/window - 0.23)), 0.01), 10)
      acc_phi <- rep(0, N)
      if (!het_cut && !choice_only) {
        s_zeta <- min(max(s_zeta*exp(1.5*(acc_zeta/window - 0.30)), 1e-3), 2); acc_zeta <- 0
      }
    }
    if (it == burn) { acc_phi <- rep(0, N); acc_zeta <- 0 }
    if (it > burn && (it - burn) %% thin == 0) {
      ki <- ki + 1
      keep_betabar[ki, ] <- phibar; keep_Sigma[ki, , ] <- Sigma
      if (!is.null(keep_cut)) keep_cut[ki, ] <- par_to_cut(zeta, 0, W)
      keep_ll[ki] <- sum(cur_ll); keep_ll_resp[ki, ] <- cur_ll
      if (keep_phi) keep_Phi[ki, , ] <- Phi
    }
  }
  post <- R_iter - burn
  structure(list(draws = list(phibar = keep_betabar, Sigma = keep_Sigma,
                              cut = keep_cut, Phi = keep_Phi),
                 ll = keep_ll, ll_resp = keep_ll_resp,
                 accept = list(phi = mean(acc_phi/post),
                               zeta = if (het_cut || choice_only) NA else acc_zeta/post),
                 settings = list(mcmc = mcmc, het_cut = het_cut,
                                 choice_only = choice_only, P = P, W = W, N = N,
                                 model = dat$model, priors = priors),
                 runtime_min = as.numeric(difftime(Sys.time(), t0, units="mins"))),
            class = "dr_hb_fit")
}
hb_beta_i <- function(fit)
  apply(fit$draws$Phi[, , 1:fit$settings$P, drop = FALSE], c(2,3), mean)

logmeanexp <- function(x) { m <- max(x); m + log(mean(exp(x - m))) }
waic <- function(ll) {
  lppd <- sum(apply(ll, 2, logmeanexp)); p <- sum(apply(ll, 2, var))
  list(waic = -2*(lppd - p), lppd = lppd, p_waic = p, elpd = lppd - p)
}
```

```{r}
#| label: config
N <- 300; T_EST <- 12; T_HOLD <- 2; J <- 4; W <- 5
beta_bar   <- c(a2 = 0.8, a3 = -0.5, b2 = 0.4, b3 = 1.0, price = -0.9)
Sigma_true <- diag(c(0.6, 0.6, 0.6, 0.6, 0.3))
cut_true   <- c(-1.0, 0.2, 1.2, 2.2)
P <- length(beta_bar)
KBUY <- 4                      # "purchase" means top-2-box throughout
REPS <- 12
ITER <- 20000
MCMC <- list(R = ITER, burn = floor(ITER * 0.4), thin = 10)

gen_panel <- function(seed, model = "B", cutmat = NULL, ...) {
  des <- make_panel_design(N, T_EST + T_HOLD, J, seed = seed)
  B <- draw_betas(N, beta_bar, Sigma_true, seed = seed + 1)
  cutarg <- if (is.null(cutmat)) cut_true else cutmat
  split_holdout(simulate_dual_hb(des, B, cutarg, model = model, seed = seed + 2, ...), T_HOLD)
}
task_mubar <- function(design, Bmat) {
  Bx <- Bmat[design$resp_row, , drop = FALSE]
  V <- matrix(rowSums(design$X * Bx), ncol = design$J, byrow = TRUE)
  m <- row_max(V); m + log(rowSums(exp(V - m)))
}
pred_exceed_panel <- function(fit, sp, k = KBUY) {
  d <- sp$holdout$design
  mu <- task_mubar(d, hb_beta_i(fit))
  if (fit$settings$het_cut) {
    cutm <- phi_to_cutmat(apply(fit$draws$Phi, c(2,3), mean),
                          fit$settings$P, fit$settings$W)
    cc <- cutm[d$resp, k - 1]
  } else {
    cv <- colMeans(fit$draws$cut); cc <- rep(cv[min(k - 1, length(cv))], length(mu))
  }
  G <- if (fit$settings$model == "B") plogis else function(z) exp(-exp(-z))
  1 - G(cc - mu)
}
hit_rate <- function(fit, sp) {
  d <- sp$holdout$design
  Bx <- hb_beta_i(fit)[d$resp_row, , drop = FALSE]
  V <- matrix(rowSums(d$X * Bx), ncol = d$J, byrow = TRUE)
  mean(max.col(V, ties.method = "first") == sp$holdout$jstar)
}
res_all <- list()
```

## 2. Wrong link

Fit Model B to Model A data and vice versa. The question is whether the two
links are distinguishable, and whether getting it wrong matters for demand.

```{r}
#| label: wronglink
m <- matrix(NA_real_, REPS, 6, dimnames = list(NULL, c(
  "bias_ex_correct_A","bias_ex_wrong_A","bias_ex_correct_B","bias_ex_wrong_B",
  "waic_picks_A","waic_picks_B")))
for (r in seq_len(REPS)) {
  row <- numeric(6)
  for (truth in c("A","B")) {
    sp <- gen_panel(seed = 70000 + 20*r + (truth == "A"), model = truth)
    d <- sp$holdout$design
    mu_true <- task_mubar(d, sp$holdout$Bmat_true)
    Gt <- if (truth == "B") plogis else function(z) exp(-exp(-z))
    tex <- 1 - Gt(cut_true[KBUY - 1] - mu_true)
    fits <- list()
    for (fl in c("A","B")) {
      datf <- sp$est; datf$model <- fl
      fits[[fl]] <- fit_dual_hb(datf, mcmc = MCMC,
                                seed = 71000 + 2*r + (fl == "A"), verbose = FALSE)
    }
    wrong <- setdiff(c("A","B"), truth)
    bc <- mean(pred_exceed_panel(fits[[truth]], sp) - tex)
    bw <- mean(pred_exceed_panel(fits[[wrong]], sp) - tex)
    picked <- if (waic(fits$A$ll_resp)$elpd > waic(fits$B$ll_resp)$elpd) "A" else "B"
    if (truth == "A") { row[1:2] <- c(bc, bw); row[5] <- picked == "A" }
    else              { row[3:4] <- c(bc, bw); row[6] <- picked == "B" }
  }
  m[r, ] <- row
}
res_all$wronglink <- list(metrics = m, mean = colMeans(m),
                          se = apply(m, 2, sd)/sqrt(REPS))
knitr::kable(rbind(mean = res_all$wronglink$mean, mc_se = res_all$wronglink$se),
             digits = 4)
```

```{r}
#| label: wronglink-read
w <- res_all$wronglink$mean
cat(sprintf("demand bias, correct link: up to %.4f on the probability scale\n",
            max(abs(c(w["bias_ex_correct_A"], w["bias_ex_correct_B"])))))
cat(sprintf("demand bias, wrong link  : up to %.4f\n",
            max(abs(c(w["bias_ex_wrong_A"], w["bias_ex_wrong_B"])))))
cat(sprintf("WAIC selects the true link in %.0f%% of replications\n",
            100 * mean(c(w["waic_picks_A"], w["waic_picks_B"]))))
```

Link misspecification is a second-order concern *for demand*. The two links
agree closely over the range of inclusive values this design produces, and
diverge only in the tails, which is where extrapolation lives. Notebook 06 shows
that the ability to tell the links apart is governed by how much the design
varies task attractiveness, not by sample size.

## 3. Ignored scale-usage heterogeneity

Now respondents genuinely differ in how they use the scale: each gets her own
cut points. Fit the common-cut model (wrong) and the heterogeneous-cut model
(right), and compare.

```{r}
#| label: hetcut
m <- matrix(NA_real_, REPS, 6, dimnames = list(NULL, c(
  "rmse_bi_common","rmse_bi_het","rmse_ex_common","rmse_ex_het",
  "hit_common","hit_het")))
theta_mu <- c(-1, log(1.2), log(1.0), log(1.0))
theta_sd <- c(0.35, 0.25, 0.25, 0.25)
for (r in seq_len(REPS)) {
  set.seed(80000 + 10*r)
  Theta <- sweep(matrix(rnorm(N*4), N, 4) %*% diag(theta_sd), 2, theta_mu, "+")
  cutmat <- t(apply(Theta, 1, function(th) th[1] + c(0, cumsum(exp(th[2:4])))))
  sp <- gen_panel(seed = 80000 + 10*r + 1, cutmat = cutmat)
  f_com <- fit_dual_hb(sp$est, mcmc = MCMC, seed = 81000 + r, verbose = FALSE)
  f_het <- fit_dual_hb(sp$est, mcmc = MCMC, het_cut = TRUE,
                       seed = 82000 + r, verbose = FALSE)
  d <- sp$holdout$design
  tex <- 1 - plogis(cutmat[d$resp, KBUY - 1] - task_mubar(d, sp$holdout$Bmat_true))
  m[r, ] <- c(sqrt(mean((hb_beta_i(f_com) - sp$est$Bmat_true)^2)),
              sqrt(mean((hb_beta_i(f_het) - sp$est$Bmat_true)^2)),
              sqrt(mean((pred_exceed_panel(f_com, sp) - tex)^2)),
              sqrt(mean((pred_exceed_panel(f_het, sp) - tex)^2)),
              hit_rate(f_com, sp), hit_rate(f_het, sp))
}
res_all$hetcut <- list(metrics = m, mean = colMeans(m),
                       se = apply(m, 2, sd)/sqrt(REPS),
                       theta_mu = theta_mu, theta_sd = theta_sd)
knitr::kable(rbind(mean = res_all$hetcut$mean, mc_se = res_all$hetcut$se), digits = 4)
```

The common-cut model loses almost nothing, even though the data-generating
process really does give every respondent her own thresholds with a standard
deviation of 0.35 on the first one. At twelve tasks per person there is not
enough information to estimate four extra parameters per respondent, so the
extra flexibility buys little and costs variance. The heterogeneous-cut
extension is robustness insurance and a diagnostic, not a default.

## 4. Non-Gumbel shocks

The entire derivation rests on Gumbel errors. Replace them with normals matched
on variance and refit.

```{r}
#| label: nongumbel
m <- matrix(NA_real_, REPS, 4,
            dimnames = list(NULL, c("rmse_bi","rmse_ex","hit","bias_c3")))
for (r in seq_len(REPS)) {
  sp <- gen_panel(seed = 90000 + 10*r, shock = "normal")
  f <- fit_dual_hb(sp$est, mcmc = MCMC, seed = 91000 + r, verbose = FALSE)
  # With normal shocks there is no closed form for demand, so the truth is
  # obtained by simulating the behavioral process at the holdout tasks.
  set.seed(92000 + r)
  d2 <- sp$holdout$design
  Bx <- sp$holdout$Bmat_true[d2$resp_row, , drop = FALSE]
  V <- matrix(rowSums(d2$X * Bx), ncol = J, byrow = TRUE)
  cnt <- numeric(nrow(V))
  for (mc in seq_len(1000)) {
    u <- V + matrix(rnorm(length(V), sd = pi/sqrt(6)), nrow(V), J)
    ustar <- u[cbind(seq_len(nrow(V)), max.col(u, ties.method = "first"))]
    cnt <- cnt + ((ustar - rgumbel(nrow(V))) >= cut_true[KBUY - 1])
  }
  ex_true <- cnt / 1000
  m[r, ] <- c(sqrt(mean((hb_beta_i(f) - sp$est$Bmat_true)^2)),
              sqrt(mean((pred_exceed_panel(f, sp) - ex_true)^2)),
              hit_rate(f, sp),
              mean(f$draws$cut[, KBUY - 1]) - cut_true[KBUY - 1])
}
res_all$nongumbel <- list(metrics = m, mean = colMeans(m),
                          se = apply(m, 2, sd)/sqrt(REPS))
knitr::kable(rbind(mean = res_all$nongumbel$mean, mc_se = res_all$nongumbel$se),
             digits = 4)
```

Preference recovery and demand prediction survive. The misspecification is
absorbed almost entirely by `bias_c3`, the displacement of the top-2-box cut
point. This is the pattern promised at the top: the cut points act as a sponge.

## 5. Correlated alternatives

Appendix D proves the framework survives any GEV error structure once the
inclusive value is redefined as $\ln G(e^V)$. Fitting the independent-Gumbel
model to nested-logit data therefore misspecifies exactly one thing: it uses
$\ln \sum_j e^{V_j}$ where the truth is $\ln G(e^V)$.

Four alternatives in two nests at dissimilarity 0.5, which is a within-nest
error correlation of 0.75.

```{r}
#| label: nested
NEST <- list(groups = c(1,1,2,2), lambda = c(0.5, 0.5))
m <- matrix(NA_real_, REPS, 4,
            dimnames = list(NULL, c("rmse_bi","rmse_ex","hit","bias_c3")))
for (r in seq_len(REPS)) {
  sp <- gen_panel(seed = 96000 + 10*r, nest = NEST)
  f <- fit_dual_hb(sp$est, mcmc = MCMC, seed = 97000 + r, verbose = FALSE)
  set.seed(98000 + r)
  d2 <- sp$holdout$design
  Bx <- sp$holdout$Bmat_true[d2$resp_row, , drop = FALSE]
  V <- matrix(rowSums(d2$X * Bx), ncol = J, byrow = TRUE)
  cnt <- numeric(nrow(V))
  for (mc in seq_len(1000)) {
    u <- V + rnested(nrow(V), J, NEST)
    ustar <- u[cbind(seq_len(nrow(V)), max.col(u, ties.method = "first"))]
    cnt <- cnt + ((ustar - rgumbel(nrow(V))) >= cut_true[KBUY - 1])
  }
  ex_true <- cnt / 1000
  m[r, ] <- c(sqrt(mean((hb_beta_i(f) - sp$est$Bmat_true)^2)),
              sqrt(mean((pred_exceed_panel(f, sp) - ex_true)^2)),
              hit_rate(f, sp),
              mean(f$draws$cut[, KBUY - 1]) - cut_true[KBUY - 1])
}
res_all$nested <- list(metrics = m, mean = colMeans(m),
                       se = apply(m, 2, sd)/sqrt(REPS), nest = NEST)
knitr::kable(rbind(mean = res_all$nested$mean, mc_se = res_all$nested$se), digits = 4)
```

Again the cut point absorbs it, and in the direction theory predicts. Nesting
makes $G(e^V) < \sum_j e^{V_j}$, so the fitted model overstates the inclusive
value, and the top-2-box cut compensates by moving upward. The sign of
`bias_c3` is therefore a prediction, not a fitted artifact.

## 6. Task-order drift

Respondents are not static across a survey. The outside good may drift as they
work through it, which is the phenomenon documented in the no-choice literature.
Two variants: a deterministic linear drift, and Hung-style probabilistic
updating toward the previous task's inclusive value.

```{r}
#| label: drift
variants <- list(linear = list(type = "linear", delta = 0.06),
                 update = list(type = "update", rho = -2.5))
drift_out <- list()
for (vn in names(variants)) {
  m <- matrix(NA_real_, REPS, 4,
              dimnames = list(NULL, c("rmse_bi","bias_c3","order_slope","order_t")))
  for (r in seq_len(REPS)) {
    sp <- gen_panel(seed = 95000 + 10*r, drift = variants[[vn]])
    f <- fit_dual_hb(sp$est, mcmc = MCMC, seed = 96000 + r, verbose = FALSE)
    d <- sp$est$design
    mu <- task_mubar(d, hb_beta_i(f))
    chat <- colMeans(f$draws$cut)[KBUY - 1]
    resid <- as.numeric(sp$est$y >= KBUY) - (1 - plogis(chat - mu))
    tt <- rep(seq_len(T_EST), times = N)
    reg <- summary(lm(resid ~ tt))
    m[r, ] <- c(sqrt(mean((hb_beta_i(f) - sp$est$Bmat_true)^2)),
                chat - cut_true[KBUY - 1],
                reg$coefficients[2,1], reg$coefficients[2,3])
  }
  drift_out[[vn]] <- list(metrics = m, mean = colMeans(m),
                          se = apply(m, 2, sd)/sqrt(REPS))
}
res_all$drift <- drift_out
knitr::kable(do.call(rbind, lapply(names(drift_out), function(v)
  data.frame(variant = v, t(round(drift_out[[v]]$mean, 4))))), row.names = FALSE)
```

The static model's cut points absorb the average drift while part-worths stay
accurate. More useful for practice: a one-line diagnostic detects it. Regress the
residual exceedance on task position, and under linear drift the $t$ statistic
is large and consistently signed, so an analyst can find this in real data
without fitting a dynamic model.

## 7. What the five have in common

```{r}
#| label: summary
knitr::kable(data.frame(
  misspecification = c("wrong link","heterogeneous cuts","normal shocks",
                       "correlated alternatives","task-order drift"),
  `preference recovery` = c("unaffected","unaffected","unaffected",
                            "slightly worse","unaffected"),
  `demand prediction` = c("second-order","unaffected","unaffected",
                          "unaffected","unaffected"),
  `where the damage lands` = c("tails only","nowhere material","cut points",
                               "cut points","cut points"),
  check.names = FALSE), row.names = FALSE)
```

Four of the five leave the quantities practitioners actually use essentially
intact, and deposit their damage in the cut points. The reason is structural
rather than lucky: the forced choice identifies preferences and is untouched by
any of these, while the cut points are the model's only free location and so
absorb whatever the second stage cannot express.

The implication cuts both ways. Demand estimates are robust. The *structural*
reading of the cut points, as probability thresholds that give scale labels
quantitative meaning, is the fragile part, and it is exactly what Model A's
probability-scale interpretation asks a reader to believe.

## 8. Results written

```{r}
#| label: save
PROJ <- if (file.exists("_quarto.yml")) "." else ".."
OUT  <- file.path(PROJ, "R", "output")
dir.create(OUT, showWarnings = FALSE, recursive = TRUE)

settings <- list(N = N, T_EST = T_EST, T_HOLD = T_HOLD, J = J, W = W,
                 beta_bar = beta_bar, Sigma = Sigma_true, cut = cut_true,
                 REPS = REPS, MCMC = MCMC, KBUY = KBUY)
for (cell in c("wronglink","hetcut","nongumbel","nested","drift")) {
  saveRDS(list(cell = cell, res = res_all[[cell]], settings = settings),
          file.path(OUT, paste0("hb_simstudy_", cell, ".rds")))
}
cat("wrote five hb_simstudy_*.rds files\n")
```

## 9. Related material

| where | what |
|---|---|
| notebook 04 | the tests that catch two of these before they do damage |
| notebook 06 | why link discrimination depends on design, not sample size |
| Appendix D | the GEV result behind the correlated-alternatives cell |
