---
title: "Specification Tests"
subtitle: "Two sharp tests of the behavioral story, and how well they work"
author:
  - name: Dan Yavorsky
  - name: Geoffery Zheng
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

{{< include ../latex_commands.qmd >}}

## What this notebook does

The model makes two claims about the second response that a reader is entitled
to doubt, and both can be turned into one-parameter extensions of the likelihood
and tested.

**Claim one: the slope on the inclusive value is exactly one.** Because both
responses come from a single, internally consistent utility draw, the ordinal
stage reads $\overline{\mu}_{it} = \ln S_{it}$ with no free loading. Relax it to

$$
\Pr(y \le w) \;=\; G\!\left( c_w - \lambda \, \overline{\mu}_{it} \right)
$$

and test $H_0 : \lambda = 1$. This is the parameter one would most naturally
estimate freely, and the theory says not to.

**Claim two: the second response depends on the choice set only through the
inclusive value, not through which alternative won.** By Lemma 1(ii) the
distribution of the best utility does not depend on which good attained it. The
competing model in the literature, DR-2Max, conditions on the chosen good's
utility instead. Nest both:

$$
\text{second-stage predictor} \;=\; \overline{\mu}_{it} + \delta \left( V_{itj^*} - \overline{\mu}_{it} \right)
$$

with $\delta = 0$ our model and $\delta = 1$ DR-2Max. Two tests fall out:
$H_0 : \delta = 0$, and, running the comparison the other way,
$H_0 : \delta = 1$.

This notebook establishes that both tests hold their nominal size and have power
against the alternatives that matter. It is the source of `spec_tests.rds`.

::: {.callout-note}
## Why the $\lambda$ test earns its place

A coauthor observed in 2024 that fitting an ordered regression with a freely
estimated slope on a latent driver can return an attenuated coefficient. That
turned out to be a reverse-regression artifact which does not arise in this
model's data-generating process, but it identified the right parameter to worry
about. Rather than assume $\lambda = 1$, we estimate it and report the test. See
`archive/identification_note.qmd`.
:::

```{r}
#| label: setup
set.seed(1)
options(digits = 5)
```

## 1. Machinery

```{r}
#| label: machinery
#| code-fold: true
#| code-summary: "Design, simulator (including both alternatives), and likelihood"
rgumbel <- function(n) -log(-log(runif(n)))

make_design <- function(n_tasks, J, seed, intercept = FALSE) {
  set.seed(seed)
  n_rows <- n_tasks * J
  a <- sample(1:3, n_rows, replace = TRUE); b <- sample(1:3, n_rows, replace = TRUE)
  price <- runif(n_rows, 0.5, 2.5)
  X <- cbind(a2 = as.numeric(a == 2), a3 = as.numeric(a == 3),
             b2 = as.numeric(b == 2), b3 = as.numeric(b == 3), price = price)
  if (intercept) X <- cbind(X, const = 1)
  list(X = X, n_tasks = n_tasks, J = J, P = ncol(X))
}
par_to_cut <- function(par, P, W) {
  if (W == 2) return(par[P + 1])
  par[P + 1] + c(0, cumsum(exp(par[(P + 2):(P + W - 1)])))
}
cut_to_par <- function(cut) if (length(cut) == 1) cut else c(cut[1], log(diff(cut)))
row_max <- function(M) do.call(pmax, as.data.frame(M))
ord_prob_z <- function(lo, hi, model) {
  if (model == "B") plogis(hi) * plogis(-lo) * (-expm1(lo - hi))
  else { ea <- exp(-lo); eb <- exp(-hi)
         p <- exp(-eb) * (-expm1(-(ea - eb))); p[!is.finite(eb)] <- 0; p }
}
ord_prob <- function(mubar, cut, y, model) {
  caug <- c(-Inf, cut, Inf); ord_prob_z(caug[y] - mubar, caug[y + 1L] - mubar, model)
}
negloglik <- function(par, dat) {
  design <- dat$design; P <- design$P; W <- dat$W
  n <- design$n_tasks; J <- design$J
  beta <- par[1:P]; cut <- par_to_cut(par, P, W)
  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  m <- row_max(V); logS <- m + log(rowSums(exp(V - m)))
  -(sum(V[cbind(seq_len(n), dat$jstar)] - logS) +
    sum(log(pmax(ord_prob(logS, cut, dat$y, dat$model), 1e-312))))
}
fit_dual_mle <- function(dat, start = NULL) {
  design <- dat$design; P <- design$P; W <- dat$W
  if (is.null(start)) {
    freq <- tabulate(dat$y, nbins = W)
    cumq <- pmin(pmax(cumsum(freq)[1:(W - 1)] / sum(freq), 1e-4), 1 - 1e-4)
    ginv <- if (dat$model == "B") qlogis else function(q) -log(-log(q))
    cut0 <- log(design$J) + ginv(cumq)
    if (W > 2) for (k in 2:(W - 1)) cut0[k] <- max(cut0[k], cut0[k - 1] + 1e-3)
    start <- c(rep(0, P), cut_to_par(cut0))
  }
  fn <- function(p) negloglik(p, dat)
  opt <- optim(start, fn, method = "BFGS", control = list(maxit = 1000, reltol = 1e-12))
  list(beta = opt$par[1:P], cut = par_to_cut(opt$par, P, W),
       nll = opt$value, convergence = opt$convergence, par = opt$par)
}

# Simulator carrying both alternatives:
#   model2 = "2max"  -> second stage reads the CHOSEN good's utility (DR-2Max)
#   lambda != 1      -> second-stage latent centred on lambda * log S
simulate_dual <- function(design, beta, cut, model = c("A", "B"), seed,
                          model2 = c("inclusive", "2max"), lambda = 1) {
  model <- match.arg(model); model2 <- match.arg(model2)
  set.seed(seed)
  n <- design$n_tasks; J <- design$J
  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  u <- V + matrix(rgumbel(n * J), n, J)
  jstar <- max.col(u, ties.method = "first")
  ustar <- u[cbind(seq_len(n), jstar)]
  m <- row_max(V); logS <- m + log(rowSums(exp(V - m)))
  latent <- if (model2 == "2max") {
    V[cbind(seq_len(n), jstar)] + rlogis(n)
  } else if (lambda != 1) {
    rlogis(n, location = lambda * logS)
  } else if (model == "A") ustar else ustar - rgumbel(n)
  list(design = design, jstar = jstar, y = findInterval(latent, cut) + 1L,
       W = length(cut) + 1L, model = model, beta_true = beta, cut_true = cut,
       model2 = model2, lambda = lambda)
}
```

## 2. The two extensions

Both tests come from one extended likelihood. The only thing that changes is
what enters the ordinal stage as the predictor.

```{r}
#| label: ext-likelihood
# ext = "lambda": predictor is theta * logS
# ext = "delta" : predictor is logS + theta * (V_{j*} - logS)
# If `fix` is supplied the extension parameter is held there instead of estimated.
negloglik_ext <- function(par, dat, ext = c("lambda", "delta"), fix = NULL) {
  ext <- match.arg(ext)
  design <- dat$design; P <- design$P; W <- dat$W
  n <- design$n_tasks; J <- design$J
  if (is.null(fix)) { theta <- par[P + W]; par_core <- par[1:(P + W - 1)] }
  else              { theta <- fix;        par_core <- par }
  beta <- par_core[1:P]; cut <- par_to_cut(par_core, P, W)

  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  m <- row_max(V); logS <- m + log(rowSums(exp(V - m)))
  Vstar <- V[cbind(seq_len(n), dat$jstar)]

  pred <- if (ext == "lambda") theta * logS else logS + theta * (Vstar - logS)
  -(sum(Vstar - logS) + sum(log(pmax(ord_prob(pred, cut, dat$y, dat$model), 1e-312))))
}

fit_dual_mle_ext <- function(dat, ext, start_core, theta0) {
  fn <- function(p) negloglik_ext(p, dat, ext = ext)
  opt <- optim(c(start_core, theta0), fn, method = "BFGS",
               control = list(maxit = 1000, reltol = 1e-12))
  Vc <- tryCatch(solve(optimHess(opt$par, fn)),
                 error = function(e) matrix(NA, length(opt$par), length(opt$par)))
  k <- length(opt$par)
  list(par = opt$par, theta = opt$par[k], se_theta = sqrt(Vc[k, k]),
       nll = opt$value, convergence = opt$convergence)
}
```

Note the economy of the $\delta$ construction. At $\delta = 0$ the predictor
collapses to $\overline{\mu}$, so the restricted fit *is* the ordinary MLE and
costs nothing extra. At $\delta = 1$ it collapses to $V_{j^*}$, which is DR-2Max.
One extension nests both models in the literature and ours.

```{r}
#| label: battery
spec_test_battery <- function(dat, mle = NULL) {
  if (is.null(mle)) mle <- fit_dual_mle(dat)
  start_core <- mle$par

  fl <- fit_dual_mle_ext(dat, "lambda", start_core, theta0 = 1)   # H0: lambda = 1
  lr_l <- 2 * (mle$nll - fl$nll)

  fd <- fit_dual_mle_ext(dat, "delta", start_core, theta0 = 0)    # H0: delta = 0
  lr_d0 <- 2 * (mle$nll - fd$nll)

  fn_d1 <- function(p) negloglik_ext(p, dat, ext = "delta", fix = 1)
  opt_d1 <- optim(start_core, fn_d1, method = "BFGS",
                  control = list(maxit = 1000, reltol = 1e-12))
  lr_d1 <- 2 * (opt_d1$value - fd$nll)                            # H0: delta = 1

  z <- c((fl$theta - 1) / fl$se_theta, fd$theta / fd$se_theta,
         (fd$theta - 1) / fd$se_theta)
  data.frame(test = c("lambda = 1", "delta = 0", "delta = 1 (DR-2Max)"),
             estimate = c(fl$theta, fd$theta, fd$theta),
             se = c(fl$se_theta, fd$se_theta, fd$se_theta),
             wald_z = z, lr = c(lr_l, lr_d0, lr_d1),
             p_wald = 2 * pnorm(-abs(z)),
             p_lr = pchisq(c(lr_l, lr_d0, lr_d1), df = 1, lower.tail = FALSE))
}
```

## 3. One dataset, worked

Before the repeated-sampling study, here is the battery on a single dataset
generated from the model.

```{r}
#| label: config
J <- 4; W <- 5
beta     <- c(a2 = 0.8, a3 = -0.5, b2 = 0.4, b3 = 1.0, price = -0.9)
cut_true <- c(-1.0, 0.2, 1.2, 2.2)

gen <- function(n_tasks, seed, model = "B", model2 = "inclusive", lambda = 1) {
  des <- make_design(n_tasks, J, seed = seed)
  simulate_dual(des, beta, cut_true, model = model, seed = seed + 1,
                model2 = model2, lambda = lambda)
}
```

```{r}
#| label: worked-null
knitr::kable(spec_test_battery(gen(3600, seed = 20007)),
             row.names = FALSE, digits = 4)
```

Both null tests fail to reject, as they should. The third row is the informative
one: $\delta = 1$ is rejected decisively, so on data from our model the test
correctly discards DR-2Max.

Now the same battery on data actually generated from DR-2Max:

```{r}
#| label: worked-2max
knitr::kable(spec_test_battery(gen(3600, seed = 70007, model2 = "2max")),
             row.names = FALSE, digits = 4)
```

The verdicts reverse. $\hat\delta$ lands near one, $\delta = 0$ is rejected, and
$\delta = 1$ is not. A test that only ever rejected in one direction would be
useless; this one discriminates both ways.

## 4. Size

Data generated from the correctly specified model. Nominal 5% tests should
reject about 5% of the time.

```{r}
#| label: study
R_SIZE <- 300
R_POW  <- 150

run_cell <- function(label, n_tasks, reps, seed0, model = "B",
                     model2 = "inclusive", lambda = 1) {
  rej <- matrix(0, reps, 3,
                dimnames = list(NULL, c("lambda=1", "delta=0", "delta=1")))
  est <- matrix(NA_real_, reps, 2,
                dimnames = list(NULL, c("lambda_hat", "delta_hat")))
  for (r in seq_len(reps)) {
    dat <- gen(n_tasks, seed = seed0 + 7 * r, model = model,
               model2 = model2, lambda = lambda)
    tb <- spec_test_battery(dat)
    rej[r, ] <- tb$p_lr < 0.05
    est[r, ] <- c(tb$estimate[1], tb$estimate[2])
  }
  list(label = label, n = n_tasks, reps = reps, reject = colMeans(rej),
       est_mean = colMeans(est), est_sd = apply(est, 2, sd))
}

cells <- list()
cells$size_B_1200 <- run_cell("size B", 1200, R_SIZE, 10000)
cells$size_B_3600 <- run_cell("size B", 3600, R_SIZE, 20000)
cells$size_A_3600 <- run_cell("size A", 3600, R_SIZE, 30000, model = "A")
```

```{r}
#| label: size-table
as_row <- function(cl) data.frame(
  cell = cl$label, n = cl$n, reps = cl$reps,
  `reject lambda=1` = round(cl$reject[1], 3),
  `reject delta=0`  = round(cl$reject[2], 3),
  `reject delta=1`  = round(cl$reject[3], 3),
  `mean lambda_hat` = round(cl$est_mean[1], 3),
  `mean delta_hat`  = round(cl$est_mean[2], 3),
  check.names = FALSE)
knitr::kable(do.call(rbind, lapply(cells, as_row)), row.names = FALSE)
```

Three things to read here. The first two rejection columns sit near 0.05, so both
tests hold their size. The estimates are centred on their theoretical values,
$\lambda$ on one and $\delta$ on zero, so neither extension is biased. And the
third column is not a size result at all: it is the power of the $\delta = 1$
test against our model, and it is **1.000 in every cell**. The
2Max-versus-inclusive-value question does not need large data to settle.

## 5. Power

Two families of alternative. Slope alternatives put $\lambda$ away from one;
the DR-2Max alternative generates the second response from the chosen good.

```{r}
#| label: power
cells$pow_lam07_3600 <- run_cell("power lambda=0.7", 3600, R_POW, 40000, lambda = 0.7)
cells$pow_lam13_3600 <- run_cell("power lambda=1.3", 3600, R_POW, 50000, lambda = 1.3)
cells$pow_lam07_1200 <- run_cell("power lambda=0.7", 1200, R_POW, 60000, lambda = 0.7)
cells$pow_2max_3600  <- run_cell("power 2max (delta=1)", 3600, R_POW, 70000, model2 = "2max")
cells$pow_2max_1200  <- run_cell("power 2max (delta=1)", 1200, R_POW, 80000, model2 = "2max")
```

```{r}
#| label: power-table
knitr::kable(do.call(rbind, lapply(cells[4:8], as_row)), row.names = FALSE)
```

Read each row against the alternative that generated it.

Under the **slope alternatives**, $\hat\lambda$ recovers the true 0.7 and 1.3,
the $\lambda$ test rejects with power 0.99 at 3,600 tasks, and the $\delta$ tests
stay quiet: a slope problem is not mistaken for a conditioning problem. At 1,200
tasks the power falls to roughly two-thirds, which is a useful design fact. A
study intending to test the unit-slope claim needs more than a thousand tasks.

Under the **DR-2Max alternative**, $\hat\delta$ lands on one, $\delta = 0$ is
rejected in every replication, and $\delta = 1$ is not rejected. The $\lambda$
test again stays quiet. Each test targets its own alternative and ignores the
other, which is what makes a rejection interpretable.

## 6. The full picture

```{r}
#| label: full-table
knitr::kable(do.call(rbind, lapply(cells, as_row)), row.names = FALSE)
```

## 7. Results written

```{r}
#| label: save
PROJ <- if (file.exists("_quarto.yml")) "." else ".."
OUT  <- file.path(PROJ, "R", "output")
dir.create(OUT, showWarnings = FALSE, recursive = TRUE)

saveRDS(list(cells = cells,
             settings = list(beta = beta, cut = cut_true, J = J, W = W,
                             R_SIZE = R_SIZE, R_POW = R_POW)),
        file.path(OUT, "spec_tests.rds"))
cat("wrote spec_tests.rds\n")
```

## 8. Related material

| where | what |
|---|---|
| notebook 02 | the aggregate estimator these tests extend |
| notebook 06 | what happens when a misspecification is *not* tested for |
| notebook 07 | why the link-discrimination question needs attractiveness variation |
| `archive/identification_note.qmd` | where the $\lambda$ test came from |
