---
title: "Designing the Scale"
subtitle: "How many points, where to put them, and what the design must vary"
author:
  - name: Dan Yavorsky
  - name: Geoffery Zheng
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

{{< include ../latex_commands.qmd >}}

## What this notebook does

The information results in the paper are abstract. This notebook turns them into
three concrete decisions a researcher has to make before fielding a study.

1. **How many scale points?** A binary follow-up is the least informative member
   of the family and a continuous report is the limit. Where between them does a
   five- or seven-point scale sit?
2. **Where do the cut points fall?** Scale size is only half the problem. The
   labels a respondent sees determine where the thresholds land, and placement
   turns out to matter about as much as count, at least for one of the two links.
3. **What must the design vary?** Separately from the scale, telling Model A from
   Model B requires the tasks to vary in attractiveness. This is a property of
   the design, not of the sample size, and the distinction has real consequences:
   a study can collect unlimited data and still learn nothing about the link.

The third result is the sharpest, and it is a negative one. It is also the
cleanest empirical demonstration of the identification argument in the model
section, since it shows exactly what is lost when the identifying variation is
removed.

This notebook is the source of `link_discrimination.rds`,
`link_discrimination_squashed.rds`, `design_guidance.rds`, and the figure the
article displays.

```{r}
#| label: setup
set.seed(1)
options(digits = 5)
```

## 1. Machinery

```{r}
#| label: machinery
#| code-fold: true
#| code-summary: "Design, simulator, likelihood, and the information weight omega"
rgumbel <- function(n) -log(-log(runif(n)))
make_design <- function(n_tasks, J, seed, intercept = FALSE) {
  set.seed(seed)
  n_rows <- n_tasks * J
  a <- sample(1:3, n_rows, replace = TRUE); b <- sample(1:3, n_rows, replace = TRUE)
  price <- runif(n_rows, 0.5, 2.5)
  X <- cbind(a2 = as.numeric(a == 2), a3 = as.numeric(a == 3),
             b2 = as.numeric(b == 2), b3 = as.numeric(b == 3), price = price)
  if (intercept) X <- cbind(X, const = 1)
  list(X = X, n_tasks = n_tasks, J = J, P = ncol(X))
}
par_to_cut <- function(par, P, W) {
  if (W == 2) return(par[P + 1])
  par[P + 1] + c(0, cumsum(exp(par[(P + 2):(P + W - 1)])))
}
cut_to_par <- function(cut) if (length(cut) == 1) cut else c(cut[1], log(diff(cut)))
row_max <- function(M) do.call(pmax, as.data.frame(M))
link_cdf <- function(z, model) if (model == "B") plogis(z) else exp(-exp(-z))
link_pdf <- function(z, model) if (model == "B") dlogis(z) else exp(-(z + exp(-z)))
ord_prob_z <- function(lo, hi, model) {
  if (model == "B") plogis(hi) * plogis(-lo) * (-expm1(lo - hi))
  else { ea <- exp(-lo); eb <- exp(-hi)
         p <- exp(-eb) * (-expm1(-(ea - eb))); p[!is.finite(eb)] <- 0; p }
}
ord_prob <- function(mubar, cut, y, model) {
  caug <- c(-Inf, cut, Inf); ord_prob_z(caug[y] - mubar, caug[y + 1L] - mubar, model)
}
negloglik <- function(par, dat) {
  design <- dat$design; P <- design$P; W <- dat$W
  n <- design$n_tasks; J <- design$J
  beta <- par[1:P]; cut <- par_to_cut(par, P, W)
  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  m <- row_max(V); logS <- m + log(rowSums(exp(V - m)))
  -(sum(V[cbind(seq_len(n), dat$jstar)] - logS) +
    sum(log(pmax(ord_prob(logS, cut, dat$y, dat$model), 1e-312))))
}
fit_dual_mle <- function(dat, start = NULL) {
  design <- dat$design; P <- design$P; W <- dat$W
  if (is.null(start)) {
    freq <- tabulate(dat$y, nbins = W)
    cumq <- pmin(pmax(cumsum(freq)[1:(W - 1)] / sum(freq), 1e-4), 1 - 1e-4)
    ginv <- if (dat$model == "B") qlogis else function(q) -log(-log(q))
    cut0 <- log(design$J) + ginv(cumq)
    if (W > 2) for (k in 2:(W - 1)) cut0[k] <- max(cut0[k], cut0[k - 1] + 1e-3)
    start <- c(rep(0, P), cut_to_par(cut0))
  }
  fn <- function(p) negloglik(p, dat)
  opt <- optim(start, fn, method = "BFGS", control = list(maxit = 1000, reltol = 1e-12))
  H <- optimHess(opt$par, fn)
  list(beta = opt$par[1:P], cut = par_to_cut(opt$par, P, W), nll = opt$value,
       par = opt$par, eigen = eigen(H, symmetric = TRUE, only.values = TRUE)$values)
}
simulate_dual <- function(design, beta, cut, model = c("A","B"), seed) {
  model <- match.arg(model); set.seed(seed)
  n <- design$n_tasks; J <- design$J
  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  u <- V + matrix(rgumbel(n * J), n, J)
  jstar <- max.col(u, ties.method = "first")
  ustar <- u[cbind(seq_len(n), jstar)]
  latent <- if (model == "A") ustar else ustar - rgumbel(n)
  list(design = design, jstar = jstar, y = findInterval(latent, cut) + 1L,
       W = length(cut) + 1L, model = model)
}
# The second-stage information weight from the paper's decomposition.
omega_weight <- function(mubar, cut, model) {
  z <- outer(mubar, cut, function(m, c) c - m)
  Gz <- cbind(0, link_cdf(z, model), 1); gz <- cbind(0, link_pdf(z, model), 0)
  W <- length(cut) + 1L
  om <- 0
  for (w in seq_len(W)) om <- om + (gz[, w + 1] - gz[, w])^2 / (Gz[, w + 1] - Gz[, w])
  om
}
```

```{r}
#| label: config
J <- 4; W <- 5
beta   <- c(a2 = 0.8, a3 = -0.5, b2 = 0.4, b3 = 1.0, price = -0.9)
cutB   <- c(-1.0, 0.2, 1.2, 2.2)
alphaA <- c(0.10, 0.30, 0.60, 0.85)
cutA   <- -log(-log(alphaA))
TAU2   <- 100
R_REPS <- 60
```

## 2. How many scale points?

The quantity to maximize is $\omega$, the second response's contribution to the
information matrix. It has a ceiling: the information a *continuous* report of
the latent would carry, which is $1$ for the Gumbel link and $1/3$ for the
logistic. Every finite scale is a partition of that continuum, and refining the
partition can only add information.

```{r}
#| label: info-by-w
I_LATENT <- c(A = 1, B = 1/3)

mean_omega <- function(cut, model, m0, s, ngrid = 81) {
  z <- seq(-4, 4, length.out = ngrid); w <- dnorm(z); w <- w / sum(w)
  sum(w * omega_weight(m0 + s * z, cut, model))
}

# Cut points giving every category the same expected share.
equal_share_cuts <- function(W, model, m0, s) {
  qs <- seq_len(W - 1) / W
  G <- if (model == "B") function(c, m) plogis(c - m) else
                          function(c, m) exp(-exp(-(c - m)))
  f <- function(c) integrate(function(m) G(c, m) * dnorm(m, m0, s),
                             m0 - 6*s - 10, m0 + 6*s + 10)$value
  vapply(qs, function(q) uniroot(function(c) f(c) - q, c(m0 - 20, m0 + 20))$root,
         numeric(1))
}

REGIMES <- list(unattractive = -1.5, typical = 0.6, attractive = 2.3)
SPREAD  <- 0.5
WGRID   <- c(2, 3, 5, 7, 9, 11)

tab_w <- do.call(rbind, lapply(names(REGIMES), function(rg) {
  m0 <- REGIMES[[rg]]
  do.call(rbind, lapply(c("A","B"), function(model) {
    do.call(rbind, lapply(WGRID, function(Wk) {
      cuts <- equal_share_cuts(Wk, model, m0, SPREAD)
      om <- mean_omega(cuts, model, m0, SPREAD)
      data.frame(regime = rg, model = model, W = Wk, omega = om,
                 frac = om / I_LATENT[model])
    }))
  }))
}))

knitr::kable(
  reshape(tab_w[, c("regime","model","W","frac")], direction = "wide",
          idvar = c("model","W"), timevar = "regime"),
  row.names = FALSE, digits = 3)
```

```{r}
#| label: fig-infoW
#| fig-cap: "Share of the continuous-report information captured by a W-point second response, cut points at equal expected shares."
suppressMessages(library(ggplot2))
tab_w$regime <- factor(tab_w$regime, levels = names(REGIMES))
g1 <- ggplot(tab_w, aes(W, frac, colour = regime, linetype = model)) +
  geom_line() + geom_point(size = 1.8) +
  scale_x_continuous(breaks = WGRID) +
  labs(x = "scale points W", y = "share of continuous-report information",
       colour = "task regime", linetype = "link",
       title = "Second-stage information by scale size",
       subtitle = "cut points at equal expected shares") +
  theme_minimal(base_size = 11)
g1
```

The shape is the finding. Going from two points to five closes most of the gap
to a continuous report; going from seven to eleven adds very little. A binary
follow-up is not merely *a* choice of scale, it is the least informative one
available, and the first few extra categories are nearly free.

## 3. Where should the cut points fall?

Scale size is the decision researchers discuss. Placement is the one they make
implicitly, by choosing labels. Compare three placements at $W = 5$: the
information-optimal one, equal expected shares, and a "conventional label"
placement reproducing the bottom-heavy distribution fielded purchase-intent
scales actually produce.

```{r}
#| label: placement
optimal_cuts <- function(W, model, m0, s) {
  start <- equal_share_cuts(W, model, m0, s)
  if (W == 2) {
    opt <- optimize(function(c) -mean_omega(c, model, m0, s),
                    interval = c(m0 - 8, m0 + 8))
    return(list(cut = opt$minimum, omega = -opt$objective))
  }
  fn <- function(p) -mean_omega(p[1] + c(0, cumsum(exp(p[-1]))), model, m0, s)
  opt <- optim(c(start[1], log(diff(start))), fn, method = "BFGS",
               control = list(maxit = 500))
  list(cut = opt$par[1] + c(0, cumsum(exp(opt$par[-1]))), omega = -opt$value)
}

placements <- list()
for (model in c("A","B")) {
  m0 <- REGIMES$typical
  eq <- equal_share_cuts(5, model, m0, SPREAD)
  op <- optimal_cuts(5, model, m0, SPREAD)
  G <- if (model == "B") function(c, m) plogis(c - m) else
                          function(c, m) exp(-exp(-(c - m)))
  conv_shares <- cumsum(c(0.35, 0.23, 0.23, 0.12))
  conv <- vapply(conv_shares, function(q) uniroot(function(c)
    integrate(function(m) G(c, m) * dnorm(m, m0, SPREAD),
              m0 - 13, m0 + 13)$value - q, c(m0 - 20, m0 + 20))$root, numeric(1))
  om <- c(optimal = op$omega,
          equal_share = mean_omega(eq, model, m0, SPREAD),
          conventional = mean_omega(conv, model, m0, SPREAD),
          binary_opt = optimal_cuts(2, model, m0, SPREAD)$omega)
  placements[[model]] <- list(optimal = op$cut, equal = eq,
                              conventional = conv, omega = om)
}

knitr::kable(do.call(rbind, lapply(c("A","B"), function(mm) {
  om <- placements[[mm]]$omega
  data.frame(link = mm, optimal = round(om[1], 4),
             `equal share` = sprintf("%.4f (%.0f%%)", om[2], 100*om[2]/om[1]),
             conventional = sprintf("%.4f (%.0f%%)", om[3], 100*om[3]/om[1]),
             `best binary` = sprintf("%.4f (%.0f%%)", om[4], 100*om[4]/om[1]),
             check.names = FALSE)})), row.names = FALSE)
```

The two links behave differently, and the difference is practical. For the
logistic link placement barely matters: equal shares is close to optimal. For the
Gumbel link it matters, because the Gumbel density is asymmetric and the
information is concentrated where it is steep. Conventional labels, which push
mass toward the bottom of the scale, cost real information under both links.

The bottom row is the one to keep in mind: even the *best possible* binary cut
sits far below a five-point scale. Choosing the scale well cannot rescue a scale
that is too coarse.

```{r}
#| label: fig-placement
#| fig-cap: "Cut placement and the cost of conventional labels at W = 5."
om_tab <- do.call(rbind, lapply(names(placements), function(mm)
  data.frame(model = mm, placement = names(placements[[mm]]$omega),
             omega = as.numeric(placements[[mm]]$omega))))
om_tab$placement <- factor(om_tab$placement,
  levels = c("optimal","equal_share","conventional","binary_opt"))
ggplot(om_tab, aes(placement, omega, fill = model)) +
  geom_col(position = "dodge") +
  labs(x = NULL, y = expression(mean~omega),
       title = "Cut placement and the cost of labels (W = 5, typical tasks)") +
  theme_minimal(base_size = 11)
```

## 4. What the design must vary

Now a different question. Scale design controls how much information the second
response carries *about the parameters*. It does not control whether the data
can tell Model A from Model B. That depends on how much the tasks vary in
attractiveness.

The reason is the identification argument in the model section. At a single value
of $\overline{\mu}$, any two strictly increasing links reproduce the $W$ category
frequencies exactly by relabelling cut points. Only variation in $\overline{\mu}$
separates them, because the two links disagree about how the response
distribution should *move* as tasks get better or worse.

We manipulate that variation directly. A task-level shift added to every
alternative's price moves the whole task's inclusive value, and because it lives
in the price column rather than a new constant, the design stays identified in
the rank-condition sense.

```{r}
#| label: link-disc
make_spread_design <- function(n_tasks, seed, sigma_task) {
  des <- make_design(n_tasks, J, seed = seed)
  if (sigma_task > 0) {
    s_t <- rnorm(n_tasks, 0, sigma_task)
    des$X[, "price"] <- des$X[, "price"] + rep(s_t, each = J)
  }
  des
}
laplace_lmd <- function(fit) {
  d <- length(fit$par)
  -fit$nll - 0.5*sum(fit$par^2)/TAU2 - 0.5*d*log(2*pi*TAU2) +
    0.5*d*log(2*pi) - 0.5*sum(log(fit$eigen))
}

run_cell <- function(sigma_task, n_tasks, reps, seed0) {
  correct_lmd <- correct_hll <- matrix(NA, reps, 2,
                                       dimnames = list(NULL, c("A","B")))
  musd <- numeric(reps)
  for (r in seq_len(reps)) {
    des <- make_spread_design(n_tasks, seed = seed0 + 11*r, sigma_task)
    n_hold <- floor(0.2 * n_tasks)
    keep <- seq_len(n_tasks) > n_hold
    subdes <- function(kv) { d2 <- des; d2$X <- des$X[rep(kv, each = J), , drop = FALSE]
                             d2$n_tasks <- sum(kv); d2 }
    des_est <- subdes(keep); des_hold <- subdes(!keep)
    for (truth in c("A","B")) {
      cut_true <- if (truth == "A") cutA else cutB
      dat_full <- simulate_dual(des, as.numeric(beta), cut_true, model = truth,
                                seed = seed0 + 11*r + 1)
      musd[r] <- sd(log(rowSums(exp(matrix(des$X %*% beta, ncol = J, byrow = TRUE)))))
      dat_est  <- list(design = des_est,  jstar = dat_full$jstar[keep],
                       y = dat_full$y[keep], W = W, model = "A")
      dat_hold <- list(design = des_hold, jstar = dat_full$jstar[!keep],
                       y = dat_full$y[!keep], W = W, model = "A")
      lmd <- hll <- c(A = NA_real_, B = NA_real_)
      for (fl in c("A","B")) {
        dat_est$model <- fl; dat_hold$model <- fl
        f <- fit_dual_mle(dat_est)
        lmd[fl] <- laplace_lmd(f)
        hll[fl] <- -negloglik(f$par, dat_hold)
      }
      correct_lmd[r, truth] <- names(which.max(lmd)) == truth
      correct_hll[r, truth] <- names(which.max(hll)) == truth
    }
  }
  list(sigma_task = sigma_task, n = n_tasks, mubar_sd = mean(musd),
       lmd = colMeans(correct_lmd), hll = colMeans(correct_hll))
}

ld_cells <- list(); seed0 <- 100000
for (sig in c(0, 0.75, 1.5)) for (n in c(1800, 3600, 7200)) {
  seed0 <- seed0 + 5000
  ld_cells[[sprintf("sig%.2f_n%d", sig, n)]] <- run_cell(sig, n, R_REPS, seed0)
}
```

```{r}
#| label: link-disc-table
knitr::kable(do.call(rbind, lapply(ld_cells, function(cl) data.frame(
  `sigma task` = cl$sigma_task, n = cl$n,
  `mubar sd` = round(cl$mubar_sd, 2),
  `P(correct | A)` = round(cl$lmd["A"], 2),
  `P(correct | B)` = round(cl$lmd["B"], 2),
  `holdout: A` = round(cl$hll["A"], 2),
  `holdout: B` = round(cl$hll["B"], 2),
  check.names = FALSE))), row.names = FALSE)
```

At this design's natural attractiveness spread, selection is already reliable
from 1,800 tasks upward, and adding engineered spread makes it more so. So on a
normal conjoint design the link is learnable.

## 5. The negative result: designs that cannot learn the link

Now remove the identifying variation entirely. Shift each task's prices by the
common amount that equalizes its inclusive value to a target. The design still
satisfies the rank condition, still has full attribute variation, still supports
estimating $\bfbeta$. It simply carries no variation in task attractiveness.

```{r}
#| label: squashed
squash_design <- function(n_tasks, seed, target_mu = 0.6) {
  des <- make_design(n_tasks, J, seed = seed)
  V <- matrix(des$X %*% beta, ncol = J, byrow = TRUE)
  m <- row_max(V); mubar <- m + log(rowSums(exp(V - m)))
  des$X[, "price"] <- des$X[, "price"] +
    rep((target_mu - mubar) / beta["price"], each = J)
  des
}

run_squashed <- function(n_tasks, reps, seed0) {
  correct <- matrix(NA, reps, 2, dimnames = list(NULL, c("A","B")))
  musd <- numeric(reps)
  for (r in seq_len(reps)) {
    des <- squash_design(n_tasks, seed = seed0 + 11*r)
    V <- matrix(des$X %*% beta, ncol = J, byrow = TRUE); m <- row_max(V)
    musd[r] <- sd(m + log(rowSums(exp(V - m))))
    for (truth in c("A","B")) {
      cut_true <- if (truth == "A") cutA else cutB
      dat <- simulate_dual(des, as.numeric(beta), cut_true, model = truth,
                           seed = seed0 + 11*r + 1)
      lmd <- c(A = NA_real_, B = NA_real_)
      for (fl in c("A","B")) { dat$model <- fl; lmd[fl] <- laplace_lmd(fit_dual_mle(dat)) }
      correct[r, truth] <- names(which.max(lmd)) == truth
    }
  }
  list(n = n_tasks, mubar_sd = mean(musd), correct = colMeans(correct))
}

sq_cells <- list()
for (n in c(3600, 7200)) sq_cells[[paste0("n", n)]] <- run_squashed(n, R_REPS, 200000 + n)

knitr::kable(do.call(rbind, lapply(sq_cells, function(cl) data.frame(
  n = cl$n, `mubar sd` = signif(cl$mubar_sd, 3),
  `P(correct | A true)` = round(cl$correct["A"], 2),
  `P(correct | B true)` = round(cl$correct["B"], 2),
  check.names = FALSE))), row.names = FALSE)
```

This is the result worth pausing on. The two probabilities do not both sit near
0.5, which is what "no information" might naively look like. Instead selection
**degenerates**: the same link gets chosen almost regardless of the truth, so one
row is near one and the other near zero. With no information in the data, the
comparison is decided by the curvature correction in the marginal likelihood
rather than by fit, and that correction does not care which model generated the
data.

Doubling the sample from 3,600 to 7,200 does not help, because the problem is
not sampling noise. The information is absent from the design, and no quantity
of data recovers it.

The practical rule follows directly. A study that wants to test the behavioral
story must engineer high- and low-attractiveness tasks. A study that only needs
demand estimates can ignore the distinction, because notebook 06 shows the two
links give nearly the same demand predictions over the range a normal design
produces.

## 6. Results written

```{r}
#| label: save
PROJ <- if (file.exists("_quarto.yml")) "." else ".."
OUT  <- file.path(PROJ, "R", "output")
dir.create(file.path(OUT, "figs"), showWarnings = FALSE, recursive = TRUE)

saveRDS(list(cells = ld_cells,
             settings = list(beta = beta, cutB = cutB, alphaA = alphaA,
                             R_REPS = R_REPS)),
        file.path(OUT, "link_discrimination.rds"))
saveRDS(list(cells = sq_cells, settings = list(beta = beta, R_REPS = R_REPS)),
        file.path(OUT, "link_discrimination_squashed.rds"))

ggsave(file.path(OUT, "figs", "info_by_W.png"), g1, width = 7, height = 4.2, dpi = 200)

saveRDS(list(info_by_W = tab_w, placements = placements,
             link_discrimination = list(cells = ld_cells),
             settings = list(regimes = REGIMES, spread = SPREAD, WGRID = WGRID)),
        file.path(OUT, "design_guidance.rds"))

cat("wrote link_discrimination.rds, link_discrimination_squashed.rds,\n",
    "design_guidance.rds, and figs/info_by_W.png\n")
```

## 7. Related material

| where | what |
|---|---|
| notebook 04 | what a coarse scale costs once the study is already fielded |
| notebook 06 | why link misspecification is second-order for demand |
| notebook 05 | the tests that use attractiveness variation to check the model |
