---
title: "Simulating and Estimating the Ordinal Dual-Response Model"
subtitle: "Model A, homogeneous preferences"
author:
  - name: Dan Yavorsky
  - name: Geoffery Zheng
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

## What this notebook does

This notebook walks through the aggregate Model A results reported under
**Parameter Recovery and Identification** in the paper's simulation study. It
does three things:

1. simulates ordinal dual-response data from the *behavioral process*, meaning
   raw Gumbel utility draws and the two reporting rules, never from the
   likelihood we are trying to validate;
2. writes out the joint likelihood and maximizes it;
3. checks that the parameters come back, and that the identification condition
   fails exactly where the theory says it should.

Everything is self-contained. The functions below are deliberate copies of the
ones in `R/lib/dgp.R` and `R/lib/likelihood.R`, unrolled and commented so the
mechanics are visible. The production versions in the replication archive are
the same calculations with more error checking and tail-stable arithmetic.

The point of simulating behaviorally rather than from the closed form deserves
emphasis. If we drew responses from `alpha_w^S - alpha_{w-1}^S` and then fit
`alpha_w^S - alpha_{w-1}^S`, recovery would confirm only that `optim` works. By
drawing utilities, taking an argmax, and applying the reporting rule, recovery
tests the derivation itself: the claim that the maximum of Gumbel utilities is
Gumbel with location equal to the log-sum inclusive value, and that it is
independent of which alternative attained it.

```{r}
#| label: setup
set.seed(1)
options(digits = 5)
```

## 1. The model in one screen

A consumer faces $J$ profiles. Profile $j$ carries attributes $x_j$ and utility

$$
u_j = x_j'\beta + \varepsilon_j, \qquad \varepsilon_j \sim \text{Gumbel}(0, 1) \text{ iid}.
$$

**First response.** She is asked which she prefers and picks the best one,
$j^* = \arg\max_j u_j$. This is the usual multinomial logit.

**Second response.** She is asked how likely she is to buy it. She knows
everything about the profiles on screen, but not her valuation of the outside
good $\eta_0 \sim \text{Gumbel}(0,1)$. So what she holds is not a decision but a
probability,

$$
p = \Pr(\eta_0 < u^* \mid u^*) = F(u^*), \qquad u^* = \max_j u_j,
$$

where $F(z) = \exp(-e^{-z})$ is the standard Gumbel CDF.

**Model A** says she reports the interval containing that probability. Given cut
points $0 = \alpha_0 < \alpha_1 < \cdots < \alpha_{W-1} < \alpha_W = 1$ on the
probability scale,

$$
y = w \iff p \in [\alpha_{w-1}, \alpha_w).
$$

From the researcher's seat $u^*$ is random. The key lemma is that
$u^* \sim \text{Gumbel}(\overline{\mu}, 1)$ with $\overline{\mu} = \ln S$ and
$S = \sum_j e^{V_j}$, and that $u^*$ is independent of $j^*$. Evaluating that
CDF at $F^{-1}(\alpha)$ collapses to a power, which gives the whole second-stage
likelihood:

$$
\Pr(y = w) = \alpha_w^{\,S} - \alpha_{w-1}^{\,S}.
$$

Equivalently, from the researcher's perspective the consumer's purchase
probability is $\text{Beta}(S, 1)$ distributed, with the inclusive value as its
only shape parameter.

## 2. A design

Two three-level attributes, dummy coded against a reference level, plus a
continuous price-like attribute. Five columns, no intercept. The absence of an
intercept is not cosmetic, as Section 9 below shows.

```{r}
#| label: design
make_design <- function(n_tasks, J, seed, intercept = FALSE) {
  set.seed(seed)
  n_rows <- n_tasks * J
  a     <- sample(1:3, n_rows, replace = TRUE)
  b     <- sample(1:3, n_rows, replace = TRUE)
  price <- runif(n_rows, 0.5, 2.5)
  X <- cbind(
    a2    = as.numeric(a == 2),
    a3    = as.numeric(a == 3),
    b2    = as.numeric(b == 2),
    b3    = as.numeric(b == 3),
    price = price
  )
  if (intercept) X <- cbind(X, const = 1)
  list(X = X, n_tasks = n_tasks, J = J, P = ncol(X))
}

N_TASKS <- 20000   # matches the paper's aggregate experiments
J       <- 4
W       <- 5       # a five-point purchase-likelihood scale

design <- make_design(N_TASKS, J, seed = 101)
head(design$X, 8)
```

Rows are stacked task-major: rows 1 to 4 are the four profiles of task 1, rows
5 to 8 are task 2, and so on.

## 3. True parameters

The scale labels are set on the probability scale, which is what makes Model A
interpretable: $\alpha_1 = 0.10$ says that answering "1" means a purchase
probability below ten percent.

```{r}
#| label: truth
beta_true  <- c(a2 = 0.8, a3 = -0.5, b2 = 0.4, b3 = 1.0, price = -0.9)
alpha_true <- c(0.10, 0.30, 0.60, 0.85)

# Cut points live on the utility scale internally. F^{-1}(a) = -log(-log a).
alpha_to_cut <- function(a) -log(-log(a))
cut_true <- alpha_to_cut(alpha_true)

rbind(alpha = alpha_true, cut = cut_true)
```

## 4. Simulate the behavioral process

Nothing here uses the likelihood. We draw utilities, take the argmax, convert the
winning utility into the consumer's purchase probability, and bin it.

```{r}
#| label: simulate
rgumbel <- function(n) -log(-log(runif(n)))

simulate_modelA <- function(design, beta, cut, seed) {
  set.seed(seed)
  n <- design$n_tasks
  J <- design$J

  # Deterministic utilities, reshaped to one row per task.
  V <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)

  # Add iid Gumbel noise: this is the consumer's single, internally
  # consistent utility draw for the task.
  u <- V + matrix(rgumbel(n * J), n, J)

  # First response: which profile is best.
  jstar <- max.col(u, ties.method = "first")

  # The utility she actually attains.
  ustar <- u[cbind(seq_len(n), jstar)]

  # Model A: she reports the interval holding p = F(ustar). Binning p against
  # alpha is the same as binning ustar against cut, since F is increasing.
  y <- findInterval(ustar, cut) + 1L

  list(design = design, jstar = jstar, y = y, W = length(cut) + 1L,
       ustar = ustar, V = V)
}

dat <- simulate_modelA(design, beta_true, cut_true, seed = 501)
```

### What the two responses look like

```{r}
#| label: responses
rbind(
  `first response (chosen profile)` = table(dat$jstar) / N_TASKS,
  `second response (scale point)`   = table(dat$y)     / N_TASKS
)
```

The first response is roughly uniform because the design randomizes attributes
across positions. The second response is not, and its shape is what the cut
points control.

### The lemma is visible in the simulated data

Two claims underlie the whole factorization. Both can be read straight off the
draws, before any estimation.

```{r}
#| label: lemma-check
S     <- rowSums(exp(dat$V))
mubar <- log(S)

# (i) ustar is Gumbel(mubar, 1). Standardize and compare to a standard Gumbel.
z <- dat$ustar - mubar
cat("mean of standardized ustar:", round(mean(z), 4),
    " (Euler-Mascheroni = 0.5772)\n")
cat("sd   of standardized ustar:", round(sd(z), 4),
    " (pi/sqrt(6) = 1.2825)\n")
cat("KS test vs standard Gumbel, p =",
    round(ks.test(sample(z, 5000), function(q) exp(-exp(-q)))$p.value, 3), "\n")

# (ii) ustar is independent of jstar: the distribution of the standardized
# maximum should not depend on which profile won.
round(tapply(z, dat$jstar, mean), 4)

# Test it rather than eyeball it. Under independence the four groups share a
# distribution, so a Kruskal-Wallis test should not reject.
kruskal.test(z, factor(dat$jstar))
```

The group means sit within sampling error of one another and the test does not
reject. That independence is what lets the joint likelihood factor into a choice
term and an ordinal term, and it is the part of the lemma that is easiest to
doubt: it says a consumer who picked profile 3 is no more or less enthusiastic,
on average, than one who picked profile 1.

## 5. The likelihood

The joint probability of one task is the product of the two pieces:

$$
\underbrace{\frac{e^{V_{j^*}}}{S}}_{\text{first response}}
\times
\underbrace{\left( \alpha_{y}^{\,S} - \alpha_{y-1}^{\,S} \right)}_{\text{second response}}.
$$

We optimize over an unconstrained parameterization that keeps the cut points
ordered by construction: the first cut point free, then the logs of successive
gaps.

```{r}
#| label: likelihood
par_to_cut <- function(par, P, W) {
  if (W == 2) return(par[P + 1])
  par[P + 1] + c(0, cumsum(exp(par[(P + 2):(P + W - 1)])))
}
cut_to_par <- function(cut) c(cut[1], log(diff(cut)))

negloglik <- function(par, dat) {
  design <- dat$design
  P <- design$P; W <- dat$W
  n <- design$n_tasks; J <- design$J

  beta <- par[1:P]
  cut  <- par_to_cut(par, P, W)

  V     <- matrix(design$X %*% beta, nrow = n, ncol = J, byrow = TRUE)
  m     <- do.call(pmax, as.data.frame(V))     # row maxima, for stability
  logS  <- m + log(rowSums(exp(V - m)))        # log inclusive value

  # First response: multinomial logit log-probability of the chosen profile.
  ll_choice <- V[cbind(seq_len(n), dat$jstar)] - logS

  # Second response: difference of Gumbel CDFs at the standardized cut points.
  caug <- c(-Inf, cut, Inf)
  lo <- caug[dat$y]      - logS
  hi <- caug[dat$y + 1L] - logS
  p_ord <- exp(-exp(-hi)) - exp(-exp(-lo))

  -(sum(ll_choice) + sum(log(pmax(p_ord, 1e-312))))
}
```

A note on that second-response line. Writing it as a plain difference of two
`exp(-exp(-z))` terms is clear but loses precision when both cut points sit far
into a tail and the two CDFs nearly cancel. The production code uses the
algebraically equivalent but numerically stable rearrangement
$e^{-e^{-h}}\left(1 - e^{-(e^{-l} - e^{-h})}\right)$. At the parameter values
here the difference is invisible; in the stress configurations of the paper it
is not.

## 6. Estimate

Starting values: $\beta = 0$, and cut points backed out of the observed category
frequencies through the link. No knowledge of the truth goes in.

```{r}
#| label: fit
fit_modelA <- function(dat) {
  P <- dat$design$P; W <- dat$W

  freq <- tabulate(dat$y, nbins = W)
  cumq <- cumsum(freq)[1:(W - 1)] / sum(freq)
  cut0 <- log(dat$design$J) + alpha_to_cut(cumq)
  start <- c(rep(0, P), cut_to_par(cut0))

  fn  <- function(p) negloglik(p, dat)
  opt <- optim(start, fn, method = "BFGS",
               control = list(maxit = 1000, reltol = 1e-12))
  H   <- optimHess(opt$par, fn)      # observed information
  Vc  <- solve(H)

  cut_hat <- par_to_cut(opt$par, P, W)

  # Delta method for the cut points, which are a nonlinear function of the
  # working parameters.
  Jc <- matrix(0, W - 1, length(opt$par))
  Jc[, P + 1] <- 1
  if (W > 2) {
    d <- exp(opt$par[(P + 2):(P + W - 1)])
    for (w in 2:(W - 1)) Jc[w, (P + 2):(P + w)] <- d[1:(w - 1)]
  }

  list(beta = opt$par[1:P], se_beta = sqrt(diag(Vc)[1:P]),
       cut = cut_hat, se_cut = sqrt(diag(Jc %*% Vc %*% t(Jc))),
       nll = opt$value, convergence = opt$convergence,
       eigen = eigen(H, symmetric = TRUE, only.values = TRUE)$values,
       par = opt$par)
}

fit <- fit_modelA(dat)
cat("converged:", fit$convergence == 0, "\n")
```

## 7. Did the parameters come back?

```{r}
#| label: recovery
recovery <- data.frame(
  parameter = c(names(beta_true), paste0("c", 1:(W - 1))),
  truth     = c(beta_true, cut_true),
  estimate  = c(fit$beta, fit$cut),
  std_error = c(fit$se_beta, fit$se_cut)
)
recovery$z <- (recovery$estimate - recovery$truth) / recovery$std_error
knitr::kable(recovery, digits = 4)
```

Every $z$ statistic is small. With nine parameters we would expect roughly one
value above two in absolute terms about forty percent of the time, so a table
with none is unremarkable rather than suspicious.

### The cut points on the probability scale

This is where Model A pays off. Pushing the estimated cut points back through
the link turns scale labels into purchase probabilities, with standard errors.

```{r}
#| label: alpha
alpha_hat <- exp(-exp(-fit$cut))
se_alpha  <- fit$se_cut * alpha_hat * (-log(alpha_hat))   # delta method

knitr::kable(
  data.frame(
    label      = paste("answer", 1:(W - 1), "or below"),
    alpha_true = alpha_true,
    alpha_hat  = alpha_hat,
    std_error  = se_alpha
  ), digits = 4)
```

A respondent choosing the top box is saying her purchase probability exceeds
`r sprintf("%.3f", alpha_hat[4])`, and we can attach a standard error to that
statement. This is the quantity that dichotomizing the scale throws away and that
assigning fixed probabilities to scale points gets wrong, because the meaning of
a label depends on what was on the screen.

## 8. Diagnostics

**Is the truth inside the confidence region?** Compare twice the
log-likelihood gap between the truth and the MLE against a $\chi^2_9$ critical
value.

```{r}
#| label: lrtest
nll_truth <- negloglik(c(beta_true, cut_to_par(cut_true)), dat)
lr   <- 2 * (nll_truth - fit$nll)
npar <- length(fit$par)
cat(sprintf("LR statistic %.2f vs chi-square(%d) 95%% critical value %.2f -> %s\n",
            lr, npar, qchisq(0.95, npar),
            ifelse(lr < qchisq(0.95, npar), "inside the region", "OUTSIDE")))
```

**Is the information matrix well conditioned?** A strongly positive definite
Hessian is the numerical face of identification.

```{r}
#| label: eigen
cat("smallest eigenvalue:", format(min(fit$eigen), digits = 4), "\n")
cat("condition number:   ", format(max(fit$eigen) / min(fit$eigen), digits = 4), "\n")
```

## 9. The identification condition, and how it fails

The paper's identification result requires $[X, \iota]$ to have full column rank
$P + 1$. The canonical violation is a constant common to every inside good: an
intercept. Adding one leaves the model's *probabilities* unchanged, because
shifting every $V_j$ by $\delta$ shifts $\overline{\mu}$ by exactly $\delta$,
which the cut points absorb one for one. What breaks is the ability to separate
the two.

```{r}
#| label: ident-fail
id_rank_check <- function(X) {
  aug <- cbind(X, 1)
  list(rank = qr(aug)$rank, required = ncol(aug))
}

cat("without intercept: "); print(unlist(id_rank_check(design$X)))
design_bad <- make_design(N_TASKS, J, seed = 101, intercept = TRUE)
cat("with intercept:    "); print(unlist(id_rank_check(design_bad$X)))
```

The rank check fails. Now fit the over-parameterized model from two different
starting points.

```{r}
#| label: ident-ridge
dat_bad <- dat
dat_bad$design <- design_bad

fit_from <- function(start_const) {
  P <- design_bad$P
  start <- c(rep(0, P - 1), start_const, cut_to_par(cut_true + start_const))
  opt <- optim(start, function(p) negloglik(p, dat_bad), method = "BFGS",
               control = list(maxit = 2000, reltol = 1e-12))
  list(beta = opt$par[1:P], cut = par_to_cut(opt$par, P, W), nll = opt$value)
}

f1 <- fit_from(-1.0)
f2 <- fit_from( 1.0)

cat(sprintf("negative log-likelihood from start 1: %.4f\n", f1$nll))
cat(sprintf("negative log-likelihood from start 2: %.4f\n", f2$nll))
cat(sprintf("difference: %.6f\n\n", abs(f1$nll - f2$nll)))

comparison <- data.frame(
  quantity = c(names(beta_true), "intercept", paste0("c", 1:(W - 1))),
  from_start_1 = c(f1$beta, f1$cut),
  from_start_2 = c(f2$beta, f2$cut)
)
knitr::kable(comparison, digits = 4)
```

Identical fit, different answers. The intercept and the cut points wander along a
flat ridge. But the *identified* combinations are pinned down:

```{r}
#| label: ident-identified
P <- design_bad$P
identified <- data.frame(
  quantity = c(names(beta_true), paste0("c", 1:(W - 1), " - intercept")),
  truth        = c(beta_true, cut_true),
  from_start_1 = c(f1$beta[1:(P - 1)], f1$cut - f1$beta[P]),
  from_start_2 = c(f2$beta[1:(P - 1)], f2$cut - f2$beta[P])
)
knitr::kable(identified, digits = 4)
```

The non-constant coefficients and the differences $c_w - \beta_{\text{const}}$
agree with the truth and with each other. This is what the proposition claims:
the model is identified up to that one location, and the design condition rules
it out.

## 10. Where this sits in the paper

The numbers above are the Model A, moderate-configuration row of the aggregate
recovery results, generated with the same seeds as the production script. The
full program in the replication archive extends this in directions the notebook
does not: three parameter configurations per link rather than one, Model B
alongside Model A, repeated sampling across forty independent datasets to check
bias and standard-error calibration, and a direct test that the joint
distribution of the two responses factors.

The scripts, in the order this notebook touches them:

| file | what it holds |
|---|---|
| `R/lib/dgp.R` | designs and behavioral simulators |
| `R/lib/likelihood.R` | the joint likelihood and the MLE fitter |
| `R/homogeneous/identification_test.R` | recovery, the intercept negative control, the factorization check |
| `R/homogeneous/replication_study.R` | repeated-sampling bias and calibration |
| `R/apollo/compare_apollo.R` | the same model estimated in Apollo, as an independent check |
