---
title: "The 2024 ordered-regression puzzle, and what it was"
subtitle: "Follow-up on Geoff's note in `_OrderedLogisticRegression.ipynb` (commit ff7d416, 2024-08-19)"
author: "Dan Yavorsky"
date: 2026-09-13
format:
  html:
    toc: true
    embed-resources: true
    code-fold: false
execute:
  warning: false
  message: false
---

::: {.callout-note}
## Purpose

Geoff raised "our identification problem" and neither of us could reconstruct
what it referred to. This note traces it to a specific finding from August 2024,
reproduces it, diagnoses it, and checks whether it affects the model as it now
stands. Short version: the finding was real and correctly described, the cause
is a reverse regression rather than an identification failure, and the paper's
current second stage does not inherit it. The concern survives in a useful form
as one of the paper's specification tests.

Not part of the manuscript. Written for discussion.
:::

## 1. What was observed

Commit `ff7d416`, 19 August 2024, titled *"observed strange phenomenon in
estimation of ordered regression"*, added `_OrderedLogisticRegression.ipynb`.
Its one markdown cell reads:

> Strange result. Here's Stata gode that replicates these phenomena. The bias is
> downwards on the $\beta$ coefficient and upwards on the magnitude of the
> estimated cutpoints.

followed by this, included so the result could not be blamed on PyMC:

```stata
clear
local N = 999999
set obs `=`N'*3'

gen grp = 1 in 1/`=`N''
replace grp = 2 in `=`N'+1'/`=2*`N''
replace grp = 3 in `=2*`N'+1'/`=3*`N''

gen mu = .
replace mu = -1 if grp == 1
replace mu = +0 if grp == 2
replace mu = +2 if grp == 3

gen x = mu + rnormal()

reg x ibn.grp, noconstant // Just to make sure it's generated correctly
estimates clear
eststo: oprobit grp x
eststo: oprobit grp, offset(x)
esttab *
```

Read the DGP carefully, because the direction matters and it is the whole
answer: the group is assigned **first**, and `x` is generated **from** the
group. The ordered probit then runs the other way, predicting the group from
`x`.

## 2. Reproduction

```{r}
#| label: repro
suppressMessages({library(MASS); library(nnet)})
set.seed(42)

N   <- 300000
grp <- rep(1:3, each = N)
mu  <- c(-1, 0, 2)[grp]
x   <- mu + rnorm(3 * N)          # x generated FROM grp

# Geoff's first specification: free coefficient on x
f1 <- polr(factor(grp) ~ x, method = "probit")
cat("free coefficient:  beta_hat =", round(coef(f1), 4),
    " (the model implies 1)\n")
cat("                   cutpoints =", round(f1$zeta, 4), "\n\n")

# His second: coefficient forced to 1, cutpoints only
f2 <- polr(factor(grp) ~ 1 + offset(x), method = "probit")
cat("offset (beta = 1): cutpoints =", round(f2$zeta, 4), "\n")
```

The finding replicates. $\hat\beta$ comes in below one, and it is not a small-
sample artifact: this is 900,000 observations.

## 3. The diagnosis

The ordered probit is misspecified for this data-generating process.

Work out what the conditional law of the group given $x$ actually is. The three
groups are equally likely, and $x \mid \text{grp} = g \sim \mathcal{N}(\mu_g, 1)$,
so by Bayes' rule

$$
\Pr(\text{grp} = g \mid x)
\;\propto\;
\tfrac{1}{3} \, \phi(x - \mu_g)
\;\propto\;
\exp\left( -\tfrac{(x - \mu_g)^2}{2} \right)
\;\propto\;
\exp\left( \mu_g x - \tfrac{\mu_g^2}{2} \right).
$$

That is a **multinomial logit in $x$ whose slope is category-specific**, equal to
$\mu_g$. An ordered probit imposes a single common slope across categories (the
parallel-lines assumption). Forcing one $\beta$ to stand in for three different
slopes attenuates it. Nothing is unidentified; the wrong model is being fit.

The prediction is checkable. If the true conditional law is multinomial logit
with slopes $\mu_g = (-1, 0, 2)$, then a multinomial logit with group 1 as
baseline should recover slope differences of $0 - (-1) = 1$ and $2 - (-1) = 3$.

```{r}
#| label: diagnosis
mn <- multinom(factor(grp) ~ x, trace = FALSE)
cat("multinomial logit slopes vs group 1:", round(coef(mn)[, "x"], 3), "\n")
cat("theory  (mu_g - mu_1)              :", c(0, 2) - (-1), "\n")
```

That settles it. The "bias" is misspecification bias from a reverse regression,
not a flat likelihood, not a normalization problem, and not an identification
failure.

## 4. Why the paper's second stage does not inherit this

Our ordinal stage generates the latent variable **first** and then bins it,
which is the ordered model's own data-generating process rather than its
reverse. Concretely: the consumer draws utilities, takes the maximum $u^*$,
and reports which interval her purchase probability falls into. The inclusive
value $\overline{\mu} = \ln S$ enters with a coefficient of exactly one, by
Lemma 1, and the same coefficient applies to every category.

```{r}
#| label: our-dgp
set.seed(7)
n <- 300000

mubar    <- rnorm(n, 0, 1.2)                 # stands in for the inclusive value
cut_true <- c(-1.0, 0.2, 1.2, 2.2)
latent   <- mubar + rlogis(n)                # Model B's second stage
y        <- findInterval(latent, cut_true) + 1L

fit <- polr(factor(y) ~ mubar, method = "logistic")
cat("slope on the inclusive value:", round(coef(fit), 4), " (theory: 1)\n")
cat("cutpoints estimated:", round(fit$zeta, 4), "\n")
cat("cutpoints true     :", cut_true, "\n")
```

No attenuation, and the cut points land on the truth. The difference between
this and §2 is entirely the direction in which the data were generated.

## 5. Where the concern survives, in a useful form

The worth of the 2024 finding is that it identified the right parameter to be
nervous about. Our model *imposes* a slope of one on the inclusive value; Geoff's
example is a case where estimating that slope freely would return something
else. Rather than assume it, the paper tests it.

The specification battery estimates

$$
\text{second-stage latent} \;=\; \lambda \, \overline{\mu}_{it} + \text{noise}
$$

and tests $\lambda = 1$ by likelihood ratio. Current results:

```{r}
#| label: lambda-test
#| echo: false
# robust to whether chunks run from archive/ or from the project root
p <- file.path("R", "output", "spec_tests.rds")
if (!file.exists(p)) p <- file.path("..", "R", "output", "spec_tests.rds")
s <- readRDS(p)
knitr::kable(
  do.call(rbind, lapply(s$cells, function(cl) data.frame(
    cell = cl$label, n = cl$n,
    `lambda hat` = round(cl$est_mean["lambda_hat"], 3),
    `reject lambda=1` = round(cl$reject["lambda=1"], 3),
    check.names = FALSE))),
  row.names = FALSE)
```

Under the null the estimate sits on one and the test holds its nominal 5% size.
Against genuine alternatives at $\lambda = 0.7$ and $1.3$ it recovers those
values and rejects with power 0.99. In Geoff's vocabulary: his `offset(x)`
specification is our model, his free-coefficient specification is the
alternative, and we report the comparison rather than assuming the answer.

Appendix E makes the same point structurally. Our model has the architecture of
a hybrid choice model, in which an ordinal measurement equation is driven by a
latent variable, except that there the loading is freely estimated and here it
is pinned at one by Lemma 1. That restriction is the theory's content, and it is
exactly the restriction Geoff's example says one should not impose casually.

## 6. A likely source of the confusion

There is now a proposition in §2 with "identification" in its name, and it is
about something else. It concerns **location**, not slope: it says $[X, \iota]$
must have full column rank $P + 1$, because an intercept common to every inside
good shifts the inclusive value one-for-one and is absorbed by the cut points.
Adding an intercept produces a genuinely flat ridge, with different starting
values giving different raw parameters at identical likelihood.

That result was added in July 2026, well after the 2024 note, and has nothing to
do with the attenuation phenomenon. If "our identification problem" is a memory
of the 2024 finding, the two are easy to run together and are unrelated. Both
are settled:

| concern | status |
|---|---|
| slope on the latent driver attenuated (2024) | reverse-regression artifact; does not arise in our DGP; tested anyway via $\lambda = 1$ |
| location not separable from cut points (2026) | real, proved in §2, demonstrated in notebook 01's negative control |

## 7. For discussion

Two things worth deciding together.

First, whether any of this belongs in the paper. My view is no: the $\lambda$
test already covers the substance, and explaining a reverse-regression artifact
that does not arise in our setting would cost a paragraph and buy confusion. But
it is a reasonable thing to disagree about, and the $\lambda$ test's motivation
would read more naturally if it said out loud *why* one might doubt that the
slope is one.

Second, whether the $\lambda$ test deserves more prominence than it currently
gets. It is reported in the specification-test table alongside the DR-2Max test
and is easy to skim past, yet it is the one result that directly answers the
question Geoff raised.
